\documentclass[a4paper,oneside, 12pt]{amsart}

\usepackage[draft,
type = paper,
externalise,
figureFolder = tikzfigures,
font = libertine,
sidenotes,
math = euler,
pkgOptions={
  hyperref = {linkcolor=red},
  geometry = {a4paper, margin=3cm}
}
]{style}

\addbibresource{local-bib.bib}
\usetikzlibrary{babel} % TODO: WIESO?!?
\pdfoutput=1
\title{Quantum Computing}
\author{Sebastian Halbig}
\author{Istvan Heckenberger}

%%% Some useful regexps:
% Word repetition: \(\<\w+\>\W-*\)\1

%%% --- Course specific definitions and macros

\newcommand*{\Pclass}{\ensuremath{\mathsf{P}}}
\newcommand*{\NPclass}{\ensuremath{\mathsf{NP}}}
\newcommand*{\BQPclass}{\ensuremath{\mathsf{BQP}}}
\newcommand*{\BPPclass}{\ensuremath{\mathsf{BPP}}}
\newcommand*{\Dtime}{\ensuremath{\mathbf{DTIME}}}


\newcommand{\underConstruction}{%
  \tikzexternaldisable
  \par\addvspace{1.5em} % Ensure new line and add space above
  \noindent
  \begingroup
  \setlength{\fboxrule}{2pt} % Thicker frame
  \setlength{\fboxsep}{10pt} % Space between frame and content
  \fbox{%
    \begin{minipage}{\linewidth-2\fboxsep-2\fboxrule}
      % Left side: The Icon
      \begin{minipage}{0.15\linewidth}
        \centering
        \scalebox{0.4}{ % Increased scale slightly for the thicker box
          \begin{tikzpicture}[limb/.style={line cap=round,line width=1.5mm,line join=bevel}]
            \draw[line width=2mm,rounded corners,fill=yellow] (-2,0) -- (0,-2) -- (2,0) -- (0,2) -- cycle;
            \fill (1.5mm,7mm) circle (1.5mm);
            \fill(0,-7.5mm) -- ++(10mm,0mm) -- ++(120:2mm)--++(100:1mm)--++(150:2mm) arc (70:170:2.5mm and 1mm);
            \draw[limb] (-7.5mm,-6.5mm)--++(70:4mm)--++(85:4mm) coordinate(a)--++(-45:5mm)--(-2.5mm,-6.5mm);
            \fill[rotate around={45:(a)}] ([shift={(-0.5mm,0.55mm)}]a) --++(0mm,-3mm)--++
            (7mm,-0.5mm)coordinate(b)--++(0mm,4mm)coordinate(c)--cycle;
            \draw[limb] ([shift={(-0.6mm,-0.4mm)}]b) --++(-120:5mm) ([shift={(-0.5mm,-0.5mm)}]c) --++
            (-3mm,0mm)--++(-100:3mm)coordinate (d);
            \draw[ultra thick] (d) -- ++(-45:1.25cm);
          \end{tikzpicture}
        }
      \end{minipage}%
      \hfill
      % Right side: The Text
      \begin{minipage}{0.80\linewidth}
        \raggedright
        \small \textbf{Under Construction:}\\
        This part of the lecture notes is still under construction and may change over the course of the semester.
      \end{minipage}
    \end{minipage}%
  }
  \endgroup
  \par\addvspace{1.5em} % Add space below
  \tikzexternalenable
}

\begin{document}
\maketitle
\tableofcontents

\cleardoublepage

\section*{About the course}
This course is an introduction to the foundations of quantum computing, specifically designed for students in computer science, data science, mathematics, and physics.
Its goal is to provide a comprehensive overview of many important aspects of theoretical quantum computing from a mathematical perspective.
Due to the complexity of the topic, however, we will not be able to cover all aspects of (theoretical) quantum computing in a single semester.

These notes are designed to be a  self-contained and detailed companion to the lectures.
At certain points, they expand beyond the topics discussed there and provide additional background information.

\subsection*{Prerequisites}
We assume a working knowledge of vector-based linear algebra.
This includes amongst other aspects in particular the ability to compute matrix-vector and matrix-matrix products, solve (small) linear equations, and to calculate the determinant of (small) matrices.
Other mathematical topics like calculus or abstract algebra can be useful at times, though they are not required.

\subsection*{Get in touch}
We strongly encourage active in-person participation in lectures and our dedicated office hours.
In case you have questions regarding the material or the exercises, please do not hesitate to reach out via email:
\begin{itemize}
  \item \textsc{Sebastian Halbig:} \url{Sebastian.Halbig@uni-marburg.de}
  \item \textsc{Istvan Heckenberger:} \url{heckenberger@mathematik.uni-marburg.de}
\end{itemize}

\subsection*{Main resources}\label{sec:main-resources}
The course is designed around the following key resources:

\begin{itemize}[leftmargin=*]
  \item  R.\ Wolf ``Quantum Computing: Lecture Notes'' (2023), \cite{wolf2023:QuantumComputing}
  \item A.\ Kockum \etal ``Lecture Notes on Quantum Computing'' (2025), \cite{kockum-soro-garcía-álvarez-et-al2025:Lecture}
  \item A.\ Kissinger, J.\ van de Wetering, ``Picturing Quantum Software'' (2026), \cite{kissinger-wetering2025:PicturingQuantumSoftware}
  \item B.\ Coecke, A.\ Kissinger ``Picturing Quantum Processes'' (2017), \cite{coecke-kissinger2017:PicturingQuantumProcesses}
  \item M.\ Nielsen, I.\ Chuang ``Quantum Computation and Quantum Information'' (2000), \cite{nielsen-chuang2000:QuantumComputationInformation}
  % \item \cite{simon2021:TopologicalQuantum}
  % \item \cite{preskill:QuantumComputing}
\end{itemize}


\numberwithin{theorem}{section}
\cleardoublepage

\section{Why quantum computing?}\label{sec:why-quantum-computing}

To motivate why quantum computers are such a fascinating topic, we will showcase---without going into technical details---how the fundamental principles of quantum mechanics can be used to solve certain problems much faster than classical computers.
This ``quantum advantage'' relies on a rather simple shift of perspective: instead of using bit logic---expressed as maps \(f\from \{0,1\}^{n}\to\{0,1\}^{m}\)---we have (almost) the full power of linear algebra at our disposal.

\addtocontents{toc}{\SkipTocEntry}
\subsection*{The Deutsch--Jozsa algorithm}\label{sec:the-deutsch--josza-algorithm}


Consider the following problem.
For a fixed number \(n \in \mathbb{N}\), we are given a map \(f \from \{0,1\}^{n}\to\{0,1\}\).
We have no information about this map except: either it is constant or balanced.
\begin{itemize}
  \item \emph{Constant:} The output is always the same (all 0s or all 1s).
  \item \emph{Balanced:} For exactly half the possible inputs, the output is \(0\) and for the other half \(1\).
\end{itemize}
We are now given the task:
\medskip

\begin{center}
  \textit{Determine which type of function we have.}
\end{center}
\medskip

\noindent
The catch: if \(f\) is balanced, we do not know which inputs belong to the half producing 0s and which inputs result in 1s.
Thus, in the worst case, a classical computer must compute the value of \(f\) for \(2^{n-1} + 1\) inputs to conclude the type of \(f\).

\subsubsection*{The quantum approach}
In contrast to the classical solution which grows exponentially in \(n\), a quantum computer can solve the above problem with a single call to (a quantum version of) \(f\).
\begin{enumerate}
  \item[Step 1:] We initialise the system by setting up \(n+1\) qubits---the quantum version of bits.
  The first \(n\) qubits are set to \(0\) and a single trailing \(1\) qubit at the end.
  We will use the first \(n\) qubits as our ``input'' and the last qubit will function as a control qubit that captures the output of \(f\).

  \item [Step 2:] While classical gates like AND and OR would allow us to transform our first \(n\) qubits (which are \((00\dots0)\)) to some other input of \(f\) (\eg \((0101\dots 01)\)), we want to use the ``magic'' of quantum mechanics to simultaneously capture all possible inputs of \(f\).
  To that end, we use Hadamard gates.
  These transform a \(0\)-qubit into a state that is equally likely \(0\)  and \(1\).
  Applying Hadamard gates to all of our \(n+1\) qubits results in particular in our desired superposition: a combination of all possible inputs of \(f\).
  Observe that we also used a Hadamard gate to alter the control qubit.
  In the next step this will be crucial in distinguishing between the balanced and the constant case.

  \item [Step 3:] We now apply the quantum version of \(f\) to this superposition and thereafter apply  Hadamard gates again to the first \(n\) qubits.
  While the details are subtle, we can roughly think of this as creating an interference pattern that depends on \(f\).
  \begin{description}
    \item [Constructive] In case \(f\) is constant, it treats all inputs the same.
    Because of the way we set up our control qubit, this results in the components of our superposition ``adding up''.
    \item [Destructive] In contrast, if \(f\) is balanced, it treats one half of our superposition differently than the other resulting in ``cancellations''.
  \end{description}


  \item [Step 4:] In order to determine the type of \(f\), we now simply have to measure the first \(n\) qubits.
  Due to the interference of the previous step we get two cases:
  \begin{enumerate}
    \item If all qubits are zero, the interference was constructive and \(f\) must have been constant.
    \item In case we observe anything else (at least one qubit being \(1\)), the function \(f\) must be balanced.
  \end{enumerate}
\end{enumerate}
Quite counterintuitively, the Deutsch--Jozsa algorithm is linear in the number of inputs \(n\) of \(f\).
For a classical computer, even a relatively small \(n=300\) would require more operations than there are atoms in the known universe\footnote{
  The number of atoms in the observable universe is estimated to be around \(10^{80}\). % TODO Some source
}, whereas a quantum computer could determine the type of \(f\) in merely \(603\) steps (601  Hadamard gates, 1 application of \(f\), and 1 measurement).

It is precisely this potential for exponential speedups, that explains the excitement around quantum computing.



\subsection{The plan for this semester}\label{sec:the-plan-for-this-semester}

\( \)\underConstruction

The preceding exposition of the Deutsch--Jozsa algorithm raises several questions that will guide us through the semester.
\bigskip

\begin{center}
  \textit{The mathematics of quantum computing}
\end{center}
\bigskip

While we tried to convey the main ideas behind an exemplary quantum algorithm, many terms (superposition, interference, \etc) remained rather opaque.
In order to properly describe and precisely analyse quantum algorithms, we must develop a rigorous mathematical framework.
For us, this framework will be that of finite-dimensional Hilbert spaces, unitary transformations,  and orthogonal projections.

\begin{thmlist}
  \item In order to describe qubits, we will explain the \emph{SCUM rules} of~\cite{kissinger-wetering2025:PicturingQuantumSoftware}:
  \begin{enumerate}
    \item [S] \emph{States} of a quantum system correspond to vectors in a Hilbert space.
    \item [C] \emph{Compound} systems are represented by tensor products.
    \item [U] \emph{Unitary} maps realise gates.
    \item [M] \emph{Measurements} are computed according to the Born rule.
  \end{enumerate}
  \item Often, explicit matrix computations can become quite cumbersome.
  As an alternative, we will discuss the \emph{ZX-calculus}---a graphical computation method that provides a lucid description of linear maps between highly compound systems.
\end{thmlist}
\bigskip

\begin{center}
  \textit{Algorithms and applications}
\end{center}
\bigskip

As demonstrated in our exposition of the Deutsch-Jozsa algorithm, the power of quantum algorithms stems from a fundamentally different approach to computation and often requires deep insights into the problem’s structure.
To further sharpen our intuition and understanding of how the paradigm of quantum computing allows for considerable speedups compared to classical computing, we now turn to what one might call the ``big four'' quantum algorithms.

\begin{thmlist}
  \item \textbf{Deutsch--Jozsa and Simon's Algorithm:} Both address pattern recognition in functions using similar methods.
  Deutsch--Jozsa determines whether a function is balanced or constant,
  while Simon's problem asks to identify a hidden periodicity.

  \item \textbf{Shor's Algorithm:} It provides an exponential speedup for factorising integers by utilising the  the quantum Fourier transform.

  \item \textbf{Grover's Algorithm:} A probabilistic algorithm that enables efficient searching of unstructured databases, achieving a quadratic speedup over classical methods.

  \item \textbf{Harrow--Hassidim--Lloyd Algorithm:}
  Offers an efficient method for gaining insights into solutions of linear systems of equations, with the solution being encoded in the quantum state amplitudes.
\end{thmlist}
\bigskip

\begin{center}
  \textit{Quantum complexity theory}
\end{center}
\bigskip

The previous algorithms suggest an advantage of quantum computers over classical machines.
We will formalise this as a conjecture phrased in the language of complexity theory and discuss the following topics:
\noindent
\begin{thmlist}
  \item \textbf{Efficient Classical Simulation:}  The \emph{Gottesman--Knill theorem} shows that certain ``incomplete'' (in particular allowing only so-called Clifford gates) quantum computers can be efficiently simulated by a classical computer.
  \item \textbf{Universal Quantum computing:} In the classical world, a set of gates is universal if any algorithm (thought of as a map \(f\from 2^{n}\to 2^{m}\)) can be realised by applying these gates finitely many times.
  In stark contrast, the ``algorithms'' for a quantum computer do not form a finite set but the topological space \(U(2^{n})\) (\ie a set with a notion of points being close to each other).
  A finite set of quantum gates is universal if finitely many gate applications can approximate any element of \(U(2^{n})\) with arbitrary precision.
  \item \textbf{The Kitaev--Solvay theorem:} Addresses the question of whether some universal gate sets are inherently more efficient than others.
  The theorem demonstrates under mild assumptions that, up to logarithmic overhead, all universal gate sets are equivalent.
  \item \textbf{An overview of complexity classes:} The class \(\Pclass\) consists of problems that are efficiently solvable by a classical computer.
  The quantum analogue is the class \(\BQPclass\) (bounded-error quantum polynomial time).
  We will formally define these classes and discuss the conjecture \(\Pclass \subsetneq \BQPclass\).
\end{thmlist}
\bigskip

\begin{center}
  \textit{Quantum error-correction}
\end{center}
\bigskip

So far, we have studied quantum computing in an idealised setting.
However, just like classical computers, we need error-correction to pass from physical to logical qubits.
Achieving fault-tolerant quantum computing is one of the most pressing challenges in the development of real-world quantum computers.

\begin{itemize}

  \item \textbf{No-Cloning Theorem:}
  Classical error correction often relies on duplicating information.
  The no-cloning theorem shows that this approach is not possible in the quantum world, necessitating alternative methods to achieving fault-tolerance.

  \item \textbf{Threshold Theorem:}
  The threshold theorem asserts that if the error rate per quantum gate is below a certain threshold, arbitrarily long quantum computations can be performed with arbitrarily small error, provided sufficient error correction is applied.
  As a consequence, universal fault-tolerant quantum computing is at least theoretically possible.

  \item \textbf{Shor’s 9-Qubit Code:}
  One of the first error-correcting codes,  demonstrating that quantum information can be protected against bit-flip and phase-flip errors.

  \item \textbf{The toric code:}
  A two-qubit system realised by combining geometric properties of the torus (\ie a doughnut) with group symmetries.
  The relevant information is encoded in global properties of these systems protecting them against all local perturbations.
\end{itemize}

\part{The mathematics of quantum computing}

\section{Linear algebra for quantum computing}\label{sec:linear-algebra-for-quantum-computing}
In the following we recapitulate basics of linear algebra necessary to discuss the axioms of quantum computing.
For further reading, see for example \cite{axler2024:LinearAlgebraDoneRight, treil2024:LinearAlgebraDoneWrong, blyth1990:ModuleTheoryLinearAlgebra}.

\subsection{Complex numbers}\label{sec:complex-numbers}
In contrast to the binary system of classical computing, quantum computing is based on the arithmetic of the complex numbers.

\begin{definition}\label{def:complex-numbers}
  The \emph{complex numbers} \( \mathbb{C} \) are the set \( \mathbb{R}^2 = \{(x, y) \mid x, y \in \mathbb{R}\} \), whose elements we write as \(x+iy\eqdef (x,y)\), endowed with two binary maps,
  \begin{align*}
    + & \from \mathbb{C} \times \mathbb{C} \to \mathbb{C}, \qquad & (a+ ib) + (x + iy) &= a+x + i(b+y), \\
    \cdot & \from \mathbb{C} \times \mathbb{C} \to \mathbb{C}, \qquad &   (a+ ib) \cdot (x +iy) &= ax - by +i(ay + bx),
  \end{align*}
  called \emph{(complex) addition} and \emph{multiplication}.
\end{definition}

Note that, while not defined as additional operations, we can \emph{subtract} complex numbers and \emph{divide} complex numbers, provided we are given a non-zero denominator.
In examples,
given \(a+ib, c+id, x+iy \in \mathbb{C}\) with \(x+iy \neq 0\), we compute
\begin{align*}
  (a+ib) - (c+id)
  & = a-c + i(b-d), \\
  \frac{1}{x+iy}
  & = \frac{x - iy}{x - iy} \frac{1}{x + iy}
    = \frac{x - iy}{x^{2} - (iy)^{2}}
    = \frac{x - iy}{x^{2}+y^{2}}.
\end{align*}

The arithmetic properties of the  complex numbers \( \mathbb{C} \) are  succinctly described by stating that \( \mathbb{C}\) forms a \emph{field}.
That is, its addition and multiplication satisfy the following axioms for all \( a, b, c \in \mathbb{C} \):

\begin{thmlist}
  \item \emph{Associativity}: \((a + b) + c = a + (b + c)\) and \((a \cdot b) \cdot c = a \cdot (b \cdot c)\);
  \item \emph{Commutativity}: \( a + b = b + a \) and \( a \cdot b = b \cdot a \);
  \item \emph{Neutral elements}: \( a + 0 = a = 0 + a \) and \( a \cdot 1 = a = 1 \cdot a \);
  \item \emph{Inverses}: For every \( d \in \mathbb{C} \), there exists an element \( -d \in \mathbb{C} \) such that \( d + (-d) = 0 \), and for every \( f \in \mathbb{C} \setminus \{0\} \), there exists an element \( f^{-1} \in \mathbb{C}\setminus\{ 0\}\) such that \( f \cdot f^{-1} = 1 \);
  \item \emph{Distributive law}: \( a \cdot (b + c) = a \cdot b + a \cdot c \).
\end{thmlist}

Beyond these basic algebraic properties, complex numbers have additional structure that will be essential throughout the course.

\begin{definition}\label{def:some-props-of-complex-numbers}
  Consider a complex number \( z = x + iy \in \mathbb{C} \).
  \begin{thmlist}
    \item Its \emph{real part} is \(\Re(z) = x\).
    \item Its \emph{imaginary part} is \(\Im(z) = y\).
    \item Its \emph{complex conjugate} is \(\overline{z}\eqdef x-iy\).
    \item Its \emph{absolute value} is \(|z| \eqdef \sqrt{z\overline{z}} = \sqrt{x^{2}+y^{2}}\).
  \end{thmlist}

  The \emph{unit circle} \(U(1)\eqdef\{z \in \mathbb{C} \mid |z| = 1\}\) consists of all complex numbers with absolute value \(1\).
\end{definition}

\begin{notation}
  For a complex number \(z \in \mathbb{C}\) with \(\Im(z) = 0\), we use the shorthand notation \(z \geq 0\) or \(z\leq 0\) instead of \(\Re(z) \geq 0\) and \(\Re(z) \leq 0\), respectively.
\end{notation}


We will often use the fact that any non-zero complex number can be written as a combination of its absolute value and some angle.

\begin{lemma}\label{lem:polar-form-uniqueness}
  Consider a non-zero complex number \(z= x +iy \in \mathbb{C}\).
  \begin{thmlist}
    \item There exist \((r, \alpha) \in \mathbb{R}^{2}\) such that
    \(z = re^{2\pi i \alpha}\).
    \item For any other pair \((s, \beta)\in \mathbb{R}^{2}\), we have \(re^{2\pi i \alpha} = s e^{2 \pi i \beta}\) if and only if \(r=s\) and \(\alpha-\beta \in \mathbb{Z}\).
    \item We have \(x = r\cos(2 \pi \alpha)\) and \(y = r\sin(2 \pi \alpha)\), implying that \(|z| = r\).
  \end{thmlist}
\end{lemma}

\begin{equation*}
  \inputtikz{complex-numbers}
\end{equation*}

\begin{definition}\label{def:polar-decomposition}
  Consider a non-zero complex number \(z=re^{2\pi i \alpha} \in \mathbb{C}\).
  We refer to \(r e^{2\pi i \alpha}\) as its \emph{polar form}, and call \(r\) the \emph{radius} and \(\alpha\) the \emph{angle} of \(z\), respectively.
\end{definition}

While it is hard to add complex numbers in polar form, it has the unquestionable advantage of simplifying complex  multiplication and division.

\begin{lemma}\label{lem:multiplication-in-polar-form}
  Let \( c = r e^{2\pi i \alpha} \) and \( d = s e^{2\pi i \beta} \) be two non-zero complex numbers in polar form.
  Their product is
  \begin{equation}
    cd = rs \; e^{2\pi i (\alpha + \beta)}.
  \end{equation}
\end{lemma}

\subsection{Vector spaces}\label{sec:vector-spaces}
The building blocks for quantum computers---qubits---will often be modelled as equivalence classes of vectors in a complex vector space.

\begin{definition}\label{def:complex-vector-space}
  A \emph{complex vector space} is a set \( V \) endowed with two maps.
  A \emph{(vector) addition} \( +\from V \times V \to V \) and a \emph{scalar multiplication} \( \cdot\from \mathbb{C} \times V \to V \) satisfying the following axioms for all \( s, v, w \in V \) and \( \alpha, \beta \in \mathbb{C} \):

  \begin{thmlist}
    \item \emph{Associativity of addition}: \( (s + v) + w = s + (v + w) \);
    \item \emph{Commutativity of  addition}: \( s + v = v + s \);
    \item \emph{Additive identity}: There exists a vector \( 0 \in V \), called the \emph{zero vector},  such that \( v + 0 = v = 0 + v \);
    \item \emph{Additive inverses}: Each vector \( v \in V \) admits an additive inverse \(-v \in V\). That is \( v + (-v) = 0 = (-v) + v \);
    \item \emph{Associativity of scalar multiplication}: \( \alpha \cdot (\beta \cdot v) = (\alpha \beta) \cdot v \);
    \item \emph{Unitality}: \( 1 \cdot v = v \).
    \item \emph{Distributivity}: \( (\alpha + \beta) \cdot (s + v) = \alpha \cdot s + \beta \cdot s + \alpha \cdot v + \beta \cdot v\).
  \end{thmlist}
  The elements of \(V\) are referred to as \emph{vectors}.
\end{definition}

A direct computation shows that the zero vector \( 0 \in V \) and additive inverses are unique.
Furthermore, the existence of additive inverses can be derived from  the distributivity between scalar multiplication and vector addition.
For any \(v\in V\) we have
\begin{equation*}
  v + (-1)\cdot v = 1\cdot v + (-1) \cdot v = (1 -1) \cdot v = 0,
\end{equation*}
implying that \(-v = (-1)\cdot v\).

\begin{example}\label{ex:one-dim-vector-space}
  The set \( \mathbb{C} \) becomes a complex vector space with its usual complex addition and multiplication as vector addition and scalar multiplication, respectively.
  The zero vector is \( 0 \in \mathbb{C} \).
\end{example}

A lot of our considerations will revolve around intelligent bookkeeping for vector spaces and their transformations.
We will achieve this using a graphical calculus.
\begin{notation}\label{not:string-diagrams.}
  We will describe vector spaces, and later matrices, or linear maps using \emph{string diagrams}.
  For now these are quite simple.
  A vector space \(V\) is simply a (horizontal) line, possibly labelled by \(V\).
  The space \( \mathbb{C}\) will be drawn transparently.
  \begin{equation}
    \inputtikz{vector-spaces-as-string-diags}
  \end{equation}
\end{notation}

Measurements play a central role in our quantum algorithms.
These will in particular ``update'' qubits such that the resulting state is contained in a (smaller) region of  of the state space.

\begin{definition}\label{def:subspace-space}
  Consider a complex vector space \(V\).
  A subset \(U\subset V\) is called a \emph{subspace} of \(V\) if it is non-empty and for all \(\lambda
  \in \mathbb{C} \) and \(x,y \in U\) we have \(\lambda x+y \in U\).
\end{definition}

\begin{proposition}\label{prop:sub-are-vs}
  Consider a subspace \(U\) of a vector space \(V\).
  Then \(U\) becomes a vector space by restricting the vector addition and scalar multiplication of \(V\) to \(U\).
\end{proposition}

Given a vector \(v\in V\) and a subspace \(U \subset V\), we write \(v+U\eqdef\{v+u \mid u \in U\}\).

\begin{proposition}\label{prop:quotients-are-vs}
  Let \(V\) be a vector space and \(U\) a subspace.
  The set \(V/U \eqdef\{x+ U \mid x \in V\}\) becomes a vector space when endowed with the vector addition and scalar multiplication
  \begin{subequations}
    \begin{align}
      (x+U) + (y+ U) &\eqdef (x+y) +U, \label{eq:quotient-vector-addition} \\
      \lambda(x + U) &\eqdef (\lambda x) + U. \label{eq:quotient-scalar-multiplication}
    \end{align}
  \end{subequations}
\end{proposition}

\begin{definition}\label{def:quotient-vs}
  Consider a vector space \(V\) and a subspace \(U\).
  We call the vector space \(V/U\) with the vector addition and scalar multiplication of Equations~\eqref{eq:quotient-vector-addition}~and~\eqref{eq:quotient-scalar-multiplication} the \emph{quotient space} of \(V\) by \(U\).
\end{definition}

Often, we want to decompose a vector space into ``complementary'' parts.

\begin{definition}\label{def:direct-sum}
  A vector space \(V\) is said to be the \emph{(internal) direct sum} of a family of subspaces \(U_{1}, \dots , U_{n} \subset V\), called \emph{direct summands}, if for every \(v\in V\) there exist for all \(1 \leq i \leq n\) unique \(u_{i}\in U_{i}\)  such that \(u_{1}+ \dots + u_{n} = v\).

  In this case, we write \(V= U_{1} \oplus  \dots \oplus U_{n}\).
\end{definition}

Suppose \(V = U \oplus W\) is decomposed into a direct sum of two subspaces \(U\) and \(W\)
While  \(U\) and \(W\) intersect trivially, \ie \(U\cap W = \{0\}\), it is not the case that \(U \cup W = V\).
In fact, \(U \cup W\) is, in general, not even a vector space.

\begin{example}\label{ex:direct-sum-c2}
  The space \( \mathbb{C}^2=\{\begin{psmallmatrix} u \\ w \end{psmallmatrix} \mid u,w \in \mathbb{C}
  \}\) is a complex vector space with component-wise addition and scalar multiplication.
  We define the subspaces
  \(U = \{\begin{psmallmatrix} u \\ 0 \end{psmallmatrix} \mid u \in \mathbb{C}\}\) and \(W = \{\begin{psmallmatrix} 0 \\ w \end{psmallmatrix} \mid w \in \mathbb{C}\}\).
  Clearly, every vector \(v =
  \begin{psmallmatrix}
    u \\ w
  \end{psmallmatrix}
  \) of \(\mathbb{C}^{2}\) is given by the sum of two vectors \(
  \begin{psmallmatrix}
    u \\0
  \end{psmallmatrix} \in U
  \) and \(
  \begin{psmallmatrix}
    0 \\ w
  \end{psmallmatrix}
  \in W
  \).
  Therefore, \( \mathbb{C}^{2}= U \oplus W\).
\end{example}

Direct sums allude to the idea of decomposing a vector space into smaller, better manageable pieces.
Based on this idea, we want to break down a vector space into its ``simplest'' non-trivial components, leading to the notion of a basis.


\begin{definition}\label{def:basis}
  An \emph{(ordered) basis} of a complex vector space \( V \) is a tuple \( \mathcal{B} = (b_i)_{i \in I} \), where \( I \) is a totally ordered index set, such that every vector \( v \in V \) can be written as a \emph{unique} linear combination
  \begin{equation*}
    v = \sum_{i \in I} \lambda_i b_i,
  \end{equation*}
  with \( \lambda_i \in \mathbb{C} \) and only finitely many \( \lambda_i \) are non-zero.
\end{definition}

One of the most fundamental theorems in linear algebra is the existence of bases.

\begin{theorem}\label{thm:existence-of-bases-and-well-defd-dimension}
  Any vector space \(V\) admits a basis \(\mathcal{B}\).
  Moreover, any other basis \(\mathcal{A}\) of \(V\) satisfies \(|\mathcal{A}| = |\mathcal{B}|\).
\end{theorem}

While bases of a vector space are highly non-unique\footnote{Every complex vector space except \(\{0\}\) admits infinitely many bases.}, the cardinality of a basis does not, by the above result, depend on a choice.

\begin{definition}\label{def:finite-dimensional-vs}
  The \emph{dimension} of a vector space \(V\) is \(\dim V \eqdef | \mathcal{B}| \) for \(\mathcal{B}\) a basis of \(V\).

  We call \(V\) \emph{finite-dimensional} if \(\dim V < \infty\).
\end{definition}

One of the most important examples for us will be the two-dimensional complex space \( \mathbb{C}^{2}\) used to model a single qubit.

\begin{example}\label{ex:c2}
  The space \( \mathbb{C}^2=\{\begin{psmallmatrix} v \\ w \end{psmallmatrix} \mid v,w \in \mathbb{C}
  \}\) is a complex vector space with component-wise addition and scalar multiplication.
  Its \emph{standard basis} is
  \begin{equation*}
    \mathcal{E}= \big(\begin{psmallmatrix} 1 \\ 0 \end{psmallmatrix},
    \begin{psmallmatrix} 0 \\ 1 \end{psmallmatrix} \big).
  \end{equation*}
  In the context of quantum computing, one writes \(\ket{0} = \begin{psmallmatrix} 1 \\ 0 \end{psmallmatrix}\) and \(\ket{1} = \begin{psmallmatrix} 0 \\ 1 \end{psmallmatrix}\) and refers to \(\mathcal{E}\) as either the \emph{computational} or \emph{\(Z\)-basis}.
\end{example}

Similar to \( \mathbb{C}^{2}\), we can define a vector space \( \mathbb{C}^{n}\) for any natural number \(n \in \mathbb{N}\).

\begin{example}\label{ex:cn}
  For any  natural number \(n\in \mathbb{N}\), we can define a vector space
  \begin{equation*}
    \mathbb{C}^n= \left\{\begin{psmallmatrix}  v_{1}\\ \vdots \\ v_{n} \end{psmallmatrix} \mid v_{1}, \dots , v_{n} \in \mathbb{C} \right\}
  \end{equation*}
  with pointwise addition and scalar multiplication.

  The current presentation of elements of \( \mathbb{C}^{n}\) as \emph{column vectors} is not particularly space-saving.
  Thus, we will often write them as \emph{row vectors} instead.
  Explicitly, we set
  \begin{equation*}
    \begin{pmatrix}
      v_{1}\\ \vdots \\ v_{n} \end{pmatrix}^{T}
    \eqdef
    ( v_{1}, \dots, v_{n} )\qquad
    \text{for all } v=     \begin{pmatrix}
      v_{1}\\ \vdots \\ v_{n} \end{pmatrix} \in \mathbb{C}^{n}
  \end{equation*}
  and call \(v^{T}\) the \emph{transpose} of \(v\).


  Similar to \( \mathbb{C}^{2}\), it has a \emph{standard basis} \( \mathcal{E} = (e_{1}, \dots , e_{n})\) given by
  \begin{equation*}
    e_{j}^{T} \eqdef
    ( 0 , \dots , 0 , 1 , 0 , \dots , 0), \qquad \text{ with \(j\)-th entry \(1\) and all other entries \(0\).}
  \end{equation*}
\end{example}

Not all vector spaces are of the form \( \mathbb{C}^{n}\) for some \(n\in \mathbb{N}\).

\begin{example}\label{ex:free-vector-space}
  Let \(S\) be a set.
  We define a vector space structure on the collection \(\mathrm{Map}(S, \mathbb{C})\) of all  maps \(f \from S \to \mathbb{C}  \) by setting
  \begin{equation*}
    (f+g)s \eqdef f(s) + g(s), \qquad (\lambda f) s \eqdef \lambda (f(s))
  \end{equation*}
  for all \(f,g \in \mathrm{Map}(S, \mathbb{C})\) and \(s \in S\).

  We refer to \(\mathrm{Map}(S, \mathbb{C})\) as the \emph{free vector space} generated by \(S\).
  By choosing a total order on \(S\), we obtain a basis \(\mathcal{B}= (\delta_{s})_{s\in S}\) of \(\mathrm{Map}(S, \mathbb{C})\), given by the \emph{delta functions}
  \begin{equation}
    \delta_{s} \from S \to \mathbb{C}\qquad \delta_{s}(t) =
    \begin{cases}
      1 & \text{ if } s= t, \\
      0 & \text{ otherwise}.
    \end{cases}, \qquad \text{ for } s\in S.
  \end{equation}

  Note that we have an inclusion
  \begin{equation*}
    \iota_{S} \from S \to \mathrm{Map}(S, \mathbb{C}), \qquad
    s \mapsto \delta_{s}.
  \end{equation*}

  If we identify,  by a slight abuse of notation, \(s\) with \(\iota_{S}(s) = \delta_{s}\), we can consider
  \begin{equation}
    \mathrm{Map}(S, \mathbb{C}) = \left\{\sum\nolimits_{s\in S} \lambda_{s} s \mid \lambda_{s}\in \mathbb{C} \text{ with } \lambda_{s}= 0 \text{ for all but finitely many } s\in S\right\}
  \end{equation}
  as the space of all (formal) linear combinations of \(S\).
\end{example}

\subsection{Matrices}\label{sec:matrices}
Matrices will play a two-fold role for us.
On the one hand, we will use them to model quantum gates.
On the other hand, they will be used to describe certain measurements.
\bigskip

For any natural number  \(n \in \mathbb{N}\), we write
\begin{equation}
  [n] \eqdef \{1, \dots , n\}.
\end{equation}
\begin{definition}\label{def:matrices}
  Consider numbers \(m, n \in \mathbb{N}\).
  An \( m \times n \) \emph{matrix} \(A\) over \( \mathbb{C} \) is a collection
  \begin{equation}
    A= (a_{ij})_{(i, j) \in [m] \times [n]} \eqdef
    \begin{pmatrix}
      a_{11} & a_{12} & \dots & a_{1n} \\
      a_{21} & a_{22} & \dots & a_{2n} \\
      \vdots& \ddots & & \vdots\\
      a_{m1} & a_{m2} & \dots & a_{mn}
    \end{pmatrix}
  \end{equation}
  with \(a_{ij} \in \mathbb{C}\) for all \(i \in [m]\) and \(j \in [n]\).
  The set of all such matrices is \( \Mat_{m,n}(\mathbb{C}) \).
  For \( m = n \), we write \( \Mat_n(\mathbb{C}) \).
\end{definition}

As we will discuss below, matrices can be used to transform vectors (via the matrix-vector multiplication).
Here, the computationally easiest case is given if all the non-zero entries of a matrix are located on the (main) diagonal.
\begin{example}\label{ex:diagonal-matrices}
  Let \(n\in \mathbb{N}\) and consider complex numbers \(\lambda_{1}, \dots , \lambda_{n} \in \mathbb{C}\).
  The matrix
  \begin{equation}
    \mathrm{diag}(\lambda_{1}, \dots , \lambda_{n}) =
    \begin{pmatrix}
      \lambda_1 & 0 & \cdots & 0 \\
      0 & \lambda_2 & & \vdots \\
      \vdots & & \ddots & 0 \\
      0 & \cdots & 0 & \lambda_n
    \end{pmatrix}
  \end{equation}
  is called a \emph{diagonal matrix}.
  In case \(\lambda_{1}= \dots = \lambda_{n} = 1\), we write \(I_{n}\eqdef \mathrm{diag}(1, \dots ,1)\in \Mat_{n}(\mathbb{C})\) and call it the \emph{identity matrix} of size \(n\).
\end{example}

The ``power'' of quantum computing arises from the calculus of non-diagonal matrices.
The two examples below will be fundamental throughout the course.

\begin{example}\label{ex:Hadamard-and-cnot-gate}
  The  matrix
  \begin{equation*}
    \begin{pmatrix}
      \tfrac{1}{\sqrt{2}} & \tfrac{1}{\sqrt{2}} \\
      \tfrac{1}{\sqrt{2}} & -\tfrac{1}{\sqrt{2}}
    \end{pmatrix}
  \end{equation*}
  is called the \emph{Hadamard gate} in quantum computing and will be used to create superpositions.

  The \emph{CNOT}-gate is given by the matrix
  \begin{equation*}
    \begin{pmatrix}
      1 & 0 & 0 & 0 \\
      0 & 1 & 0 & 0 \\
      0 & 0 & 0 & 1 \\
      0 & 0 & 1 & 0
    \end{pmatrix}.
  \end{equation*}
  We will later see how it can be used to flip a \emph{target qubit} depending on the behaviour of a \emph{control qubit}.
  This will be a source of producing entanglements.
\end{example}

One of the most important set of matrices are the Pauli-matrices.
\begin{definition}\label{def:Pauli-matrices}
  The matrices
  \begin{equation*}
    \sigma_x =
    \begin{pmatrix}
      0& 1 \\ 1& 0
    \end{pmatrix}, \qquad
    \sigma_{y} =
    \begin{pmatrix}
      0 & -i \\ i & 0
    \end{pmatrix},\qquad
    \sigma_{z} =
    \begin{pmatrix}
      1 & 0 \\ 0 & -1
    \end{pmatrix}
  \end{equation*}
  are called \emph{Pauli matrices}.
\end{definition}
For convenience, we also write \(\sigma_{0} = I_{2}\eqdef
\begin{psmallmatrix}
  1 & 0 \\ 0 & 1
\end{psmallmatrix}.
\)

\begin{remark}\label{rmk:matrices-are-a-vector-space}
  The set \( \Mat_{m,n}(\mathbb{C}) \) of \( m \times n \) matrices forms a complex vector space under entry-wise addition and scalar multiplication.
  Specifically, for \( \lambda \in \mathbb{C} \) and \( A, B \in \Mat_{m,n}(\mathbb{C}) \), we define
  \begin{subequations}
    \begin{gather}
      A + B =
      \begin{pmatrix}
        a_{11} & \dots &  a_{1n} \\
        \vdots & \ddots & \vdots \\
        a_{m1} & \dots &  a_{mn}
      \end{pmatrix}
      +
      \begin{pmatrix}
        b_{11} & \dots &  b_{1n} \\
        \vdots & \ddots & \vdots \\
        b_{m1} & \dots &  b_{mn}
      \end{pmatrix}
      \eqdef \begin{pmatrix}
        a_{11} + b_{11} & \dots & a_{1n} + b_{1n} \\
        \vdots & \ddots & \vdots \\
        a_{m1} + b_{m1} & \dots & a_{mn} + b_{mn}
      \end{pmatrix}, \\
      \lambda A =
      \lambda
      \begin{pmatrix}
        a_{11} & \dots &  a_{1n} \\
        \vdots & \ddots & \vdots \\
        a_{m1} & \dots &  a_{mn}
      \end{pmatrix}
      \eqdef \begin{pmatrix}
        \lambda a_{11} & \dots & \lambda a_{1n} \\
        \vdots & \ddots & \vdots \\
        \lambda a_{m1} & \dots & \lambda a_{mn}
      \end{pmatrix}.
    \end{gather}
  \end{subequations}
\end{remark}

While qubits are often described as equivalence classes of vectors in a complex vector space, there is an equivalent formulation via density matrices.
These are matrices which are, amongst other properties, invariant under the transformations discussed in the next definition.

\begin{definition}\label{def:adjoint}
  Consider a matrix \( A = (a_{ik}) \in \Mat_{m,n}(\mathbb{C}) \).
  \begin{thmlist}
    \item The \emph{transpose} of \( A \) is \( A^T = (a_{ki}) \in \Mat_{n,m}(\mathbb{C}) \).
    \item The \emph{conjugate} of \( A \) is \( \overline{A} = (\overline{a_{ik}}) \in \Mat_{m,n}(\mathbb{C}) \).
    \item The \emph{adjoint} of \( A \) is \( A^\dagger = \overline{A^T} \in \Mat_{n,m}(\mathbb{C}) \).
  \end{thmlist}
  In the case \( m = n \) and \( A = A^\dagger \), we call \( A \) \emph{self-adjoint} or \emph{Hermitian}.
\end{definition}


We want to perform two operations with our quantum gates:
we want to apply them to a state and we want to compose them (\ie, apply multiple ones in sequence).
The former operation will be modelled by matrix-vector multiplication.
The latter by matrix-matrix multiplication.

\begin{definition}\label{def:matrix-multiplication}
  Let \( A = (a_{ik}) \in \Mat_{m,n}(\mathbb{C}) \) be a matrix.
  \begin{thmlist}
    \item Consider a vector \(v\in \mathbb{C}^{n}\) with \(v^{T}= (v_{1}, \dots , v_{n})\).
    The \emph{matrix-vector multiplication} of \(A\) and \(v\) is the vector \(Av \in \mathbb{C}^{m}\) defined by
    \begin{equation}
      (Av)_{i} \eqdef \sum_{k=1}^{n} a_{ik} v_{k}.
    \end{equation}
    \item Given another matrix \( B = (b_{kj}) \in \Mat_{n,p}(\mathbb{C}) \), the \emph{matrix product} of \(A\) and \(B\) is the matrix \( AB \in \Mat_{m,p}(\mathbb{C}) \) whose entries are given for all \(1\leq i \leq m\) and \(1\leq j \leq p\) by
    \begin{equation*}
      (AB)_{ij} = \sum_{k=1}^n a_{ik} b_{kj}.
    \end{equation*}
  \end{thmlist}
\end{definition}

The above formula for the product of two matrices can be memorised as follows.
Consider matrices \(A\in \Mat_{m,n}(\mathbb{C})\) and \(B \in \Mat_{n,p}( \mathbb{C})\).
Suppose, we want to determine the entry \(C_{ij}\) of the product \(C= AB \in \Mat_{m,p} (\mathbb{C})\).
To that end, we proceed in two steps.
\begin{thmlist}
  \item We consider the (row)  vector \(a= \begin{psmallmatrix}
    a_{i1} & \dots & a_{in} \end{psmallmatrix}\) given by the \(i\)-th row of \(A\) and the (column) vector \(b = \begin{psmallmatrix} b_{1j}\\ \vdots \\b_{nj}  \end{psmallmatrix}
  \) given by the \(j\)-th column of \(B\).
  \item Then \( C_{ij} \) is the ``dot product'' of \( a \) and \( b \). That is,
  \begin{equation*}
    C_{ij}=
    \begin{pmatrix}
      a_{i1} & a_{i2} & \dots & a_{in}
    \end{pmatrix}
    \begin{pmatrix}
      b_{1j} \\
      b_{2j} \\
      \vdots \\
      b_{nj}
    \end{pmatrix}
    =
    a_{i1}b_{1j} + a_{i2} b_{2j} + \dots + a_{in}b_{nj}.
  \end{equation*}
\end{thmlist}
To see this pattern of combining columns and rows, consider the matrices
\begin{equation*}
  A = \begin{pmatrix}
    {\color{red}{\mathbf{a_{11}}}} & {\color{red}{\mathbf{a_{12}}}} \\
    {\color{dark blue}{a_{21}}} & {\color{dark blue}{a_{22}}}
  \end{pmatrix}, \quad
  B = \begin{pmatrix}
    {\color{dark green}{\mathbf{b_{11}}}} & {\color{pastel orange}{b_{12}}} \\
    {\color{dark green}{\mathbf{b_{21}}}} & {\color{pastel orange}{b_{22}}}
  \end{pmatrix}.
\end{equation*}
The product \( C = AB \) is
\begin{equation*}
  C =
  \begin{pmatrix}
    {\color{red}{\mathbf{a_{11}}}}{\color{dark green}{\mathbf{b_{11}}}} + {\color{red}{\mathbf{a_{12}}}}{\color{dark green}{\mathbf{b_{21}}}} & & {\color{red}{a_{11}}}{\color{pastel orange}{b_{12}}} + {\color{red}{a_{12}}}{\color{pastel orange}{b_{22}}} \\
    {\color{dark blue}{a_{21}}}{\color{dark green}{b_{11}}} + {\color{dark blue}{a_{22}}}{\color{dark green}{b_{21}}} & & {\color{dark blue}{a_{21}}}{\color{pastel orange}{b_{12}}} + {\color{dark blue}{a_{22}}}{\color{pastel orange}{b_{22}}}
  \end{pmatrix}.
\end{equation*}
A pure colour comparison shows that \eg the entry \(C_{11}\) of \(C\) is obtained by combining the first row of \(A\) with the first column of \(B\).

\begin{notation}\label{not:string-diags-for-matrices}
  In string diagrams, we represent matrices as labelled boxes between lines.
  To keep the diagrams lucid, we omit drawing identity matrices.
  As a convention, we read our diagrams from right to left.
  \begin{equation}
    \inputtikz{matrices-as-string-diags}
  \end{equation}
  Matrix multiplication is given by horizontally concatenating diagrams.
  \begin{equation}
    \inputtikz{matrix-matrix-mult-as-string-diags}
  \end{equation}

  We can also express taking transposes, adjoints, and conjugates in our diagrammatic formalism.
  \begin{equation}
    \inputtikz{matrix-transpose-adjoint-conjugate}
  \end{equation}

  To also express matrix-vector multiplication, we may consider a vector \(v \in \mathbb{C}^{n}\) as a matrix \(v\in \Mat_{n,1}( \mathbb{C})\).
  To signify that \(v\) is a vector, we draw it as a triangle.
  \begin{equation}
    \inputtikz{matrix-vector-mult-as-string-diags}
  \end{equation}
\end{notation}

To ensure that the diagrammatic approach to matrix operations is well-defined, we implicitly use that matrix multiplication is associative and the identity matrix is a multiplicative identity.

\begin{lemma}\label{lem:associativity-of-matrix-multiplication}
  Matrix multiplication is associative and unital.
  That is, for all \(A \in \Mat_{m,n}( \mathbb{C})\), \(B \in \Mat_{n,o}( \mathbb{C})\), and \( C \in \Mat_{o,p}( \mathbb{C})\), we have
  \begin{equation*}
    A(BC) = (AB)C \qquad \text{and}\qquad
    I_{m} A = A = A I_{n}.
  \end{equation*}
\end{lemma}
Note however, that matrix multiplication is \emph{not} commutative.
For every \(n\geq 2\) there are square matrices \(A, B \in \Mat_{n,n}(\mathbb{C})\), such that \(AB \neq BA\).

\begin{example}\label{ex:multiplication-table}
  The multiplication table of the Pauli matrices of Definition~\ref{def:Pauli-matrices} (with \(\sigma_{0}=I_{2}\)) is as follows:
  \begin{equation*}
    \begin{array}{l| c c c c}
      & \sigma_{0} & \sigma_{x} & \sigma_{y} & \sigma_{z} \\\hline
      \sigma_{0} & \sigma_{0} & \sigma_{x} & \sigma_{y} & \sigma_{z} \\
      \sigma_{x} & \sigma_{x} & \sigma_{0} & i \sigma_{z} & i\sigma_{y} \\
      \sigma_{y} & \sigma_{y} & -i \sigma_{z} & \sigma_{0} & i\sigma_{x} \\
      \sigma_{z} & \sigma_{z} & -i \sigma_{y} & - i\sigma_{x} & \sigma_{0}
    \end{array}
  \end{equation*}
\end{example}

\begin{definition}\label{def:invertible-matrix}
  Let \(n\in \mathbb{N}\) be a natural number.
  A matrix \( A \in \Mat_n(\mathbb{C}) \) is \emph{invertible} if there exists a matrix \( B \in \Mat_n(\mathbb{C}) \) such that \( AB = I_n = BA \), where \( I_n \) is the \( n \times n \) identity matrix.
  In this case, \( B \) is called the \emph{inverse} of \( A \) and is denoted by \( A^{-1} \eqdef B\).

  The set of invertible matrices is \(\GL_{n}( \mathbb{C})\).
\end{definition}
The inverse of an invertible matrix \(A\in \Mat_{n}( \mathbb{C})\) can be computed for example using Gaussian elimination.
Starting with
\begin{equation*}
  \left(\begin{array}{cccc|cccc}
    a_{11}& a_{12} & \dots & a_{1n}& 1 & 0 & \dots & 0 \\
    a_{21}& \ddots & \ddots & \vdots& 0 & \ddots &\ddots & \vdots \\
    \vdots& \ddots & \ddots & a_{n-1 \;n}& \vdots & \ddots & \ddots & 0 \\
    a_{n1}& \ddots & a_{n\; n-1} & a_{nn}& 0 & \dots & 0 & 1
  \end{array}\right),
\end{equation*}
we apply elementary (row) transformations until we reach a matrix of the form
\begin{equation*}
  \left(\begin{array}{cccc|cccc}
     1 & 0 & \dots & 0 & b_{11}& b_{12} & \dots & b_{1n} \\
     0 & \ddots & \ddots & \vdots & b_{21}& \ddots & \ddots & \vdots \\
    \vdots & \ddots & \ddots & 0 & \vdots& \ddots & \ddots & b_{n-1 \;n} \\
    0 & \dots & 0 & 1 & b_{n1}& \ddots & b_{n\; n-1} & b_{nn}
  \end{array}\right).
\end{equation*}
The matrix \(B=(b_{ij})\) is the inverse of \(A\).
\bigskip

Three interconnected invariants of matrices will play a central role in the study of quantum computing: the rank, trace and the determinant of a matrix (the latter indirectly, and mostly via its connection to eigenvalues).

\begin{definition}\label{def:rank}
  The \emph{rank} \(\rk(A)\) of a matrix \(A \in \Mat_{m,n}(\mathbb{C})\) is the dimension of the subspace \(\{ Av \mid v\in \mathbb{C}^{n}\}\subset \mathbb{C}^{m}\).
\end{definition}

The rank of a matrix \( A \in \Mat_{m,n}(\mathbb{C}) \) can be computed using Gaussian elimination.
By applying elementary row operations (swapping rows, multiplying a row by a non-zero scalar, and adding a multiple of one row to another), we can transform \( A \) into a matrix \(A'\) in  \emph{row echelon form}.
That is, a matrix of the form
\begin{equation*}
  A' = \left(
    \begin{array}{cccccccccc}
      0 & \dots & 0 &\multicolumn{1}{|c}{\mathbf{a_{1j_{1}}}} & \dots  & \dots& \dots & \dots & \dots  & a_{1n} \\
      \cline{4-5}
      0 & \dots & \dots & \dots & 0 &  \multicolumn{1}{|c}{\mathbf{a_{2j_{2}}}} & \dots & \dots & \dots & a_{2n} \\
      \cline{6-7}
      \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots  & \vdots\\
      0 & \dots & \dots & \dots & \dots & \dots &  0 & \multicolumn{1}{|c}{\mathbf{a_{kj_{k}}}} &\dots &  a_{3n} \\
      \cline{8-10}
      0 & \dots & \dots & \dots  & \dots & \dots & \dots & \dots & \dots & 0\\
      \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots  & \vdots\\
    \end{array}
  \right)
\end{equation*}
with \((a_{1j_{1}}, \dots , a_{kj_{k}})\)  non-zero.
These elements are called the \emph{pivot elements} of \(A'\).
The rank \(\rk (A) = k\) simply equals the number of \emph{pivot} elements of \(A'\).


\begin{definition}\label{def:trace}
  The \emph{trace} of a matrix \(A\in \Mat_{n}( \mathbb{C})\) is
  \begin{equation}
    \tr(A)= a_{11}+ \dots + a_{nn}.
  \end{equation}
\end{definition}

In the density matrix formalism, traces will be used to describe probabilities of outcomes of measurements.

\begin{proposition}\label{prop:basic-properties-of-trace}
  Let \(A, B\in \Mat_{n}( \mathbb{C})\), \(\lambda \in \mathbb{C}\), and \(G \in \GL_{n}( \mathbb{C})\).
  The trace  is
  \begin{thmlist}
    \item linear:  \(\tr(\lambda A + B) = \lambda \tr(A) + \tr(B)\),
    \item cyclic: \(\tr(AB) = \tr(BA)\), and
    \item invariant under conjugation: \(\tr(GAG^{-1}) = \tr(A)\).
  \end{thmlist}
\end{proposition}

Eigenvalues play a crucial role in quantum computing, for example via their connection to positive (semi-)definiteness.

\begin{definition}\label{def:eigenvalue-and-eigenspace}
  Let \( A \in \Mat_{n}(\mathbb{C}) \). A scalar \( \lambda \in \mathbb{C} \) is called an \emph{eigenvalue} of \( A \) if there exists a non-zero vector \( v \in \mathbb{C}^n \) such that
  \begin{equation}
    A v = \lambda v.
  \end{equation}

  The vector \( v \) is called an \emph{eigenvector} associated with \( \lambda \).
  The subspace \(\{v\in \mathbb{C}^{n} \mid Av = \lambda v\}\) is the \emph{eigenspace} of the eigenvalue \( \lambda \).
\end{definition}

For computations, it is often favourable to work with \emph{diagonal matrices}.
That is, matrices with all off-diagonal entries being zero.

\begin{definition}\label{def:diagonalisable}
  A matrix \( A \in \Mat_{n}(\mathbb{C}) \) is \emph{diagonalisable} if there exists an invertible matrix \( P \in \Mat_{n}(\mathbb{C}) \) such that
  \begin{equation*}
    P^{-1} A P =
    \begin{pmatrix}
      \lambda_1 & 0 & \dots & 0 \\
      0 & \lambda_2 & & \vdots \\
      \vdots & & \ddots & 0 \\
      0 & \cdots & 0 & \lambda_n
    \end{pmatrix}.
  \end{equation*}
  is a diagonal matrix.
\end{definition}

The connection between eigenvalues and diagonalisable matrices is due to the following result.

\begin{proposition}\label{prop:diag-iff-basis-of-eigenvalues}
  A matrix \( A \in \Mat_{n}(\mathbb{C}) \) is diagonalisable if and only if there exists a basis \( \mathcal{B} = (b_1, \dots, b_n) \) of \( \mathbb{C}^n \) consisting of eigenvectors of \( A \).

  In this case, let \( B \) be the matrix whose columns are the eigenvectors \( b_1, \dots, b_n \), and set \( \lambda_k \in \mathbb{C} \)  such that \(Ab_{k}=\lambda_{k}b_{k}\) for all \(1\le k \leq n\).
  Then,
  \begin{equation}
    B^{-1} A B =
    \begin{pmatrix}
      \lambda_1 & 0 & \cdots & 0 \\
      0 & \lambda_2 & & \vdots \\
      \vdots & & \ddots & 0 \\
      0 & \cdots & 0 & \lambda_n
    \end{pmatrix}.
  \end{equation}
\end{proposition}

The determinant is a fundamental tool for finding the eigenvalues of a matrix.
Specifically, eigenvalues of a matrix \(A\) correspond to roots of the so-called characteristic polynomial \(\chi_{A}(\lambda)\).

\begin{definition}\label{def:permutation}
  For a given \( n \in \mathbb{N} \), the set of all \emph{permutations} of \( \{1, \dots, n\} \) is
  \begin{equation}
    \Sym(n) \eqdef \{\sigma \from [n] \to [n] \mid \sigma \text{ is bijective}\}.
  \end{equation}
  The \emph{signum} of a permutation \( \sigma \) is
  \begin{equation}
    \sgn(\sigma) = \prod_{1\leq i<j \leq n}\frac{\sigma(j)-\sigma(i)}{j-i} \in \{\pm 1\}.
  \end{equation}
\end{definition}

The following definition of the determinant is called the \emph{ Leibnitz formula}.
Equivalent definitions can be given in terms of the Laplace formula, or the universal property of the determinant as alternating and normalised multilinear form.
\begin{definition}\label{def:determinant}
  The \emph{determinant} of a matrix \( A = (a_{ij}) \in \Mat_n(\mathbb{C}) \) is
  \begin{equation}
    \det(A) = \sum_{\sigma \in \Sym(n)} (-1)^{\sgn(\sigma)} a_{1\sigma(1)} \cdots a_{n\sigma(n)}.
  \end{equation}
\end{definition}

Unlike traces, the determinant is not linear\footnote{%
  It is however, linear in the columns and rows of a matrix, respectively.}.

\begin{proposition}\label{prop:basic-properties-of-determinant}
  Consider matrices \(A, B \in\Mat_{n}( \mathbb{C}) \) and \(P \in \GL_{n}( \mathbb{C})\).
  The determinant is
  \begin{thmlist}
    \item multiplicative: \(\det(AB) = \det(A) \det (B)\),
    \item normalised: \(\det(I_{n}) = 1\), and
    \item invariant under conjugation: \(\det(PAP^{-1})= \det(A)\).
  \end{thmlist}
\end{proposition}

One shows that \(A\in \Mat_{n}( \mathbb{C})\) is invertible if and only if \(\det(A) \neq 0\).
This can be cleverly used to study eigenvalues.
Suppose that \(\lambda\) is an eigenvalue of \(A\) and \(v \in \mathbb{C}^{n}\) is an eigenvector
Then
\begin{equation*}
  (\lambda I_{n} - A) v = \lambda I_{n} v - Av = \lambda v - \lambda v =0.
\end{equation*}
Since \(0 \neq v\) the matrix \(\lambda I_{n} - A\) cannot be invertible and \(\det (\lambda I_{n} - A) = 0\).
Thus, we want to systematically find all \(t \in \mathbb{C}\) such that  \(\det(t I_{n} - A) = 0\).
c
\begin{definition}\label{def:char-poly}
  The \emph{characteristic polynomial} of a matrix \( A \in \Mat_n(\mathbb{C}) \) is
  \begin{equation}
    \chi_A(t) = \det(t I_n - A) \in \mathbb{C}[t].
  \end{equation}
\end{definition}
We can now state our previously made observation about the connection between eigenvalues and determinants as a theorem.
\begin{theorem}\label{thm:eigenvalues-and-char-poly}
  A scalar \(\lambda \in \mathbb{C}\) is an eigenvalue of a matrix \(A\in \Mat_{n}( \mathbb{C})\) if and only if \(\chi_{A}(\lambda) = 0\).
\end{theorem}

The coefficients of the characteristic polynomial are closely connected with the invariants of matrices we have previously discussed.

\begin{proposition}\label{prop:coeffs-of-char-poly}
  The characteristic polynomial of a matrix \(A \in \Mat_{n}( \mathbb{C})\) satisfies
  \begin{equation}
    \chi_{A}(t) = t^{n} - \tr(A)t^{n-1} + \dots + (-1)^{n}\det(A).
  \end{equation}
\end{proposition}
Let \(v\in \mathbb{C}^{n}\) be a vector and \(A \in \Mat_{n}( \mathbb{C})\) a matrix.
By considering \(v^{\dagger}= (\overline{v_{1}}, \dots , \overline{v_{n}})\in \Mat_{1,n}( \mathbb{C})\) as a matrix, we can compute the scalar
\begin{equation*}
  v^{\dagger}Av =\sum_{i,k=1}^{n} \overline{v_{i}}a_{ik}v_{k} \in \mathbb{C}.
\end{equation*}

\begin{definition}\label{def:positive-semidefinite}
  A self-adjoint matrix \(A\in \Mat_{n}( \mathbb{C})\) is \emph{positive semidefinite} if one of the following equivalent conditions hold
  \begin{thmlist}
    \item For all \(v\in \mathbb{C}^{n}\) we have \(v^{\dagger}Av \geq 0\).
    \item All eigenvalues \(\lambda\) of \(A\) are real and \(\lambda \geq 0\).
    \item For each \( 1 \leq i \leq n \), the \emph{leading principal minor} \( A_i \)---the \( i \times i \) submatrix of \( A \) formed by its first \( i \) rows and columns---satisfies \( \det(A_i) \geq 0 \).
  \end{thmlist}
\end{definition}

Albeit an abstract result at first glance, the Cayley--Hamilton theorem describes many of the structural properties of matrices.
\begin{theorem}[Cayley--Hamilton]\label{thm:cayley-hamilton}
  Every matrix \( A \in \Mat_n(\mathbb{C}) \) satisfies
  \begin{equation*}
    \chi_{A}(A) = 0 \in \Mat_{n}( \mathbb{C}).
  \end{equation*}
\end{theorem}
Here, \( \chi_A(A) \) is the matrix obtained by substituting \( A^k \) for \( t^k \) in the characteristic polynomial \( \chi_A(t) \), with the convention that \( A^0 = I_n \).
For example,  the \(2 \times 2\)-matrix \(A =\begin{psmallmatrix} a & b \\ c & d \end{psmallmatrix}\)
has the characteristic polynomial
\begin{equation*}
  \chi_A(t) = \det(t I_2 - A) = t^2 - (a + d)t + (ad - bc)
\end{equation*}
and inserting \(A\) yields
\begin{equation*}
  \chi_{A}(A) = A^{2} - (a + d) A + (ad-bc) I_{2} = A^{2} - \tr(A)A + \det(A)I_{2}.
\end{equation*}

\begin{corollary}\label{cor:consequences-of-caley-hamilton}
  Consider a square matrix \( A \in \Mat_n(\mathbb{C}) \) with characteristic polynomial \(\chi_{A}(t) = t^{n} + c_{n-1} t^{n-1} + \dots + c_{1}t + c_{0}\).
  \begin{thmlist}
    \item \( A \) is invertible if and only if \( \det(A) \neq 0 \).
    In this case, the inverse is
    \begin{equation}
      A^{-1}= - \frac{1}{c_{0}}(A^{n-1} + c_{n-1} A^{n-2} + \dots + c_{1} I_{n}).
    \end{equation}
  \end{thmlist}
  Suppose that \(\lambda_{1}, \dots , \lambda_{k}\in \mathbb{C}\) are the pairwise distinct roots of \(\chi_{A}(t)\) with multiplicities  \(m_{1}, \dots , m_{k}\in \mathbb{N}\), \ie \(\chi_{A}(t) = (t-\lambda_{1})^{m_{1}} \cdot \dots \cdot (t-\lambda_{k})^{m_{k}}\).
  \begin{thmlist}[resume]
    \item  The eigenvalues of \(A\) are \(\lambda_{1}, \dots , \lambda_{k}\).
    \item  The trace of \(A\) is \(\tr(A) =  \lambda_{1}m_{1} + \dots \lambda_{k}m_{k}\).
    \item  The determinant of \(A\) is \(\det(A) = \lambda_{1}^{m_{1}} \cdot \dots \cdot \lambda_{k}^{m_{k}}\)
    \item The matrix \(A\) is diagonalisable if for every \(1\leq j\leq k\) the eigenspace \(E(A, \lambda_{j})\) of \(A\) for the eigenvalue \(\lambda_{j}\) has dimension \(\dim(E(A, \lambda_{j})) = m_{j}\).
  \end{thmlist}
\end{corollary}

\subsection{Linear maps}\label{sec:linear-maps}

While matrix-vector multiplication provides a way to act on vectors in \( \mathbb{C}^n \), we will later encounter more general vector spaces that cannot be canonically identified with \( \mathbb{C}^{n}\).
This necessitates a more abstract and general notion of transformation.

\begin{definition}\label{def:linear-maps}
  A \emph{map} \( f \from V \to W \) between complex vector spaces \(V\) and \(W\) is called  a \emph{linear map} or \emph{homomorphism of vector spaces } if
  \begin{subequations}
    \begin{align}
      f(x + y) &= f(x) + f(y), \\
      f(\lambda x) &= \lambda f(x),
    \end{align}
  \end{subequations}
  for all \( x, y \in V \) and \( \lambda \in \mathbb{C} \).
  In case there exists a linear map \(g \from W \to V\) such that \(g f = \id_{V} \) and \(f g = \id_{W}\), we call \(f\) an \emph{isomorphism} and \(g\) the \emph{inverse} of \(f\).

  The set of all linear maps from \( V \) to \( W \) is \(\Hom_{ \mathbb{C}}(V, W) \).
  If \(V=W\), we call \(f\) an \emph{endomorphism} and write \(\End_{ \mathbb{C}}(V) \eqdef \Hom_{ \mathbb{C}}(V,V)\).
\end{definition}

In Proposition~\ref{prop:basic-properties-of-trace}, we have already seen that the trace map \(\tr \from \Mat_{n}( \mathbb{C}) \to \mathbb{C}\) is linear.

\begin{example}\label{ex:matrices-lead-to-linear-multiplication}
  Consider a matrix \(A\in \Mat_{m,n}( \mathbb{C})\).
  The map
  \begin{equation}
    \mu_{A} \from \mathbb{C}^{n}\to \mathbb{C}^{m}, \qquad v \mapsto Av.
  \end{equation}
  is linear.

  The image \(\mu_{A}(e_{j})\) of the \(j\)-th standard basis vector coincides with the \(j\)-th column of \(A\).
\end{example}

\begin{definition}\label{def:image-and-kernel}
  Consider two vector spaces \(V\) and \(W\) and a linear map \(f\ \from V \to W\).
  Its \emph{kernel} is the subspace \(\{v \in V \mid f(v) = 0\} \subset V\).
  The \emph{image} of \(f\) is the subspace \(\{f(v) \mid v\in V\} \subset W\).
\end{definition}

Similar to matrices, the collection of all linear maps between two fixed vector spaces forms again a vector space.

\begin{remark}\label{rmk:vector-space-structure}
  The set \( \Hom_{\mathbb{C}}(V, W) \) naturally forms a vector space under pointwise addition and scalar multiplication.
  That is, for all \(f, g \in \Hom_{ \mathbb{C}}(V,W)\), \(\lambda \in \mathbb{C}\), and \(v\in V\) we have
  \begin{equation*}
    (f+g)v = f(v) + g(v)\qquad \text{ and } \qquad
    (\lambda f)v = \lambda (f (v)).
  \end{equation*}
  Here, the zero vector is given by the map constant zero map
  \begin{equation*}
    V\to W, \qquad v \mapsto 0.
  \end{equation*}
\end{remark}

\begin{definition}\label{def:dual space}
  We call \(\rd*{V} \eqdef \Hom_{ \mathbb{C}}(V,   \mathbb{C} )\) the \emph{dual space} of \(V\).
\end{definition}

We now want to link linear maps to matrices.
As an auxiliary step, we first notice that any linear map is uniquely determined by its values on a chosen basis of the domain.

\begin{theorem}\label{thm:universal-property-free-vector-space}
  Consider a vector space \(V\) with basis \( \mathcal{B}\) and let \(W\) be any vector space.
  Any (set) map \(g \from \mathcal{B} \to W\) can be extended to a unique linear map \(f\from V \to W\) such that \(f(b) = g(b)\) for all \(b\in \mathcal{B}\).
\end{theorem}

As an immediate consequence, we can identify any \(n\)-dimensional space (non-uniquely!) with \( \mathbb{C}^{n}\).

\begin{corollary}\label{cor:basis-isomorphism}
  For any basis \( \mathcal{B} = (b_{1}, \dots , b_{n})\)  of a vector space \(V\), the map
  \begin{equation*}
    \can_{\mathcal{B}} \from
    \mathbb{C}^{n} \to V, \qquad
    \begin{pmatrix}
      \lambda_{1}\\ \vdots \\ \lambda_{n}
    \end{pmatrix}
    \mapsto
    \lambda_{1} b_{1} + \dots + \lambda_{n} b_{n}
  \end{equation*}
  is an isomorphism.
\end{corollary}

One of the most powerful tools in linear algebra is to translate between linear maps and matrices by choosing bases.

\begin{proposition}\label{prop:matrices-linear-maps}
  Consider a vector space \(V\) with basis \( \mathcal{B} = (b_{1}, \dots , b_{n})\), a space \(W\) with basis \( \mathcal{C} = (c_{1}, \dots , c_{m})\), and a linear map \(f\from V \to W\).
  \begin{thmlist}
    \item There is a  unique matrix  \(A= (a_{ij}) \in \Mat_{m,n}( \mathbb{C})\) such that
    \begin{equation*}
      f(b_{j})= a_{1j}c_{1} + \dots + a_{mj} c_{m}\qquad \text{ for all } 1\leq j \leq n.
    \end{equation*}
    \item The following diagram commutes
    \begin{equation*} %TODO fix labels.
      \begin{tikzcd}[ampersand replacement=\&,cramped]
        V \&\& W \\\\
        {\mathbb{C}^n} \&\& {\mathbb{C}^m}
        \arrow["f", from=1-1, to=1-3]
        \arrow["{\can_{\mathcal{B}}}", from=3-1, to=1-1]
        \arrow["{\mu_A}"', from=3-1, to=3-3]
        \arrow["{\can_{\mathcal{C}}}"', from=3-3, to=1-3]
      \end{tikzcd}
    \end{equation*}
    In other words, we have \(f \can_{ \mathcal{B}} = \can_{\mathcal{C}}\mu_{A}\).
  \end{thmlist}
\end{proposition}

\begin{definition}\label{def:transformation-matrix}
  Retain the notation of the previous proposition: \( V \)  is a vector space with basis \( \mathcal{B} \), \( W \) a space with basis \( \mathcal{C} \), and \( f \from V \to W \) a linear map.
  The matrix \( A \) such that \( f \circ \can_{\mathcal{B}} = \can_{\mathcal{C}} \circ \mu_A \) is called the \emph{transformation matrix} of \( f \) with respect to the bases \( \mathcal{B} \) and \( \mathcal{C} \).

  To emphasise the choice of bases, we also  denote this matrix by \({_{\mathcal{C}}}{f}{_{\mathcal{B}}} \eqdef A\).
\end{definition}

We write \(\id_{V} \from V\to V\), \(v \mapsto v\) for the identity map of a vector space \(V\).

\begin{lemma}\label{lem:composition-linear-maps}
  Let \(f \from U \to V\),  \(g \from V \to W\), and \(h \from W\to X\) be linear maps.
  \begin{enumerate}
    \item The composition \(g\circ f \from U \to X\) of linear maps  is a linear map.
    \item We have \((h\circ g)\circ f = h \circ (g \circ f)\).
    \item The composition satisfies \(\id_{V} \circ f = f = f \circ \id_{U}\).
  \end{enumerate}
\end{lemma}

Under the identification between linear maps and matrices (once bases are chosen), the composition of linear maps corresponds to matrix multiplication.
In particular, the representing matrix of the composition \( g \circ f \) is the product of the representing matrices of \( g \) and \( f \).

\begin{proposition}\label{prop:comp-is-matrix-mult}
  Consider linear maps \(f \from V \to W\) and \(g \from W\ \to X\) and bases \(\mathcal{B}= (b_{1}, \dots , b_{n})\) of \(V\), \(\mathcal{C}= (c_{1}, \dots , c_{m})\) of \(W\), and \(\mathcal{D} = (d_{1}, \dots , d_{p})\) of \(X\).
  Let \(F = _{ \mathcal{C}}f_{\mathcal{B}}\) and \( G = _{\mathcal{D}} g_{\mathcal{C}}\) be transformation matrices of \(f\) and \(g\), respectively.
  Then, the following diagram commutes:
  \begin{equation}
    \begin{tikzcd}[ampersand replacement=\&,cramped]
      V \&\& W \&\& X \\
      \\
      {\mathbb{C}^n} \&\& {\mathbb{C}^m} \&\& {\mathbb{C}^p}
      \arrow["f", from=1-1, to=1-3]
      \arrow["gf", curve={height=-24pt}, from=1-1, to=1-5]
      \arrow["g", from=1-3, to=1-5]
      \arrow["{\can_{\mathcal{B}}}", from=3-1, to=1-1]
      \arrow["{\mu_F}"', from=3-1, to=3-3]
      \arrow["{\mu_{GF}}"', curve={height=24pt}, from=3-1, to=3-5]
      \arrow["{\can_{\mathcal{C}}}", from=3-3, to=1-3]
      \arrow["{\mu_G}"', from=3-3, to=3-5]
      \arrow["{\can_{\mathcal{D}}}"', from=3-5, to=1-5]
    \end{tikzcd}
  \end{equation}
\end{proposition}

Note that representing a linear map \(f\from V\to W\) by a matrix \({}_{\mathcal{C}}f_{\mathcal{B}}\) depends on choices on bases \(\mathcal{B}\) of \(V\) and \(\mathcal{C}\) of \(W\).
This naturally raises the question, how \({}_{\mathcal{C}}f_{\mathcal{B}}\) transforms under a change of bases.

\begin{corollary}\label{cor:change-of-basis}
  Let \( \mathcal{B}, \mathcal{B}' \) be bases of \(V\) and \( \mathcal{C}, \mathcal{C}' \) bases of \(W\).
  For any linear map \(f \from V\to W\), we have
  \begin{equation*}
    {}_{\mathcal{C}' }f_{\mathcal{B}' }= \big({}_{\mathcal{C}' }\id_{\mathcal{C}}\big)\big({}_{\mathcal{C} }f_{\mathcal{B} }\big)\big({}_{\mathcal{B} }\id_{\mathcal{B}' }\big).
  \end{equation*}
\end{corollary}

\begin{notation}\label{not:string-diags-for-linear-maps}
  One take-away of the previous results is that upon taking bases, linear maps can be translated into matrices and vice-versa.
  Subsequently, we can extend our string diagrams to general vector spaces and their transformations by representing linear maps  as labelled boxes between lines.
  Compare this in particular with the conventions of Notation~\ref{not:string-diags-for-matrices}.
  \begin{equation}
    \inputtikz{lin-maps-as-string-diags}
  \end{equation}
  Similar to the identity matrix, we ommit a box for the identity map \(\id_{V} \from V \to V\).

  Compositions of linear maps(\ie multiplying their transformation matrices) is given by horizontally concatenating diagrams.
  \begin{equation}
    \inputtikz{composition-as-string-diags}
  \end{equation}
  To also apply a linear map to a vector, we consider a vector \(v \in \mathbb{C}^{n}\) as the linear map

  \begin{equation*}
    \nu \from \mathbb{C} \to V,\quad \lambda \mapsto \lambda v.
  \end{equation*}
  Using the convention that \(\mathbb{C}\) is drawn transparently, we get the following diagrammatic representation.
  \begin{equation}
    \inputtikz{vectors-in-string-diagrams}
  \end{equation}
\end{notation}

For any pair of bases \(\mathcal{B}, \mathcal{B}'\)  of \(V\) the two matrices \({}_{\mathcal{B} }\id_{\mathcal{B}'}\) and \({}_{\mathcal{B}'}\id_{\mathcal{B}}\) are mutually inverse to each other.
Therefore, by using the cyclic property of the trace, we get for any endomorphism \(f \from V \to V\) that
\begin{equation*}
  \tr\big({}_{\mathcal{B}' }f_{\mathcal{B}'}\big)
  = \tr\bigg(\big({}_{\mathcal{B}' }\id_{\mathcal{B}}\big)\big({}_{\mathcal{B} }f_{\mathcal{B} }\big)\big({}_{\mathcal{B} }\id_{\mathcal{B}' }\big)\bigg)
  = \tr\bigg(\big({}_{\mathcal{B} }f_{\mathcal{B} }\big)\big({}_{\mathcal{B} }\id_{\mathcal{B}' }\big)\big({}_{\mathcal{B}' }\id_{\mathcal{B}}\big)\bigg)
  = \tr\big({}_{\mathcal{B}}f_{\mathcal{B}}\big)
\end{equation*}

Many properties of matrices we discussed earlier are invariant under base change, allowing us to formulate them abstractly for linear maps.
\begin{definition}\label{def:trace-of-an-endomorphism}
  Suppose \(f\from V \to V\) is linear map between finite-dimensional vector spaces and \(\mathcal{B}\) is a basis of \(V\).
  \begin{thmlist}
    \item The \emph{trace} of \(f\from V \to V\) is \(\tr(f) = \tr(_{\mathcal{B}}f_{\mathcal{B}})\);
    \item The \emph{determinant} of \(f \from V\to V\) is \(\det(f) = \det(_{\mathcal{B}}f_{\mathcal{B}})\);
    \item The \emph{rank} of \(f\) is \(\rk(f) = \dim \im f\).
  \end{thmlist}
\end{definition}

We conclude our detour into linear algebra by discussing a special kind of linear map---projectors---which play a fundamental role in quantum mechanics.
These will be used to represent measurement outcomes in the quantum formalism. Moreover, in the density formalism, certain states are represented by rank-1 projectors.

\begin{definition}\label{def:projection}
  A \emph{projection} onto a subspace \(U \subset V\) is a linear map \(f\from V \to V\) such that \(\im f \eqdef\{ f(v) \mid v \in V\} = U\) and \(f(u) = u \) for all \(u \in U\).
\end{definition}
In other words, \( f \) acts as the identity on \( U \) and maps every vector in \( V \) to its ``component'' in \( U \) (with respect to a suitable decomposition of \( V= \ker f \oplus U \)).

\begin{lemma}\label{lem:equivalent-characterisations-projections}
  Suppose \(f \from V\to V\) is linear.
  The following are equivalent.
  \begin{thmlist}
    \item \(f\) is a projection onto its image \(\im f =\{f(v) \mid v\in V\}\).
    \item\(f^{2} = f\),
    \item for every  \(v\in V\) we have \(v\in \im(f)\) if and only if \(f(v) = v\), and
    \item every vector \(v\in V\) can be written as a unique linear combination \(v= x+y\) with \(f(x) = 0\) and \(y\in \im(f)\).
    In this case \(x= (\id_{V} -f) v\) and \(y = f(v)\).
  \end{thmlist}
\end{lemma}

Any transformation matrix \(_{\mathcal{B}}f_{\mathcal{B}}\) of a projector must square to itself.
Subsequently, its eigenvalues can be either \(0\) or \(1\), leading to the following result.
\begin{proposition}\label{prop:matrix-representation-of-projection}
  For any projector \(f\from V \to V\) on an \(n\)-dimensional space \(V\), there exists a basis \(\mathcal{B}\) of \(V\) such that \(_{\mathcal{B}}f_{\mathcal{B}} = \mathrm{diag}(1, \dots , 1, 0, \dots , 0)\) is a diagonal matrices with \(\dim \im f\) diagonal entries equal to \(1\).
\end{proposition}


\cleardoublepage
\subsection{Exercises}\label{sec:lin-alg-exercises}
\begin{exercise}
  Let \(z_1 = 3 + 4i\) and \(z_2 = 1 - i\).
  \begin{thmlist}
    \item Express \(z_1 + z_2\), \(z_1 z_2\), and \(z_1 / z_2\) in Cartesian form (\ie as linear combinations \(a + ib \in \mathbb{C})\).
    \item Write \(z_2\) in polar form \(re^{2\pi i \alpha}\).
    \item Find a complex number \(u\) such that \(u^{3}= z_{2}\).
    \item Determine \emph{integers} \(p,q \in \mathbb{Z}\) such that \(z_{1}^{2} + p z_{1} + q = 0\).
    \item Prove that for any \(z \in \mathbb{C}\setminus\{0\}\), we have \(z^{-1} = \tfrac{\overline{z}}{||z||}\).
  \end{thmlist}
\end{exercise}

\begin{exercise}
  Recall that \(e^{i\theta} = \cos\theta + i\sin\theta\).
  Using the properties of complex multiplication in polar form, derive the addition theorems:
  \begin{align*}
    \cos(\alpha + \beta) &= \cos\alpha \cos\beta - \sin\alpha \sin\beta \\
    \sin(\alpha + \beta) &= \sin\alpha \cos\beta + \cos\alpha \sin\beta
  \end{align*}
\end{exercise}

\begin{exercise}\label{ex:props-of-Pauli-matrices}
  Consider the Pauli matrices \(\sigma_x, \sigma_y, \sigma_z\) defined in Definition~\ref{def:Pauli-matrices}.
  \begin{thmlist}
    \item Verify that \(\sigma_x^2 = \sigma_y^2 = \sigma_z^2 = I_2\).
    \item Show that \(\tr(\sigma_j) = 0\) for \(j \in \{x, y, z\}\).
    \item Prove that the Pauli matrices are self-adjoint (Hermitian), \ie \(\sigma_j^\dagger = \sigma_j\) for all \(j \in\{x,y,z\}\).
    \item Show that \((\sigma_{0}, \sigma_{x}, \sigma_{y}, \sigma_{z})\) is a basis of \(\Mat_{2}( \mathbb{C})\).
  \end{thmlist}
  Define the two matrices \(P_{\pm} = \frac{1}{2}(I_2 \pm \sigma_x)\).
  \begin{thmlist}[resume]
    \item Show that \(P_{\pm}^{2} = P_{\pm}\) and that \(P_+ P_- = 0\).
    \item Find an invertible matrix \(U\in \GL_{2}( \mathbb{C})\) such that \(U^{-1}P_{\pm}U\) are diagonal matrices.
  \end{thmlist}
\end{exercise}

\begin{exercise}
  Let \(A \in \Mat_n(\mathbb{C})\) with eigenvalues \(\lambda_{1}, \lambda_{2}, \dots , \lambda_{n}\) (not necessarily distinct).
  \begin{thmlist}
    \item Formulate what it means for a number \(\mu \in \mathbb{C}\) to be an eigenvalue of \(A\).
    \item Let \(v \in \mathbb{C}^{n}\) be an eigenvector of \(A\) for the eigenvalue \(\lambda_{1}\).
    What is \(A^{2}v\)?
    \item Determine the eigenvalues of \(A^{2}\).
  \end{thmlist}
\end{exercise}

\begin{exercise}[very hard]
  Let \(A \in \Mat_n(\mathbb{C})\) with eigenvalues \(\lambda_{1}, \lambda_{2}, \dots , \lambda_{n}\) (not necessarily distinct).
  One can express the determinant of \(A\) in terms of traces of powers of \(A\).
  The following sketches this idea in the case \(n\in\{2,3\}\).
  \begin{thmlist}
    \item For \(n=2\), show that \(2\det(A) = (\tr(A)^2 - \tr(A^2))\).
    \textit{Hint}: Use the Cayley-Hamilton theorem or the relationship between trace/determinant and eigenvalues.
  \end{thmlist}
  For \(n=3\), prove that \(6 \det(A) = \tr(A)^{3} - 3 \tr(A)\tr(A^{2}) + 2 \tr(A^{3})\).
  Proceed as follows
  \begin{thmlist}
    \item  The characteristic polynomial of \(A\) is
    \begin{equation*}
      \chi_{A}(t) = t^{3} - \alpha t^{2} + \beta t - \gamma = (t- \lambda_{1})(t- \lambda_{2})(t-\lambda_{3})
    \end{equation*}
    Determine the coefficient \(\beta\) as a combination of the \(\lambda_{i}\)s.
    \item Using the previous exercise, compute  \(\tr(A)^{2}\) and \(\tr(A^{2})\).
    Express \(\beta\)  in terms of \(\tr(A)^{2}\) and \(\tr(A^{2})\).
    \item Now use the connection between \(\alpha\) and \(\gamma\) and the trace and characteristic polynomial of \(A\) to conclude the proof.
  \end{thmlist}
\end{exercise}


\begin{exercise}
  Let \(V\) be a finite-dimensional vector space and \(p \in \End_{\mathbb{C}}(V)\).
  Prove that \(p\) is a projection if and only if there exists a subspace \(U \subset V\), a surjective linear map \(f \from V \to U\), and an injective linear map \(g \from U \to V\) such that \(f \circ g = \id_U\) and \(p = g \circ f\).
\end{exercise}

\begin{exercise}
  Let \(V = \mathbb{C}^2\) with the computational basis \(\mathcal{B} = (\ket{0}, \ket{1})\).
  Let \(H\) be the Hadamard gate (Example~\ref{ex:Hadamard-and-cnot-gate}).
  \begin{thmlist}
    \item Determine the linear map \(\mu_{H}\from \mathbb{C}^{2}\to \mathbb{C}^{2}\).
    \item Let \(\mathcal{B}' = (\ket{+}, \ket{-})\) where \(\ket{+} = H\ket{0}\) and \(\ket{-} = H\ket{1}\).
    Show that \(\mathcal{B}'\) is a basis.
    \item Find the change-of-basis matrix \(P = {}_{\mathcal{B}}\id_{\mathcal{B}'}\).
    \item Determine the transformation matrix of \(\mu_{H}\) with respect to the basis \(\mathcal{B}'\).
  \end{thmlist}
\end{exercise}

\begin{exercise}
  Use string diagrams to express the following identities.
  \begin{thmlist}
    \item Write the composition \((f\circ \id_{V})\circ h\) with \(h \from W \to V\) and \(f \from V\to W\).
    \item Visualise the defining identity of a projection \(p \from V \to V\) in terms of string diagrams.
    \item A linear map \(f \from V \to W\) is invertible with inverse \(g \from W \to V\).
    \item Given a matrix \(A\in \Mat_{n}( \mathbb{C})\), express \(v^{T}Av\) for \(v\in \mathbb{C}^{n}\).
  \end{thmlist}
\end{exercise}

\cleardoublepage
\subsection{Study goals}\label{sec:lin-alg-study-goals}
The questions below are designed to help you build a deep, conceptual understanding of the linear algebra required in this course.
You should aim to answer and explain your reasoning independently, without relying on external aids such as search engines or AI tools.
The level of difficulty is aligned with what you can expect in the exam.

\subsubsection*{Complex numbers}
\begin{thmlist}
  \item Express the multiplicative inverse of a complex number \(z \in \mathbb{C}\setminus\{0\}\) as  \(a + bi\) with \(a, b \in \mathbb{R}\).
  \item What are the real and imaginary part of a complex number \(z = re^{2\pi i \alpha} \in \mathbb{C}\)?
  \item Consider \(z=a+ib \in \mathbb{C}\) with \(a>0, b<0\) and suppose \(z = re^{2\pi i \alpha}\). What can be said about \(\alpha \in [0,1]\)?
\end{thmlist}

\subsubsection*{Vector Spaces}
\begin{thmlist}
  \item Is the set of all \(n \times n\) Hermitian matrices a complex subspace of \(\Mat_n(\mathbb{C})\)?
  \item Consider two matrices \(A, B \in \Mat_{4}( \mathbb{C})\) and let \(U =\{v \in \mathbb{C}^{4} \mid Uv = 0\}\) and \(W =\{Bv \mid v \in \mathbb{C}^{4}\}\).
  How do you check if they form a direct sum \(U \oplus W = \mathbb{C}^4\)?
  \item Suppose \(V\) is an \(n\)-dimensional vector space and \(\mathcal{B}=(b_{1}, \dots , b_{m})\) is a collection of vectors.
  In case \(m\geq n\), is it true that \(\lambda_{1}b_{1} + \dots + \lambda_{m} b_{m} = 0\)  if and only if \(\lambda_{1} = \dots = \lambda_{m} = 0 \).
\end{thmlist}

\subsubsection*{Matrices}
\begin{thmlist}
  \item Given a matrix \(A \in \Mat_{m,n}(\mathbb{C})\), what is the vector \(Ae_i\), where \(e_i\) is the \(i\)-th standard basis vector?
  \item How do you check if a vector \(v\) is an eigenvector of a matrix \(A \in \Mat_{n}( \mathbb{C})\) with eigenvalue \(\lambda\)?
  \item If a square matrix \(A \in \Mat_n(\mathbb{C})\) has full rank (\(\rk(A) = n\)), is it necessarily invertible? Justify your answer.
  \item Let \(A, B \in \Mat_{n}( \mathbb{C})\) be self-adjoint matrices.
  Is their sum \(A+B\) always self-adjoint?
  For \(\lambda \in \mathbb{C}\), under what conditions is \(\lambda A\) self-adjoint?
  \item State two equivalent conditions for a self-adjoint matrix to be positive semidefinite.
  \item Is the matrix \(\begin{psmallmatrix} 1 & 2 \\ 2 & -1 \end{psmallmatrix}\) positive semidefinite?
  \item Suppose \(A\) is conjugate to the diagonal matrix \(\mathrm{diag}(\lambda, \mu)\). What is \(\tr(A)\)?
  \item Can a matrix \(A\) with eigenvalues \(\{1, 2, 3\}\) be a projection?
\end{thmlist}

\subsubsection*{Linear maps}
\begin{thmlist}
  \item Let \(f\from V \to W\) be a linear map between a \(p\)-dimensional space \(V\) with basis \(\mathcal{B}\) and a \(q\)-dimensional space \(W\) with basis \(\mathcal{C}\).
  Is \(_{\mathcal{C}}f_{\mathcal{B}} \in \Mat_{p,q}( \mathbb{C})\) or \(_{\mathcal{C}}f_{\mathcal{B}} \in \Mat_{q,p}( \mathbb{C})\)?
  Justify your answer.
  \item Consider the  map
  \begin{equation*}
    f \from \Mat_2(\mathbb{C}) \to \mathbb{C}, \qquad
    \begin{psmallmatrix}
      a & b \\
      c & d
    \end{psmallmatrix}
    \mapsto
    b+c.
  \end{equation*}
  Show that it is linear.
  \item Consider the map \(f \from \Mat_{2} ( \mathbb{C}) \to \mathbb{C}\) of the previous exercise.
  Furthermore, let \(\mathcal{B} = \left(
    \begin{psmallmatrix}
      1 & 0 \\ 0 & 0
    \end{psmallmatrix},
    \begin{psmallmatrix}
      0 & 1 \\ 0 & 0
    \end{psmallmatrix},
    \begin{psmallmatrix}
      0 & 0 \\ 1 & 0
    \end{psmallmatrix},
    \begin{psmallmatrix}
      0 & 0 \\ 0 & 1
    \end{psmallmatrix}
  \right)\) be the standard basis of \(\Mat_2(\mathbb{C})\) and \( \mathcal{C}= (1)\)  the standard basis of \( \mathbb{C}\).
  Determine the transformation matrix \(_{ \mathcal{C}}f_{\mathcal{B}}\in \Mat_{1,4}( \mathbb{C})\).
  \item For a projection \(p \in \End(V)\), what is the relationship between \(\tr(p)\) and the dimension of its image \(\im(p)\)?
  \item Suppose \(p\from V\to V\) is a projection onto \(U\subset V\). Given a vector \(v\in V\), draw \(p(v)\),
  and explain the components of the diagram. Now suppose \(v\in U\), simplify the diagram.
\end{thmlist}
\cleardoublepage

\section{The quantum circuit model and the SCUM axioms}\label{sec:the-quantum-circuit-model-and-the-scum-axioms}
% TODO Norm in \mathbb{C} is |z| not \|z\|!
As suggested by their name, quantum computers rely on the principles of quantum mechanics.
Following \cite[Section~2]{kissinger-wetering2025:PicturingQuantumSoftware}, we present a simplified and tailored version of the basic axioms called in their terminology the  \emph{SCUM postulates}.

\subsection{Finite-dimensional Hilbert spaces}\label{sec:finite-dimensional-hilbert-spaces}
The systems we consider will be modeled by finite-dimensional Hilbert spaces---complex vector spaces endowed with an inner product which allows us to define basic geometric concepts like angles and lengths.


\begin{definition}\label{def:unitary-vector-space}
  Let \(V\) be a complex vector space.
  An \emph{inner product} is a map
  \begin{equation*}
    \braket{\blank}{\bblank} \from V\times V \to \mathbb{C}
  \end{equation*}
  which is
  \begin{thmlist}
    \item linear in the second argument: \(\braket{a}{\lambda b + c} = \lambda \braket{a}{b} + \braket{a}{c}\) for all scalars \(\lambda \in \mathbb{C}\) and vectors \(a,b,c \in V\);
    \item Hermitian: \(\braket{a}{b} = \overline{\braket{b}{a}}\) for all \(a, b \in V\);
    \item positive definite: \(\braket{a}{a} \geq 0 \)  for all \(a\in V\) and \(\braket{a}{a} = 0\) if and only if \(a = 0\).
  \end{thmlist}
  We call a pair \((V, \braket{\blank}{\bblank})\) of a complex vector space together with an inner product a \emph{unitary vector space}.
\end{definition}

Combining the Hermitian symmetry of an inner product with the linearity in its second argument implies that an inner product is up-to-conjugation linear in its first argument.

\begin{remark}\label{rmk:sesquilinear}
  A  map \(\phi \from V \times W \to \mathbb{C}\) is \emph{sesquilinear} if for all \(v, v' \in V\), \(w, w' \in W\) and \(\lambda, \mu \in \mathbb{C}\), we have
  \begin{gather*}
    \phi(v+v', w+ w') = \phi(v,w) + \phi(v',w) + \phi(v, w') + \phi(v',w'), \\
    \phi(\lambda v, w) = \overline{\lambda} \phi(v, w) \qquad \text{ and } \phi(v, \mu w) = \mu \phi(v,w).
  \end{gather*}
  If \(\braket{\blank}{\bblank} \from V\times V \to \mathbb{C}\) is an inner product, it is in particular sesquilinear, see Exercise~TODO
\end{remark}



The inner product of \(V\) induces the \emph{norm}
\begin{equation}
  \| \blank \| \from V \to \mathbb{R}_{\geq0} ,\qquad   v \mapsto \|v\| \eqdef  \sqrt{\braket{v}{v}}.
\end{equation}

A sequence \((v_n)_{n \in \mathbb{N}}\) in \( V \) is called a \emph{Cauchy sequence} if for every \(\varepsilon > 0\), there exists an \(N \in \mathbb{N}\) such that
\begin{equation}
  \|v_m - v_n\| \leq \varepsilon \qquad \text{for all } m, n \geq N.
\end{equation}
A \emph{limit} of \((v_{n})_{n\in \mathbb{N}}\), if it exists, is a vector \(w \in V\) such that for every \(\varepsilon>0\) there is an \(N\in \mathbb{N}\) with
\(\|v_{n}- w\|< \varepsilon\) for all \(n \geq N\).
In this case,  we write \(\lim\limits_{n\to \infty} v_{n} \eqdef w\) and say that \((v_{n})_{n\in \mathbb{N}}\) \emph{converges} to \(w\).

\begin{definition}\label{def:Hilbert-space}
  A \emph{Hilbert space} is a unitary vector space \((V, \braket{\blank}{\bblank})\) such that
  every Cauchy sequence \((v_{n})_{n\in \mathbb{N}}\) converges.
\end{definition}

In case the inner product of a Hilbert space or unitary vector space is apparent from the context, we simply write \(V\) instead of \((V, \braket{ \blank }{\bblank})\).
\begin{lemma}\label{lem:fin-dim-implies-Hilbert}
  Any finite-dimensional unitary vector space is a Hilbert space.
\end{lemma}
In the infinite-dimensional case, there are Cauchy sequences within  unitary spaces that do not converge.
In particular, not every unitary space is automatically a Hilbert space.

\begin{convention}
  For the rest of this section, all spaces will be finite-dimensional Hilbert spaces.
\end{convention}

In a Hilbert space \(V\) the \emph{angle} \(\theta \in [0, \pi)\) between two non-zero vectors \(v,w \in V\) can be defined as
\begin{equation*}
  \theta \eqdef \cos^{-1}\left(\frac{\braket{u}{v}}{\|u\| \|v\|} \right).
\end{equation*}
In particular \(u\) and \(v\) are perpendicular to each other if and only if \(\braket{u}{v} = 0\).

\begin{definition}\label{def:orthogonal-complement}
  Consider a subspace \(U\subset V\) of a Hilbert space \(V\).
  The \emph{orthogonal complement} of \(U\) is the subspace
  \begin{equation}
    U^{\perp} \eqdef\{v \in V \mid \braket{u}{v} = 0\} \subset V.
  \end{equation}
\end{definition}

Note that  for any subspace \(U \subset V\), we have \(U \oplus U^{\perp} = V\), see Exercise~\ref{exe:orth-complement-direct-sum}.

\begin{definition}\label{def:orthonormal-basis}
  An \emph{orthonormal basis} of a finite-dimensional unitary vector space \((V, \braket{\blank}{\bblank})\) is a sequence of vectors \(\mathcal{B}= (b_{1}, \dots , b_{n})\)  such that \(\mathcal{B}\) is a basis and
  \begin{equation*}
    \braket{b_{l}}{b_{k}}=
    \begin{cases}
      1 & \text{ if } l = k, \\
      0 & \text{otherwise,}
    \end{cases} \qquad \text{ for all } b_{l}, b_{k} \in \mathcal{B}.
  \end{equation*}
\end{definition}

\begin{example}\label{ex:c2-as-Hilbert-space}
  We turn the 2-dimensional complex vector space  \(\mathbb{C}^2=\{\begin{psmallmatrix} v \\ w \end{psmallmatrix} \mid v,w \in \mathbb{C}\}\) into a Hilbert space by endowing it with its \emph{standard inner product}
  \begin{equation*}
    \braket{ \blank }{\bblank}_{\mathrm{std}} \from \mathbb{C}^{2}\times \mathbb{C}^{2}\to \mathbb{C}, \qquad
    \left( \begin{psmallmatrix} v_1 \\ v_2 \end{psmallmatrix}, \begin{psmallmatrix} w_1 \\ w_2 \end{psmallmatrix} \right) \mapsto \overline{v_{1}}w_{1} + \overline{v_{2}}w_{2}.
  \end{equation*}

  Its standard basis \(\mathcal{E}= (\ket 0, \ket 1)\)---also called the \emph{computational} or \emph{\(Z\)-basis}---is given by
  \begin{equation*}
    \ket{0} = \begin{pmatrix} 1 \\ 0 \end{pmatrix}\qquad \text{and}\qquad
    \ket{1} = \begin{pmatrix} 0 \\ 1 \end{pmatrix},
  \end{equation*}
  and is an orthonormal basis.

  Other common orthonormal bases for \(\mathbb{C}^2\) include the \emph{\(X\)-basis} \((\ket{+}, \ket{-})\), where
  \begin{equation*}
    \ket{+} = \frac{1}{\sqrt{2}} \begin{pmatrix} 1 \\ 1 \end{pmatrix}, \qquad \text{ and } \qquad
    \ket{-} = \frac{1}{\sqrt{2}} \begin{pmatrix} 1 \\ -1 \end{pmatrix},
  \end{equation*}
  and the \emph{\(Y\)-basis} \((\ket{+i}, \ket{-i})\)
  \begin{equation*}
    \ket{+i} = \frac{1}{\sqrt{2}} \begin{pmatrix} 1 \\ i \end{pmatrix}, \qquad \text{ and } \qquad
    \ket{-i} = \frac{1}{\sqrt{2}} \begin{pmatrix} 1 \\ -i \end{pmatrix}.
  \end{equation*}
\end{example}

Similar to the previous example, we can turn \( \mathbb{C}^{n}\) for all \(n\in \mathbb{N}\) into a Hilbert space.

\begin{example}\label{ex:cn-as-Hilbert-space}
  The \emph{standard scalar product} on \( \mathbb{C}^{n}\) is
  \begin{equation*}
    \braket{ \blank }{ \bblank }_{\mathrm{std}}\from \mathbb{C}^{n}\times \mathbb{C}^{n}\to \mathbb{C}, \qquad
    \left(
      \begin{pmatrix}
        v_{1} \\ \vdots \\ v_{n}
      \end{pmatrix},
      \begin{pmatrix}
        w_{1} \\ \vdots \\ w_{n}
      \end{pmatrix}
    \right)
    \mapsto \overline{v_{1}}w_{1} + \dots + \overline{v_{n}}w_{n}.
  \end{equation*}
  The pair \(( \mathbb{C}^{n}, \braket{ \blank }{ \bblank }_{\mathrm{std}})\) is a Hilbert space and the standard basis \(\mathcal{E}\) defined in Example~\ref{ex:cn} is an orthonormal basis.
\end{example}

While any Hilbert space has a basis, such a basis need not a priori be orthonormal.
This can be achieved using Gram--Schmidt.
\begin{proposition}\label{prop:existence-of-onb}
  Let \(\mathcal{B} = (b_{1}, \dots , b_{n})\) be a vector space basis of a Hilbert space \(V\).
  Then \(\mathcal{O} = (w_{1}, \dots , w_{n})\), given by
  \begin{equation*}
    w_{i} = \frac{1}{\|w'_{i}\|} w'_{i} \qquad \text{with } w'_{i} = b_{i} - \sum_{j=1}^{i-1} \frac{\braket{ b_{i}}{ b_{j}}}{\braket{b_{j}}{b_{j}}} b_{j}
  \end{equation*}
  is  an orthonormal basis.
\end{proposition}
Take for example the basis \(\mathcal{B} = \left(
  \begin{psmallmatrix}
    1  \\ 1
  \end{psmallmatrix},
  \begin{psmallmatrix}
    0 \\ 1
  \end{psmallmatrix}
\right)\) of \( \mathbb{C}^{2}\).
The procedure of the previous proposition, called the \emph{Gram-Schmidt procedure}, produces the vectors
\begin{equation*}
  w_{1}'=
  \begin{pmatrix}
    1 \\ 1
  \end{pmatrix}, \qquad \text{ and }\qquad
  w_{2}' =
  \begin{pmatrix}
    0 \\ 1
  \end{pmatrix} - \frac{1}{2}
  \begin{pmatrix}
    1 \\ 1
  \end{pmatrix}
  = \frac{1}{2}
  \begin{pmatrix}
    -1 \\ 1
  \end{pmatrix}
\end{equation*}
After normalisation, this leads to the orthonormal basis \(\mathcal{O}=(w_{1}, w_{2})\) with
\begin{equation*}
  w_{1} = \frac{1}{\sqrt{2}}
  \begin{pmatrix}
    1 \\ 1
  \end{pmatrix}, \qquad \text{ and } \qquad
  w_{2} = \frac{1}{\sqrt{2}}
  \begin{pmatrix}
    -1 \\ 1
  \end{pmatrix}.
\end{equation*}
In pictures this process can be visualised as follows.
\begin{equation*}
  \inputtikz{gram-schmidt-example}
\end{equation*}


Recall that the adjoint \(A^{\dagger} = \overline{A}^{T}\) of a matrix  \(A \in \Mat_{m,n}( \mathbb{C})\) is given by transposing \(A\) (\ie changing rows to columns and vice versa) and taking complex conjugates entry-wise.

\begin{remark}\label{rmk:adjoint-and-standard-product}
  For any matrix \(A\in \Mat_{m,n}( \mathbb{C})\), we have
  \begin{equation}
    \braket{ v }{Aw} = \braket{A^{\dagger}v}{w} \qquad \text{for all \(v \in \mathbb{C}^{m}\) and \(w \in \mathbb{C}^{n}\)}.
  \end{equation}
  Subsequently, we  write
  \begin{equation}
    \braket{v}{A \mid w} \qquad \text{ for all } v\in \mathbb{C}^{m}, w \in \mathbb{C}^{n} \text{, and }A \in \Mat_{m,n}( \mathbb{C}).
  \end{equation}
  In particular, one can show that a square matrix \(A\in \Mat_{n}( \mathbb{C})\) is self-adjoint (\ie \(A^{\dagger} = A\)) if and only if
  \(\braket{ v }{Aw} = \braket{Av}{w}\) for all \(v, w \in \mathbb{C}^{n}\).
\end{remark}

We now want to describe the interaction between inner products and linear transformations more abstractly.

\begin{definition}\label{def:adjoint-of-linear-map}
  Let \(f\from V \to W\) be a linear map between two Hilbert spaces \((V, \braket{ \blank }{ \bblank})\) and \((W, [ \blank \mid \bblank])\).
  We call a linear map \(f^{\dagger} \from W \to V\) an \emph{adjoint} of \(f\) if
  \begin{equation*}
    \braket{f^{\dagger}(w)}{v} = [w\mid f(v)]\qquad\quad \text{ for all } v\in V \text{ and } w\in W.
  \end{equation*}

  In case \(g\from V\to V\) satisfies \(g^{\dagger} = g\), we call \(g\) \emph{self-adjoint}.
\end{definition}

This notion recovers our previously discussed concept of adjoint matrices.

\begin{proposition}\label{prop:adjoint-morph-is-adjoint-matrix}
  Any linear map \(f \from V \to W\) between Hilbert spaces has a unique adjoint \(f^{\dagger} \from W \to V\).
  Moreover, if \(\mathcal{B}\) is an orthonormal basis of \(V\) and \(\mathcal{C}\) is an orthonormal basis of \(W\), the transformation matrices \(A = {_{\mathcal{C}}f_{\mathcal{B}}}\) and \(B = {_{\mathcal{B}}{f^{\dagger}}{}_{\mathcal{C}}}\) are adjoint to each other.
  That is, \(A^{\dagger} = B\) and \(B^{\dagger} =A\).
\end{proposition}
\begin{proof}
  Let us first show by that \(f\) has at most one adjoint.
  Suppose both \(g,h \from W\to V\) are adjoints of \(f\).
  For all \(w\in W\) we have
  \begin{align*}
    \braket{g(w)- h(w)}{g(w) - h(w)}
    & = \braket{g(w)}{g(w) - h(w)} - \braket{h(w)}{g(w) - h(w)}\\
    & = \braket{w}{f(g(w) - h(w))} - \braket{w}{f(g(w) - h(w))}
      = 0.
  \end{align*}
  This implies  \(g(w) = h(w)\) for all \(w\in W\) and so \(g = h\).

  To prove existence, let \( \mathcal{B} = (b_1, \dots, b_n) \) and \( \mathcal{C} = (c_1, \dots, c_m) \) be orthonormal bases for \( V \) and \( W \), respectively. The canonical maps \( \can_{\mathcal{B}} \from \mathbb{C}^n \to V \) and \( \can_{\mathcal{C}} \from \mathbb{C}^m \to W \) satisfy
  \begin{equation*}
    \braket{\can_{\mathcal{B}}(v)}{\can_{\mathcal{B}}(w)} = \braket{v}{w}_{\mathrm{std}} \quad \text{for all } v, w \in \mathbb{C}^n,
  \end{equation*}
  and similarly for \( \can_{\mathcal{C}} \).

  Let \(A= {}_{\mathcal{C}}f_{\mathcal{B}}\) be the transformation matrix of \(f\) with respect to the bases \(\mathcal{B}\) and \(\mathcal{C}\).
  We define \(f^{\dagger} = \can_{\mathcal{B}} \circ \mu_{A^{\dagger}} \circ \can_{\mathcal{C}}^{-1}\) as the composition
  \begin{equation*}
    W \xrightarrow{\can_{\mathcal{C}}^{-1}} \mathbb{C}^{m}\xrightarrow{\mu_{A^{\dagger}}} \mathbb{C}^{n}\xrightarrow{\can_{\mathcal{B}}} V.
  \end{equation*}
  Now, consider \(v \in V\) and \(w\in W\).
  Using that there are unique \(x \in \mathbb{C}^{n}\) and \(y \in \mathbb{C}^{m}\) such that
  \(\can_{\mathcal{B}}(x) = v\) and \(\can_{\mathcal{C}}(y) = w\), we compute
  \begin{align*}
    \braket{f^{\dagger}(w)}{v}
    & = \braket{f^{\dagger} (\can_{\mathcal{C}}(y))}{\can_{\mathcal{B}}(x)}
      = \braket{(\can_{\mathcal{B}} \circ \mu_{A^{\dagger}} \circ \can_{\mathcal{C}}^{-1}) (\can _{C} (y))}{\can_{B}(x)} \\
    & = \braket{A^{\dagger}(y)}{x}_{\mathrm{std}}
      = \braket{y}{Ax}_{\mathrm{std}}
      = [\can_{\mathcal{C}}(y) \mid \can_{\mathcal{C}}(Ax)]\\
    & = [\can_{\mathcal{C}}(y) \mid (\can_{\mathcal{C}}\circ \mu_{A} \circ \can_{\mathcal{B}}^{-1})(x)]
      = [w\mid f(v)].
  \end{align*}
  By construction, we have \(_{\mathcal{B}}f^{\dagger}{}_{\mathcal{C}} = A^{\dagger}\) which concludes the proof.
\end{proof}

We can now define a notion of positive-semidefiniteness for linear maps by mimicking the matrix version of Definition~\ref{def:positive-semidefinite}.
\begin{definition}\label{def:positive-semidiefinite-maps}
  A self-adjoint map \(f\from V \to V\) is positive semidefinite if
  \begin{equation*}
    \braket{v}{f(v)} \geq 0\qquad \text{ for all } v\in V.
  \end{equation*}
\end{definition}

A key fact about self-adjoint operators is the spectral theorem.
\begin{theorem}\label{thm:spectral-theorem}
  Any self-adjoint map \(f\from V\to V\) admits an orthonormal basis of eigenvectors \(\mathcal{B}=(b_{1}, \dots , b_{n})\) with real eigenvalues \(\lambda_{1}\geq \dots \geq \lambda_{n}\).

  Moreover, \(f\) is self-adjoint if and only if \(\lambda_{1} \geq \dots \geq \lambda_{n}\geq 0\).
\end{theorem}
The proof of the spectral theorem can be done in an inductive fashion.
Suppose the statement holds for all self-adjoint maps \(f\from V \to V\) between \(n\)-dimensional spaces and let \(g\from W\to W\) be a self-adjoint map on an \(n+1\)-dimensional space.
Since \( \mathbb{C}\) is algebraically closed, it must have a maximal eigenvalue \(\lambda\).
We pick a normalised eigenvector \(v \in W\)  with respect to the eigenvalue \(\lambda\) and set \(V=\{w\in W\mid \braket{w}{v} = 0\}\).
A direct computation shows that \(g(V)\subset V\).
Using the induction hypothesis, we chose an orthonormal basis \(\mathcal{C} = (c_{1}, \dots , c_{n})\) of \(V\) given by eigenvectors of \(g\) and set \(\mathcal{B} = (v, c_{1}, \dots , c_{n})\).

\begin{remark}\label{rmk:spectral-theorem-results}
  Our formulation of the spectral theorem does not require eigenvalues to be mutually distinct.
  In particular, there can be linearly independent (non-zero) vectors \(v\in V\) such that for the same eigenvalue \(\lambda \in \mathbb{C}\), we have  \(f(v) = \lambda v\) and \(f(w)= \lambda w\).
  That is, both \(v\) and \(w\) are eigenvectors for the eigenvalue \(\lambda \in \mathbb{C}\).

  Moreover, the choice to arrange the eigenvalues in descending order is purely for convenience and not a mathematical necessity.
\end{remark}


Our approach to quantum computing is based on the study of three interconnected types of maps between Hilbert spaces.
We will discuss these both abstractly as linear map \(f \from V \to W\) and concretely as matrices \(A \in \Mat_{m,n}( \mathbb{C})\).

\begin{definition}\label{def:orthogonal-projection}
  Consider a Hilbert space \(V\).
  An \emph{orthogonal projection} onto a subspace \( U \subset V \) is a linear map \( p \from V \to V \) such that
  \begin{thmlist}
    \item \( p(v) \in U \) for all \( v \in V \), and
    \item \( \braket{p(v) - v}{u} = 0 \) for all \( v \in V \) and \( u \in U \).
  \end{thmlist}
\end{definition}

The above definition is equivalent to \( p \) being a projection (\ie \( p^2 = p \)) with the additional property that \( \im(p) \) is orthogonal to \( \ker(p) \).

\begin{lemma}\label{lem:representing-orthogonal-projection}
  A linear map \( p \from V \to V \) is an orthogonal projection onto its image if and only if, for any orthonormal basis \( \mathcal{B} \) of \( V \), the transformation matrix \( A = {_{\mathcal{B}}} f_{\mathcal{B}}\) satisfies \(A^{2} = A\) and \(A^{\dagger} = A\).
\end{lemma}

\begin{proof}
  From Lemma~\ref{lem:equivalent-characterisations-projections}, we know that \( p \) is a projection if and only if \( A^2 = A \).
  It remains to show that the orthogonality condition \( \ker(p) = (\im(p))^\perp \) is equivalent to \( A \) being self-adjoint.
  To that end, we note that \(A\) is self-adjoint if and only if \(\braket{p(v)}{w} = \braket{v}{p(w)}\) for all \(v,w \in V\).

  \begin{description}
    \item[(\( \Rightarrow \))]
    Suppose \( p \) is an orthogonal projection.
    For any \( v, w \in V \), we have
    \begin{equation*}
      \braket{p(v)}{w}
      = \braket{p(v)}{p(w) + (w - p(w))}
      = \braket{p(v)}{p(w)} + \braket{p(v)}{w - p(w)}.
    \end{equation*}
    Since \( p(v) \in \im(p) \) and \( w - p(w) \in \ker(p) \), and \( \im(p)^{\perp} = \ker(p) \), it follows that \( \braket{p(v)}{w - p(w)} = 0 \).
    Thus,
    \begin{equation*}
      \braket{p(v)}{w} = \braket{p(v)}{p(w)} = \braket{v}{p(w)},
    \end{equation*}
    which shows that \( A \) is self-adjoint.

    \item[(\( \Leftarrow \))]
    Suppose \( A \) is self-adjoint.
    For any \( v \in V \) and \( u \in \im(p) \), we have:
    \begin{equation*}
      \braket{p(v) - v}{u}
      = \braket{p(v)}{u} - \braket{v}{u}
      = \braket{v}{p(u)} - \braket{v}{u}
      = \braket{v}{u} - \braket{v}{u}
      = 0,
    \end{equation*}
    since \( p(u) = u \) for \( u \in \im(p) \).
    This shows that \( p(v) - v \in (\im(p))^\perp \), and thus \( p \) is an orthogonal projection. \qedhere
  \end{description}
\end{proof}

The second kind of map we are interested in is defined by being inner-product on a subspace of its domain.
\begin{definition}\label{def:partial-isometry}
  A \emph{partial isometry}  \(f \from (V, \braket{\blank}{\bblank})\to (W, [\blank \mid \bblank ])\) between Hilbert spaces is a linear map \(f \from V\to W\) such that
  \begin{equation*}
    [f(v) \mid f(w) ] = \braket{v}{w} \qquad \text{ for all } v, w \in \ker f^{\perp}\subset V.
  \end{equation*}
  In case \(\ker f =\{0\}\), we call \(f\) a \emph{(linear) isometry}.
\end{definition}

Suppose \( f \from V \to W \) is a partial isometry and write \(U = \ker(f)^\perp \) for the orthogonal complement of the kernel of \( f \).
There exists an orthogonal projection \( p \from V \to V \) onto \( U \) and an isometry \( g \from U \to W \) such that \( f = g \circ \pi \), where \( \pi \from V \to U \), \(v \mapsto p(v)\) is the obvious restriction map.

\begin{lemma}\label{lem:transformation-matrix-of-isometry}
  Consider a linear map \(f \from V\to W\) between an \(n\)-dimensional space \(V\) and an \(m\)-dimensional space \(W\).
  Moreover, let \(\mathcal{B}\) and \(\mathcal{C}\) be orthonormal bases of \(V\) and \(W\), respectively, and \(A={}_{\mathcal{C}}f_{\mathcal{B}}\) the transformation matrix of \(f\) with respect to \(\mathcal{B}\) and \(\mathcal{C}\).
  Then, \(f\) is a partial isometry if and only if \(P= A^{\dagger} A\) satisfies \(P^{2}= P\).
\end{lemma}
\begin{proof}
  We write \(U = \ker f^{\perp}\) for the orthogonal complement of \(\ker f\).
  First, let us show that for any linear map \(f\from V \to W\), the image of \(f^{\dagger}f\) is contained in \(U\).
  To that end, consider a vector \(v\in V\) and \(w \in \ker f\).
  We compute
  \begin{equation*}
    \braket{f^{\dagger}f(v)}{w}
    = [f(v) \mid f(w)]
    = [f(v) \mid 0 ]
    = 0,
  \end{equation*}
  implying that \(f(v) \in U\).

  Next, given \(v \in V\) and \(u\in U\), we observe that
  \begin{equation*}
    \braket{f^{\dagger}f(v) - v}{u}
    = \braket{f^{\dagger}f(v)}{u} - \braket{v}{u}
    = [f(v) | f(u)] - \braket{v}{u} = 0 \iff [f(v) | f(u)] = \braket{v}{u}.
  \end{equation*}
  Therefore, \(f\) is a partial isometry if and only if \(f^{\dagger}f\) is an orthogonal projection onto \(U\) and the claim follows from Lemma~\ref{lem:representing-orthogonal-projection} and Proposition~\ref{prop:adjoint-morph-is-adjoint-matrix}.
\end{proof}

Third kind of map, is a special case of an isometry \(f\from V\to V\)  with equal domain and codomain.

\begin{definition}\label{def:unitary-matrix}
  Consider a Hilbert space \(V\).
  A linear map \(f \from V \to V\) is a \emph{unitary map} if \(f^{\dagger} f = \id_{V} = f f^{\dagger}\).


  Similarly, a matrix \( A \in \Mat_n(\mathbb{C}) \) is called \emph{unitary} if
  \begin{equation}
    A^\dagger A = I_n = A A^\dagger.
  \end{equation}
  We denote the set of all \( n \times n \) unitary matrices by \( U(n) \).
\end{definition}
One easily verifies that upon chosing orthonormal bases, any unitary map is represented by a unitary matrix, see Exercise~\ref{exe:unitary-matrix-unitary-map}.


The concept of the dual space arises naturally in the context of Hilbert spaces.
Fixing a vector \( v \in V \) in the first entry of the inner product \( \braket{v}{\blank} \) defines a linear map \( \delta_{v}\from V \to \mathbb{C} \).
It ``measures'' the projection of any vector \( u \in V \) onto \( v \), and is a prototypical element of the dual space \( \rd*{V}= \Hom_{ \mathbb{C}}(V, \mathbb{C})\).

\begin{lemma}\label{lem:canonical-iso-to-dual-space}
  Consider a finite-dimensional Hilbert space \((V, \braket{\blank}{\bblank})\) and the
  Riesz-map
  \begin{equation*}
    d \from V \to \rd*{V}, \qquad v \mapsto \braket{v}{\blank}.
  \end{equation*}
  \begin{thmlist}
    \item There Riesz-map is well-defined and anti-linear. That is,
    \begin{equation}
      d(\lambda v+w ) = \overline{\lambda} d(v) + d(w), \qquad \text{ for all }\lambda\in \mathbb{C}, v, w \in V.
    \end{equation}
    \item The Riesz-map is a bijection.
    \item The dual \(\rd*{V}\) of \(V\) becomes a Hilbert space via the inner product
    \begin{equation*}
      \braket{\phi}{\psi} = \braket{d^{-1}(\phi)}{d^{-1}(\psi)} \qquad \text{for all } \phi, \psi \in \rd*{V}.
    \end{equation*}
  \end{thmlist}
\end{lemma}
Exercise~\ref{exe:proof-of-lem-canonical-iso-to-dual-space} provides a guided proof of this statement.


\begin{notation}\label{not:bra-ket}
  Following physics literature, we will use \emph{bra-ket notation}.
  Any vector of a Hilbert space  \((V, \braket{\blank }{\bblank})\) will be written as a \emph{ket} \(\ket v\).
  Similarly, a vector of the dual space \((\rd*{V}, \braket{ \blank }{ \bblank })\) is written as a \emph{bra} \(\bra{\phi}\).

  To keep our notation concise, we write \(f \ket v\eqdef f(\ket v)\) for any linear map \(f\from V\to W\) and \(\ket v\in V\).
  This has the additional advantage that the \emph{canonical pairing}, also called \emph{evaluation},
  between \(\ket{v}\in V\) and \(\bra{\phi}\in \rd*{V}\) will be written as
  \begin{equation*}
    \braket{\phi}{v}\eqdef \bra{\phi} \big( \ket v\big)
  \end{equation*}

  While we considered in our diagrammatic calculus vectors \(\ket{v}\) as linear maps \( \mathbb{C} \to V\), \(\lambda \mapsto \lambda \ket{v}\),
  vectors of the dual space are---by definition---linear maps \(\bra{\phi} \from V \to \mathbb{C}\).

  This allows us to form diagrams of the following type:
  \begin{thmlist}[leftmargin=6.5em + \parindent, align = left,labelwidth = 6.5em]
    \item[ \textit{initialisation}:] We consider a vector \(\ket{v}\in V\);
    \item[ \textit{transformation}:] We apply a linear map (often a partial isometry)  \(f\from V\to W\);
    \item[ \textit{evaluation}:] We evaluate the outcome against a vector \(\bra{\psi}\in \rd*{V}\) and obtain a scalar.
  \end{thmlist}
  \begin{equation}
    \inputtikz{init-transform-eval}
  \end{equation}
\end{notation}

\subsection{States}

In the simplest case, quantum algorithms follow the same three-step pattern highlighted in the previous string diagram.
First, we prepare a state \(\ket{v}\), then we apply a unitary transformation \(f\), and finally, we measure the outcome.
Even at this superficial level we have an oddity to consider.
Everything we did was \emph{deterministic}: applying a linear transformation to some vector will always yield the same result.
But in quantum computing we will utilise the probabilistic nature of quantum mechanics\footnote{
  The proclaimed advantages of quantum computing over classical computing arise not just from probabilistic effects, but from the intricate interplay of superposition, entanglement, and interference in compound quantum systems.
  In the language of complexity theory, the class \(\BQPclass\) of polynomial-time problems solvable by quantum computers is believed to properly include those solvable by probabilistic Turing machines, denotes \(\BPPclass\)
}.

As a motivational example we apply the the Hadamard gate \(H= \tfrac{1}{\sqrt{2}}
\begin{psmallmatrix}
  1 & 1 \\ 1 & -1
\end{psmallmatrix}
\) to the vector \(\ket{s} = \ket{0} \in \mathbb{C}^{2}\).
The resulting vector \(\ket{t} = H\ket{s}= \tfrac{1}{\sqrt{2}}(\ket 0 + \ket 1)\) is neither \(\ket{0}\) nor \(\ket{1}\).
Instead, following the Copenhagen interpretation its probability to be \(\ket{0}\) is
\begin{equation*}
  \|\braket{0}{t}\|^{2} = \|\tfrac{1}{\sqrt{2}}\|^{2} = \tfrac{1}{2}
\end{equation*}
and similarly \(\ket{t}\) is with probability \(\tfrac{1}{2}\) in state \(\ket{1}\).
While we will discuss measurements in more detail in Section~\ref{sec:measurements-and-the-born-rule}, these computations motivate two observations:
First, for the probabilities to sum to \(1\) the vector \(\ket{t}\) must have length \(1\).
Second, if we replace \(\ket{t}\) with \(e^{2\pi \alpha}\ket{t}\) the outcomes of measurement remain unchanged \ie such scalars are not ``witnessed'' by measurements.

\begin{definition}\label{def:states}
  Consider a Hilbert space \(V\).
  \begin{thmlist}
    \item A vector \(\ket{v}\in V\) is \emph{normalised} if \(\|\ket{v}\| = 1\).
    \item Two normalised vectors \(\ket{v},\ket{w}\in V\) are \emph{equivalent up to global phase} if there exists an \(\alpha\in \mathbb{R}\) such that \(e^{2\pi i \alpha}\ket{v} = \ket{w}\).
    \item The equivalence class \([\ket{v}]\eqdef \{ e^{2\pi i \alpha} \ket{v}\}\) of a normalised vector \(\ket{v} \in V\) is called a \emph{(pure) state}.
  \end{thmlist}
\end{definition}

In quantum mechanics,the choice of basis \(\mathcal{B}\) for a Hilbert space is often motivated by the physical interpretation of the system.
For example, in the context of qubits, the standard basis \( \{\ket{0}, \ket{1}\} \) corresponds to classical binary states, and measurements are typically performed with respect to this basis.
Superposition describes the idea of having a state \([\ket{v}]\) that cannot be represented by a single basis vector alone.

\begin{definition}\label{def:superposition} %TODO ket-ify the vectors
  Let \(V\) be a Hilbert space and \(\mathcal{B}= (\ket{b_1}, \dots, \ket{b_n})\) a basis.
  A state \([\ket{v}]\) is in a \emph{superposition} with respect to \(\mathcal{B}\) if  \(\ket{v} = \lambda_{1} \ket{b_{1}} + \dots +\lambda_{n}\ket{b_{n}}\) and \(\lambda_{i}, \lambda_{j}\neq 0\) for some \(i\neq j\).
\end{definition}

We will now discuss two equivalent ways of describing states, eliminating the ambiguity introduced by global phase factors.

\begin{lemma}\label{lem:projective-space-and-phases}
  The canonical map
  \begin{equation}
    \mathrm{pro} \from \{\text{states in } V\} \to\{L\subset V\mid L \text{ is a 1-dimensional subspace}\} \qquad
    [\ket{v}]\mapsto \mathbb{C} \ket{v} %or some \otimesher nate \instead of pro...
  \end{equation}
  is a bijection.
\end{lemma}
\begin{proof}
  We construct the inverse map explicitly.
  For each \(1\)-dimensional subspace \( L \subset V \), choose a non-zero vector \( \ket{v} \in L \setminus \{0\} \).
  There exists a unique positive real number \( \lambda \in \mathbb{R}_{>0} \) such that \( \|\lambda \ket{v}\| = 1 \).
  We define the inverse map via
  \begin{equation*}
    \mathrm{st} \from \{L\subset V\mid L \text{ is a 1-dimensional subspace}\} \to \{\text{states in } V\} \qquad L \mapsto [\lambda \ket{v}].
  \end{equation*}
  Given a \(1\)-dimensional subspace \( L = \mathbb{C} \ket{v} \), we compute
  \begin{equation*}
    \mathrm{pro}(\mathrm{st}(L)) = \mathrm{pro}([\lambda \ket{v}]) = \mathbb{C} \lambda \ket{v} = \mathbb{C} \ket{v}.
  \end{equation*}

  Conversely, if \([\ket{v}]\) is a state, we get
  \begin{equation*}
    \mathrm{st}(\mathrm{pro}[\ket{v}]) = \mathrm{st}( \mathbb{C} \ket{v}) = [\lambda \ket{v}']
  \end{equation*}
  where \(\ket{v}' =\beta \ket{v}\).
  Since \( \|\ket{v}\| = 1 \), we have
  \begin{equation*}
    \|\beta\| = \|\beta\| \;\|\ket{v}\| = \|\beta \ket{v}\| = \|\lambda \ket{v}'\| = 1.
  \end{equation*}
  Thus, \( \beta = e^{2\pi i \alpha} \) for some \( \alpha \in \mathbb{R} \), and \( [\ket{v}] = [\lambda \ket{v}'] \).
\end{proof}

The set \( \mathbb{C}P^{n-1} =\{L \subset \mathbb{C}^{n}\mid \dim L = 1\} \) is called the \emph{complex projective space}.
By the above result, states in \( \mathbb{C}^{n}\) can be thought of as points in the projective space.
\bigskip

A different perspective on states is given by viewing them as matrices.

\begin{definition}\label{def:mixed-state-density-matrix}
  A \emph{mixed state} is a self-adjoint, positive semidefinite map \(\rho\from V\to V\) satisfying \(\tr \rho = 1\).

  A \emph{density matrix} is a self-adjoint, positive definite matrix \(A \in \Mat_{n}( \mathbb{C})\) with \(\tr(A)=1\).
\end{definition}

\begin{remark}\label{rmk:density-matrices-are-mixed-states}
  Let \(\mathcal{B}=(b_{1}, \dots , b_{n})\) be an orthonormal basis of \(V\).
  By construction, a linear map \(f\from V\to V\) is a mixed state if and only if \(_{\mathcal{B}}f_{\mathcal{B}} \in \Mat_{n}{ \mathbb{C}}\) is a density matrix.
\end{remark}


Consider any  vector \(\ket{v}= a \ket{0} + b \ket{1} \in \mathbb{C}^{2}\).
A direct computation shows that
\begin{equation*}
  \rho_{\ket v}\eqdef\ket{v}\bra{v} \from \mathbb{C}^{2} \to \mathbb{C}^{2}, \qquad \ket{w} \mapsto (\braket{v}{w})\ket{v}.
\end{equation*}
is self-adjoint and positive semidefinite.
Moreover the trace of \(\rho_{\ket{v}}\) is
\begin{align*}
  \tr(\rho_{\ket v})
  & = \braket{0}{\rho_{\ket v}(\ket 0)} + \braket{1}{\rho_{\ket v}(\ket{1})}
    = (\braket{v}{0}\braket{0}{v}) + \braket{v}{1}\braket{1}{v} \\
  & = \|\braket{v}{0}\|^{2} + \|\braket{v}{1}\|^{2}
    = a^{2} + b^{2} = \|\ket{v}\|^{2}.
\end{align*}
Thus, \(\tr(\rho_{\ket v})=1\) is equivalent to \(\ket{v}\) being normalised.

\begin{proposition}\label{prop:density-matrix-perspective}
  There is an injective map
  \begin{equation*}
    \rho:\{\text{states of }V\}\to\{ \text{density matrices on } V\}, \qquad \ket{v} \mapsto \rho_{\ket{v}} \eqdef \ket{v}\bra{v}
  \end{equation*}
  Its image coincides with the rank 1 density matrices.
\end{proposition}
\begin{proof}
  Let \( \mathcal{B} = (\ket{v}, \ket{b_2}, \dots, \ket{b_n}) \) be an orthonormal basis containing \( \ket{v} \). The transformation matrix of \( \rho_{\ket{v}} \) with respect to \( \mathcal{B} \) is \( \mathrm{diag}(1, 0, \dots, 0)\), which is clearly self-adjoint, positive semidefinite, and of rank 1.

  Conversely, assume that \(\varrho \in \End(V)\) is a rank 1 density matrix.
  By the spectral Theorem~\ref{thm:spectral-theorem}, there exists an orthonormal basis \(\mathcal{B}= (b_{1}, b_{2}, \dots , b_{n})\) of eigenvalues \(\lambda_{1}\geq  \dots \geq \lambda_{n} \geq 0\) of \(\varrho\).
  Since \(\rk \varrho\) coincides with the number of non-zero eigenvalues and \(\tr(\varrho) = \lambda_{1} + \dots + \lambda_{n} = 1\), we have \(\lambda_{1} = 1\) and \(\lambda_{2} = \dots = \lambda_{n} = 0\).
  In particular \(_{\mathcal{B}}\varrho_{\mathcal{B}} = \mathrm{diag}(1,0, \dots , 0) = {}_{\mathcal{B}}{\rho_{\ket{b_{1}}}}_{\mathcal{B}}\) which implies that \(\varrho = \rho_{\ket{b_{1}}}\).


  To show injectivity of \(\rho\), suppose \( \rho_{\ket{v}} = \rho_{\ket{w}} \) for two normalised vectors \(\ket{v}, \ket{w} \in V\).
  For any \( \ket{x} \in V \), we obtain
  \begin{equation*}
    \ket{v} \braket{v}{x}
    = \rho_{\ket{v}}(\ket{x})
    = \rho_{\ket{w}}(\ket{x})
    = \ket{w} \braket{w}{x}.
  \end{equation*}
  Choosing \( \ket{x} = \ket{v} \), we get
  \begin{equation*}
    \ket{v} = \ket{v} \braket{v}{v} = \ket{w} \braket{w}{v}.
  \end{equation*}
  Setting \(\lambda = \braket{w}{v}\), we observe that  \(\lambda \ket{w}= \ket{v}|\).
  Moreover, since \(\ket{w}\) and \(\ket{v}\) are normalised, we obtain
  \begin{equation*}
    \|\lambda\| = \|\lambda\| \; \|\ket{w}\| = \|\lambda \ket{w}\| = \|v\| = 1
  \end{equation*}
  implying that \([\ket v] = [\ket w]\).
\end{proof}

In string diagrams, the density matrix of a pure state is given as follows.
\begin{equation*}
  \inputtikz{density-matrix-example}
\end{equation*}


\subsubsection*{Qubits and the Bloch sphere}\label{sec:qubits-and-the-bloch-sphere}

As the name suggests, a \emph{qubit} is the quantum analogue of a bit.
It is modelled by a system with two degrees of freedom andtypically represented by the Hilbert space \( \mathbb{C}^2 \).

\begin{definition}\label{def:qubit}
  A \emph{(pure) qubit state} is a state of \( \mathbb{C}^2 \).
\end{definition}

As discussed in Example~\ref{ex:c2-as-Hilbert-space}, the space \( \mathbb{C}^2 \) has the \emph{computational basis} or \emph{Z-basis}  \((\ket{0}, \ket{1})\) with
\begin{equation*}
  \ket{0} = \begin{psmallmatrix} 1 \\ 0 \end{psmallmatrix}, \qquad
  \ket{1} = \begin{psmallmatrix} 0 \\ 1 \end{psmallmatrix}.
\end{equation*}
These basis vectors can represent physical quantities such as spin states.
\bigskip

Before moving on to more complex systems, we will study the geometry of a qubit in detail.
To do this, we use that pure states are precisely the rank-1 density matrices, see Proposition~\ref{prop:density-matrix-perspective}, and that \( \Mat_2(\mathbb{C}) \) has a basis in terms of the identity matrix and the Pauli matrices of Definition~\ref{def:Pauli-matrices}, see Exercise~\ref{ex:props-of-Pauli-matrices}.

\begin{example}\label{ex:density-matrices-of-some-bases}
  As a warm-up, we compute the density matrices of the \( Z \)-, \( X \)-, and \( Y \)-bases of \( \mathbb{C}^2 \) discussed in Example~\ref{ex:c2-as-Hilbert-space}.

  For the \( Z \)-basis \( (\ket{0}, \ket{1}) \), we have
  \begin{equation*}
    P_{\ket{0}} = \ket{0}\bra{0} = \begin{pmatrix} 1 & 0 \\ 0 & 0 \end{pmatrix} = \tfrac{1}{2}(I_2 + \sigma_z), \quad
    P_{\ket{1}} = \ket{1}\bra{1} = \begin{pmatrix} 0 & 0 \\ 0 & 1 \end{pmatrix} = \tfrac{1}{2}(I_2 - \sigma_z).
  \end{equation*}

  The \( Y \)-basis \( (\ket{i+}, \ket{i-}) \), with \( \ket{i\pm} = \frac{1}{\sqrt{2}}(\ket{0} \pm i\ket{1}) \), yields
  \begin{equation*}
    P_{\ket{i+}} = \ket{i+}\bra{i+} = \tfrac{1}{2} \begin{pmatrix} 1 & -i \\ i & 1 \end{pmatrix} = \tfrac{1}{2}(I_2 + \sigma_y), \quad
    P_{\ket{i-}} = \ket{i-}\bra{i-} = \tfrac{1}{2} \begin{pmatrix} 1 & i \\ -i & 1 \end{pmatrix} = \tfrac{1}{2}(I_2 - \sigma_y).
  \end{equation*}


  Finally, for the \( X \)-basis \( (\ket{+}, \ket{-}) \), where \( \ket{\pm} = \frac{1}{\sqrt{2}}(\ket{0} \pm \ket{1}) \), we get
  \begin{equation*}
    P_{\ket{+}} = \ket{+}\bra{+} = \tfrac{1}{2} \begin{pmatrix} 1 & 1 \\ 1 & 1 \end{pmatrix} = \tfrac{1}{2}(I_2 + \sigma_x), \quad
    P_{\ket{-}} = \ket{-}\bra{-} = \tfrac{1}{2} \begin{pmatrix} 1 & -1 \\ -1 & 1 \end{pmatrix} = \tfrac{1}{2}(I_2 - \sigma_x).
  \end{equation*}
\end{example}

We will now describe arbitrary states as linear combinations of \(I_{2}\) and the Pauli matrices.

\begin{lemma}\label{lem:characterising-the-qubit}
  A matrix \(P \in \Mat_{2}( \mathbb{C})\) is a density matrix of a pure state if and only if there are real numbers \(r_{x}, r_{y}, r_{z}\in \mathbb{R}\) with \(r_{x}^{2}+ r_{y}^{2}+r_{z}^{2} = 1\) such that
  \begin{equation*}
    P = \tfrac{1}{2}(I_{2} +r_{x}\sigma_{x} + r_{y}\sigma_{y}+ r_{z}\sigma_{z})
  \end{equation*}
\end{lemma}
\begin{proof}
  Let \(P=\tfrac{1}{2}(r_{0}I_{2} +r_{x}\sigma_{x} + r_{y}\sigma_{y}+ r_{z}\sigma_{z})\in \Mat_{2}( \mathbb{C})\) be any matrix.
  Since Pauli matrices are self-adjoint, \(P = P^{\dagger}\) if and only if the scalars \(r_{0}, r_{x}, r_{y}, r_{z}\) are real.
  Moreover, since the trace is linear and Pauli matrices are trace-less, \(\tr(P) = 1\) is equivalent to \(r_{0} = 1\).
  We now compute
  \begin{align*}
    (2P)^{2}
    & = (I_{2} +r_{x}\sigma_{x} + r_{y}\sigma_{y}+ r_{z}\sigma_{z})(I_{2} +r_{x}\sigma_{x} + r_{y}\sigma_{y}+ r_{z}\sigma_{z})\\
    & = (1+ r_{x}^{2} + r_{y}^{2}+ r_{z}^{2}) I_{2}
      + (2r_{x} + i r_{y}r_{z} -i r_{y}r_{z} ) \sigma_{x}\\
    &\qquad  + (2r_{y} + -i r_{x}r_{z} + i r_{x}r_{z}) \sigma_{y}
      + (2r_{z}+ i r_{x}r_{y} - i r_{x}r_{y} ) \sigma_{z} \\
    & = 4P
  \end{align*}
  Implying that \(P^{2}=P\) if and only if \((r_{x}^{2} + r_{y}^{2}+ r_{z}^{2}) =1\).
\end{proof}

Therefore, we can model the states of \( \mathbb{C}^{2}\) as unit vectors of \( \mathbb{R}^{3}\).

\begin{definition}\label{def:bloch-sphere}
  We call  \(S^{2}=\{(r_{x}, r_{y}, r_{z})^{T} \in \mathbb{R}^{3} \mid r_{x}^{2}+ r_{y}^{2}+ r_{z}^{2}=1\}\) the \emph{Bloch sphere}.
\end{definition}

The following corollary is a reformulation of Lemma~\ref{lem:characterising-the-qubit}

\begin{corollary}\label{cor:Bloch-sphere-pure-state}
  There is a bijection between the Bloch sphere and the states of \( \mathbb{C}^{2}\) given by
  \begin{equation*}
    S^{2} \mapsto\{ p \in \Mat_{2}( \mathbb{C}) \mid p \text{ density matrix, } \rk(p) = 1\}, % <- the last bracket could be two lines below each \otimesher... replace \mid \blank by vert...
    \qquad
    \begin{pmatrix}
      r_{x} \\
      r_{y}\\
      r_{z}
    \end{pmatrix}
    \mapsto \frac{1}{2}(I_{2} + r_{x}\sigma_{x} + r_{y}\sigma_{y} + r_{z} \sigma_{z}).
  \end{equation*}
\end{corollary}

The figure below show the Bloch sphere with \(\ket{0}\) and \(\ket{1}\) as north and south pole and the \(X\)- and \(Y\)-basis as points on the equator.

\begin{equation*}
  \inputtikz{bloch-sphere}
\end{equation*}


\subsection{Compound systems}\label{sec:compound-systems}
Having established a single qubit, we now want to discuss the mathematical foundations for multi-qubit systems.
In order to see what kind of construction we need, we can look at the translation between the classical and the quantum world.
\begin{equation*}
  \inputtikz{motivation-tensor-product}
\end{equation*}
In other words we need a construction that combines vector spaces \(V\) and \(W\) such that the resulting space \(T\) has dimension \(\dim V \cdot \dim W\).
This space \(T\) will be the tensor product of \(V\) and \(W\).

\subsubsection{Tensor products of vector spaces}\label{sec:tensor-products-of-vector-spaces}

In order to get some intuition for what will become the tensor product \(T\) of two vector spaces \(V\) and \(W\), suppose \(\mathcal{B}\) is a basis for \(V\) and \(\mathcal{C}\) is a basis of \(W\).
The requirement \(\dim T = \dim V \cdot \dim W\) suggests that \(T\) has a basis behaving like the Cartesian  product \( \mathcal{B} \times \mathcal{C} \).
To understand \(\mathcal{B}\times \mathcal{C}\) better, let us consider some vector space \(X\) and an assignment \(f\from \mathcal{B}\times \mathcal{C} \to X \).
Since \(\mathcal{B}\) is a basis of \(V\) and \(\mathcal{C}\) is a basis of \(W\) we can extend \(f\) to a map \(\widehat{f} \from V\times W \to X\) which is linear \emph{in each argument}.

\begin{definition}\label{def:bilinear-map}
  Let \(V,W,X\) be three vector spaces.
  A map \(b \from V\times W\to X\) is called \emph{bilinear} if for all \(\lambda, \mu \in \mathbb{C}\), \(v,v' \in V\), and \(w,w \in W\) we have
  \begin{gather*}
    b(v+v', w+w') = b(v, w) + b(v, w') + b(v', w) + b(v', w'), \\
    b(\lambda v, w) = \lambda b(v, w) \quad  \text{ and }\quad
    b(v, \mu w) = \mu b(v, w).
  \end{gather*}
\end{definition}
Alternatively, \(b\from V\times W \to X\) is a bilinear map if and only if the associated map \(B \from V\to \Hom_{ \mathbb{C}}(W,X)\), \(v\mapsto b(v, \blank)\) is linear, see Exercise~\ref{exe:bilinear-vs-linear-hom}.

\begin{remark}\label{rmk:bilinear-uniquely-def-by-values-on-basis}
  Let \( V \) and \( W \) be vector spaces with bases \( \mathcal{B} \) and \( \mathcal{C} \), respectively, and let \( X \) be a vector space.
  A direct computation shows that any bilinear map \(f\from V\times W \to X\) is uniquely determined by its value on \(\mathcal{B}\times \mathcal{C}\).
  Conversely, any function \( g \from \mathcal{B} \times \mathcal{C} \to X \) extends uniquely to a bilinear map
  \begin{equation*}
    \widehat{g} \from V \times W \to X, \quad \big(\sum_{b\in \mathcal{B}}\lambda_{b} b, \sum_{c \in \mathcal{C}} \mu_{c}c\big) \mapsto \sum_{b\in \mathcal{B}} \sum_{c\in \mathcal{C}} \lambda_{b} \mu_{c}\; g(b,c).
  \end{equation*}
  Thus, we obtain a bijection
  \begin{align*}
    \{ \text{bilinear maps } V \times W \to X \} &\longleftrightarrow \{ \text{set maps } \mathcal{B} \times \mathcal{C} \to X \}, \\
    f &\mapsto \left( (b, c) \mapsto f(b, c) \right), \\
    \widehat{g} &\longmapsfrom g.
  \end{align*}
\end{remark}
If functions \(\mathcal{B}\times \mathcal{C}\to X\) determine bilinear maps and the tensor product of \(V\) and \(W\) is ought to have a basis behaving like \(\mathcal{B}\times \mathcal{C}\), it needs to turn bilinear maps into linear ones.

\begin{definition}\label{def:tensor-product}
  The \emph{tensor product} of two vector spaces \(V, W\) is a pair \((V\otimes W, \blank \otimes \bblank)\) comprising a vector space \(V\otimes W\) and a bilinear map
  \begin{equation*}
    (\blank \otimes \bblank)\from V \times W \to V\otimes W, \qquad (v,w) \mapsto v\otimes w
  \end{equation*}
  such that for every vector space \(X\) and  bilinear map \(b \from V \times W \to X\) there exists a unique linear map \(f \from V\otimes W \to X\) satisfying
  \begin{equation*}
    b(v, w) = f(v \otimes w) \qquad \text{for all } v \in V, w \in W.
  \end{equation*}
\end{definition}

We can express the defining condition of the tensor product in terms of the commuting diagram:
\begin{equation*}
  \begin{tikzcd}[ampersand replacement=\&,cramped]
    {V\times W} \&\& X \\
    \\
    {V\otimes W}
    \arrow["b \text{ (bilinear)}", from=1-1, to=1-3]
    \arrow["{\blank \otimes \bblank}"', from=1-1, to=3-1]
    \arrow["{\exists! f}"', dashed, from=3-1, to=1-3]
  \end{tikzcd}
\end{equation*}

A priori, we do not know whether tensor products exist and, in case they do, if they are unique.
However, we get a first hint towards a potential construction of tensor products by reconsidering the analogy between classical bits and quantum bits.
In the classical world, we have 4 possible \(2\)-bit words: \(00\), \(01\), \(10\), and  \(11\).
These words are all possible concatenations of the letters \(0\) and \(1\).
For qubits, we want a vector space with a basis \(\ket{00}\), \(\ket{01}\), \(\ket{10}\), \(\ket{11}\), and this basis should result from ``concatenating'' the basis \((\ket{0}, \ket{1})\) of \( \mathbb{C}^{2}\) with itself.
We have already seen such a construction  in the context of free vector spaces discussed in~Example~\ref{ex:free-vector-space}.

\begin{remark}\label{rmk:realisation-of-tensor-products}
  Consider a  pair of vector spaces \((V,W)\) with bases \(\mathcal{B}= (b_{1}, \dots , b_{n})\) and \(\mathcal{C}= (c_{1}, \dots , c_{m})\), respectively.
  By Example~\ref{ex:free-vector-space}, the set of all maps \(\mathrm{Map}(\mathcal{B}\times \mathcal{C}, \mathbb{C})\) becomes a vector space under pointwise addition.
  Given \(1\leq i \leq n\) and \(1 \leq j \leq m\), we write
  \begin{equation*}
    b_{i} \boxtimes c_{j} \from \mathcal{B} \times \mathcal{C} \to \mathbb{C}, \qquad (b_{k}, c_{l}) \mapsto
    \begin{cases}
      1 & i=k, j= l \\
      0 & \text{ otherwise.}
    \end{cases}
  \end{equation*}
  By construction \((b_{1}\boxtimes c_{1}, \dots , b_{1}\boxtimes c_{m}, b_{2} \boxtimes c_{1}, \dots , b_{n} \boxtimes c_{m})\) is a basis of \(\mathrm{Map}(\mathcal{B}\times \mathcal{C}, \mathbb{C})\) allowing us to express
  \begin{equation*}
    \mathrm{Map}(\mathcal{B}\times \mathcal{C}, \mathbb{C}) =  \big\{ \sum\nolimits_{i=1}^{n}\sum\nolimits_{j=1}^{m}\lambda_{ij}(b_{i}\boxtimes c_{j})\mid \lambda_{ij} \in \mathbb{C}\big\}.
  \end{equation*}
\end{remark}

\begin{theorem}\label{thm:tensor-product-existence-and-uniqueness}
  Let \( V \) and \( W \) be vector spaces with bases \( \mathcal{B} = (b_1, \dots, b_n) \) and \( \mathcal{C} = (c_1, \dots, c_m) \), respectively.
  Retaining the notation of the previous remark, we write
  \begin{equation*}
    V\boxtimes W \eqdef \mathrm{Map}(\mathcal{B}\times \mathcal{C}, \mathbb{C}) \cong \big\{ \sum\nolimits_{i=1}^{n}\sum\nolimits_{j=1}^{m}\lambda_{ij}(b_{i}\boxtimes c_{j})\mid \lambda_{ij} \in \mathbb{C}\big\}
  \end{equation*}
  and define the map
  \begin{equation*}
    (\blank \boxtimes \bblank )\from V\times W \to V \boxtimes W, \quad  (\lambda_{1} b_{1} + \dots \lambda_{n} b_{n})\boxtimes (\mu_{1}c_{1}+ \dots \mu_{m}c_{m}) = \sum \lambda_{i}\mu_{j}\; b_{i}\boxtimes c_{j}.
  \end{equation*}
  The following statements hold:
  \begin{thmlist}
    \item The map \( \blank \boxtimes \bblank \) is bilinear.
    \item The pair \( (V \boxtimes W, \blank \boxtimes \bblank) \) is a tensor product of \( V \) and \( W \).
    \item For any other tensor product \( (V \otimes W, \blank \otimes \bblank) \) of \( V \) and \( W \), there exists a unique linear isomorphism \( f \from V \boxtimes W \to V \otimes W \) such that \( f(v \boxtimes w) = v \otimes w \) for all \( v \in V \) and \( w \in W \).
  \end{thmlist}
\end{theorem}
\begin{proof}
  \begin{thmlist}
    \item The map \( \blank \boxtimes \bblank \) is the bilinear extension of the map
    \begin{equation*}
      \mathcal{B}\times \mathcal{C} \to V\boxtimes W, \qquad (b_{i}, c_{j})\mapsto b_{i}\boxtimes c_{j}
    \end{equation*}
    and therefore in particular bilinear.
    \item Given a bilinear map \( g \from V \times W \to X \), we define the linear map \( f \from V \boxtimes W \to X \) by \( f(b_i \boxtimes c_j) = g(b_i, c_j) \).
    Note that
    \begin{equation*}
      (f\circ( \blank \boxtimes \bblank)) (b_{i}, c_{j}) =  f(b_{i} \boxtimes c_{j}) = g(b_{i}, c_{j}),
    \end{equation*}
    implying that \(f(v\boxtimes w) = g(v,w)\) for all \(v\in V\), \(w\in W\).
    In case \(f' \from V\boxtimes W \to X \) also satisfies \(f'(v \boxtimes w) = g(v,w)\) for all \(v\in V\) and \(w\in W\), we in particular have \(f'(b_{i}\boxtimes c_{j}) = f(b_{i} \boxtimes c_{j})\) for all \(b_{i}\in \mathcal{B}\) and \(c_{j}\in \mathcal{C}\), implying that \(f = f'\).

    \item Since \( \blank \otimes \bblank \from V \times W \to V \otimes W \) is bilinear, there exists a unique linear map \( k \from V \boxtimes W \to V \otimes W \) such that \( k(v \boxtimes w) = v \otimes w\).
    Similarly, since \(( V \otimes W, \blank \otimes \bblank) \) is a tensor product and \( \blank \boxtimes \bblank \from V\times W \to V\boxtimes W\) is bilinear, there exists a unique linear map \( r \from V \otimes W \to V \boxtimes W \) such that \( r(b_i \otimes c_j) = b_i \boxtimes c_j \).
    In order to show that \(r\circ k = \id_{V\boxtimes W}\), we observe that for all \(v\in V\) and \(w\in W\) we have
    \begin{equation*}
      r\circ k (v\boxtimes w) = r(v\otimes w) = v \boxtimes w = \id_{V\boxtimes W} (v\boxtimes w).
    \end{equation*}
    By uniqueness, we get \(r\circ k = \id_{V\boxtimes W}\).
    One shows \(k \circ r = \id_{V\otimes W}\) analogously. \qedhere
  \end{thmlist}
\end{proof}

 Due to the above theorem, tensor products are unique up to unique isomorphism.
With a small abuse of notation, we will speak of the tensor product of \(V\) and \(W\) and denote it by \((V\otimes W, \blank \otimes \bblank)\).

\begin{definition}\label{def:elementary-tensors}
  Let \((V \otimes W, \blank \otimes \bblank)\) be the tensor product of \(V \otimes W\).
  An element \(x\in V\otimes W\) is called a \emph{tensor}.
  In case there are \(v\in V\) and \(w\in W\), such that \(v\otimes w = x\), we refer to \(x\) as an \emph{elementary tensor}.
\end{definition}

Not all tensors in \( V \otimes W \) are elementary tensors
However, every tensor in \( V \otimes W \) can be expressed as a finite sum of elementary tensors, thus endowing elementary tensors with a quite important role.

\begin{remark}\label{rmk:bilinearity-on-the-level-of-elementary-tensors}
  The bilinearity of the  map \( \blank \otimes \bblank \from V\times W \to V\otimes W\) translates directly to computation rules for elementary tensors.
  Explicitly, for all \(\lambda \in \mathbb{C}\), \(v, v' \in V\), and \(w, w' \in W\), we have
  \begin{subequations}
    \begin{gather}
      (v+v')\otimes (w+w') = v\otimes w + v \otimes w' + v' \otimes w + v' \otimes w', \label{eq:tensor-additive}\\
      (\lambda v) \otimes w = \lambda(v\otimes w) = v \otimes (\lambda w).\label{eq:tensor-scalar-mult}.
    \end{gather}
  \end{subequations}
  It is customary to define a linear map \( f \from V \otimes W \to X \) by specifying its action on elementary tensors
  \begin{equation*}
    f \from V \otimes W \to X, \quad v \otimes w \mapsto f(v \otimes w),
  \end{equation*}
  with the implicit understanding that \( f \) is extended linearly to all of \( V \otimes W \).
  That is, for any tensor \( t = \sum_{i=1}^k v_i \otimes w_i \), the map \( f \) is defined by
  \begin{equation*}
    f(t) = \sum_{i=1}^k f(v_i \otimes w_i).
  \end{equation*}

  \textbf{Warning:} While this notation is suggestive and compact, it can be misleading.
  When defining a map \( f \) in this way, one must verify that \( f \) is well-defined by ensuring that it respects the bilinearity relations of Equations~\eqref{eq:tensor-additive} and~\eqref{eq:tensor-scalar-mult}.
\end{remark}


Let us now examine the tensor product \( \mathbb{C}^2 \otimes \mathbb{C}^2 \).
\begin{example}\label{ex:c2-otimes-c2}
  Given vectors \(\ket{v}, \ket{w} \in \mathbb{C}^{2}\), we set \(\ket{vw} \eqdef \ket{v} \otimes \ket{w}\).
  By Theorem~\ref{thm:tensor-product-existence-and-uniqueness} and Remark~\ref{rmk:realisation-of-tensor-products}, the tensor product \( \mathbb{C}^2 \otimes \mathbb{C}^2 \) is four-dimensional and has a basis \( (\ket{00}, \ket{01}, \ket{10}, \ket{11}) \).
  Thus, it will serves as a quantum analogue of a 2-bit system.
\end{example}

We now want to iterate this procedure to describe multi-qubit systems.

\begin{remark}\label{rmk:iterated-tensor-products}
  Let \( V, W, X \) be vector spaces.
  Although the spaces \( (V \otimes W) \otimes X \) and \( V \otimes (W \otimes X) \) are distinct, there exists a canonical isomorphism between them.
  By a slight abuse of notation, we will simply write \( V \otimes W \otimes X \) for either space.
  This construction can be iterated further, allowing us to define the \( n \)-fold tensor product of \( V \) as
  \begin{equation*}
    V^{\otimes n} = \underbrace{V \otimes V \otimes \cdots \otimes V}_{n \text{ times}},
  \end{equation*}
  with the convention that \( V^{\otimes 0} \eqdef \mathbb{C} \).

  If \( \mathcal{B} = \{b_1, b_2, \ldots, b_k\} \) is a basis of \( V \), then the set of all tensors of the form
  \begin{equation*}
    b_{i_1} \otimes b_{i_2} \otimes \cdots \otimes b_{i_n},
  \end{equation*}
  where \( i_1, i_2, \ldots, i_n \in \{1, 2, \ldots, k\} \), forms a basis for \( V^{\otimes n} \).
\end{remark}

We can now apply this to describe \( n \)-qubit systems.

\begin{example}\label{ex:n-qubits}
  Given \(\ket{v_{1}}, \dots , \ket{v_{n}} \in \mathbb{C}^{2} \), we write
  \begin{equation}
    \ket{v_{1} v_{2} \dots v_{n}} \eqdef v_{1} \otimes v_{2} \otimes \dots \otimes v_{n}.
  \end{equation}
  Then, the \(2^{n}\)-dimensional space \( (\mathbb{C}^{2})^{\otimes n}\)  has a basis \(\mathcal{B}= (\ket{s_{1} s_{2} \dots s_{n}})_{s\in\{0,1\}^{n}}\).
\end{example}

With tensor products, we can consider multiple qubits, allowing us to complete the infographic at the beginning of the section.
\begin{equation*}
  \inputtikz{motivation-tensor-product-complete}
\end{equation*}


\begin{remark}\label{rmk:tensor-product-string-diags}
  The tensor product is a new operation between vector spaces.
  Since we use in string diagrams the horizontal axis for composition, we will use the vertical axis for tensoring
  \begin{equation*}
    \inputtikz{tensor-product-of-vector-spaces}
  \end{equation*}
  This notation suggests that we should also be able to tensor linear maps, a topic we will discuss later.
\end{remark}

\begin{lemma}\label{lem:unitary-on-tensor-product}
  Let \( V \) and \( W \) be finite-dimensional Hilbert spaces with inner products \( \braket{\blank}{\bblank}_V \) and \( \braket{\blank}{\bblank}_W \), respectively.
  The tensor product \( V \otimes W \) becomes a Hilbert space with the inner product \(\braket{\blank}{\bblank}_{V \otimes W} \from (V \otimes W) \times (V \otimes W) \to \mathbb{C}\) defined on elementary tensors by
  \begin{equation*}
    \braket{v \otimes w}{v' \otimes w'}_{V \otimes W} = \braket{v}{v'}_V \braket{w}{w'}_W \qquad \text{for all } v, v' \in V \text{ and }w, w' \in W.
  \end{equation*}
\end{lemma}
\begin{proof}
  To show that \( \braket{\blank}{\bblank}_{V \otimes W} \) is well-defined, we use the universal property of tensor products.
  To that end, we want to utilise a small trick.
  Recall that for any vector space \(V\), we can define a ``conjugate'' space \(\overline{V}\) with the same underlying set and addition as \(V\) but the conjugate scalar multiplication \(\lambda \cdot v \eqdef \overline{\lambda}v\) for all \(\lambda \in \mathbb{C}\)  and \(v \in V\).
  To better distinguish between \(V\) and \(\overline{V}\), we write \(\overline{v} \in \overline{V}\) for the corresponding element \(v\in V\).

  By Exercise~\ref{exe:translating-bilinear-to-sesquilinear},
  and the universal property of tensor products, we obtain linear maps
  \begin{align*}
    [ \blank \mid \bblank]_{V} &\from \overline{V}\otimes V \to \mathbb{C}, \qquad &&\overline{v} \otimes v' \mapsto \braket{v}{v'}_{V}\\
    [ \blank \mid \bblank]_{W} &\from \overline{W}\otimes W \to \mathbb{C}, \qquad &&\overline{w} \otimes w' \mapsto \braket{w}{w'}_{W}
  \end{align*}
  Since complex multiplication is bilinear, the universal property of tensor products implies the existence of a linear map
  \begin{equation*}
    \psi \from \overline{V} \otimes \overline{W} \otimes V \otimes W \to \mathbb{C},\qquad
    \overline{v} \otimes \overline{w} \otimes v' \otimes w' \mapsto \braket{v}{v'}_{V}\braket{w}{w'}_{W}.
  \end{equation*}
  Applying Exercise~\ref{exe:translating-bilinear-to-sesquilinear} again, shows that
  \begin{equation*}
    \braket{\blank}{\bblank}_{V \otimes W} =  \psi \circ ( \blank \otimes \bblank) \from (V \otimes W) \times (V \otimes W) \to \mathbb{C}
  \end{equation*}
  is a well-defined sesquilinear map.

  It remains to show that \( \braket{\blank}{\bblank}_{V \otimes W} \) is Hermitian and positive definite.
  Since it suffices to verify these properties on elementary tensors, we fix elements  \(v, v' \in V\) and \(w,w' \in W\), and compute
  \begin{equation*}
  \braket{v \otimes w}{v' \otimes w'}_{V \otimes W} = \braket{v}{v'}_V \braket{w}{w'}_W = \overline{\braket{v'}{v}_V} \; \overline{\braket{w'}{w}_W} = \overline{\braket{v' \otimes w'}{v \otimes w}_{V \otimes W}}.
  \end{equation*}
  Similarly, we observe that
  \begin{equation*}
      \braket{v \otimes w}{v \otimes w}_{V \otimes W} = \braket{v}{v}_V \braket{w}{w}_W \geq 0,
  \end{equation*}
  with equality if and only if \( v = 0 \) or \( w = 0 \).
  Since \( 0 \otimes w = 0 = v \otimes 0 \), the inner product is positive definite.
\end{proof}

From now on, we endow \( V \otimes W \) with the tensor product inner product defined above.

\begin{corollary}\label{cor:onbs-tensor-to-onbs}
  Let \(\mathcal{B}\) be an orthonormal basis of \(V\) and \(\mathcal{C}\) an orthonormal basis of \(W\).
  Then \(\mathcal{B}\times \mathcal{C}\) is an orthonormal basis of \(V\otimes W\)
\end{corollary}


\begin{definition}\label{def:entangled-state}
  A pure state \( \ket{x} \in V \otimes W \) is called \emph{entangled} if there are no \(\ket{v} \in V\) and \(\ket{w} \in W\) such that \(\ket{x}= \ket{vw}\). % TODO Establish Notation!
\end{definition}

\begin{remark}\label{rmk:comparison-superposition-vs-entanglement}
  Let \(V\), \(W\) Hilbert spaces with bases \(\mathcal{B}\) and \(\mathcal{C}\).
  If  \(\ket{x}\in V\otimes W \) is an entangled state, it is in particular in a superposition with respect to the basis \(\mathcal{B}\times \mathcal{C}\) of \(V\otimes W\).
\end{remark}


\subsubsection{Tensor products of matrices}\label{sec:tensor-products-of-matrices}
Having established \( (\mathbb{C}^{2})^{\otimes n}\) as the model for the memory of a quantum computer,
we now want to study its gates.
That is, linear maps and matrices between tensor products.

\begin{definition}\label{def:kronecker-product}
  The \emph{Kronecker product} of the matrices \( A \in \Mat_{m,n}(\mathbb{C}) \) and \( B \in \Mat_{p,q}(\mathbb{C}) \) is the matrix \(A\otimes B \in \Mat_{mp,nq} ( \mathbb{C})\) given by
  \((A\otimes B)_{ip+k,jq+l}= a_{ij}b_{kl}\) where \((i,k,j,l)\in[m]\times[p]\times[n]\times[q]\).
\end{definition}
The Kronecker product \(A\otimes B\) of \(A\) and \(B\)  can be subdivided into blocks of size \(p\times q\) with entries of the  form \(a_{ij}B \in \Mat_{p,q}( \mathbb{C})\):
\begin{equation*}
  A \otimes B = \begin{pmatrix}
    a_{11}B & a_{12}B & \cdots & a_{1n}B \\
    a_{21}B & a_{22}B & \cdots & a_{2n}B \\
    \vdots  & \vdots  & \ddots & \vdots  \\
    a_{m1}B & a_{m2}B & \cdots & a_{mn}B
  \end{pmatrix} \in \Mat_{mp,nq}(\mathbb{C}).
\end{equation*}

Matrix-vector multiplication involving the Kronecker product can be cumbersome due to its block structure.
However, by a slight abuse of notations, certain cases can be handled quite efficiently.
\begin{remark}\label{rmk:matrix-vector-multiplication-Kronecker-product}
  Let \(A\in \Mat_{m,n}( \mathbb{C})\) and \(B \in \Mat_{p,q}( \mathbb{C})\).
  Given a vector \(v\in \mathbb{C}^{n}\) and \(w \in \mathbb{C}^{q}\), we write by a slight abuse of notation
  \begin{equation*}
    v\otimes w =
    \begin{pmatrix}
      v_{1} w\\
      v_{2} w\\
      \vdots
      v_{n} w
    \end{pmatrix} \in \mathbb{C}^{nq},
  \end{equation*}
  where each \( v_i w \) represents the \(q\) entries \(v_{i}w_{1}, \dots , v_{i}w_{q}\).

  With this notation, the action of the Kronecker product \(A\otimes B\) on the tensor product \(v\otimes w\) is given by
  \begin{equation*}
    (A\otimes B) (v\otimes w) = Av \otimes Bw.
  \end{equation*}
\end{remark}

In Remark~\ref{rmk:tensor-product-string-diags}, we discussed how to represent tensor products of vector space diagrammatically.
We can now extend these string diagrams to include matrices.

\begin{remark}\label{rmk:matrices-as-string-diagrams}
  Consider matrices \(A\in \Mat_{m,n}( \mathbb{C})\) and \(B \in \Mat_{p,q}( \mathbb{C})\).
  We visualise the Kronecker product \(A\otimes B \in \Mat_{mp,nq}( \mathbb{C})\) as shown below.
  \begin{equation*}
    \inputtikz{tensor-product-of-matrices}
  \end{equation*}
  As before, we will omit drawing identity matrices, leading to pictures where only one strand has a box on it.
  \begin{equation*}
    \inputtikz{tensor-product-of-matrices-w-identities}
  \end{equation*}
  Conversely, not every matrix \(C \in \Mat_{mp,nq}(\mathbb{C})\) is a product \(C = A\otimes B\).
  In this case, we represent \(C\) by a box spanning multiple strands.
  \begin{equation*}
    \inputtikz{indecomposable-matrix}
  \end{equation*}
\end{remark}

A natural question at this stage is whether we need to be careful with respect to the horizontal alignment of the  boxes, or if they can be moved freely on their strings.

\begin{lemma}\label{lem:eck-hilton}
  Consider matrices \(A \in Mat_{m,n}( \mathbb{C})\), \(C \in \Mat_{n,p}( \mathbb{C})\), \(B \in \Mat_{r,s}( \mathbb{C})\), and \(D \in \Mat_{t,u}( \mathbb{C})\).
  The Kronecker product and matrix  multiplication satisfy the compatibility relation
  \begin{equation}\label{eq:Kronecker-product-vs-matrix-mult}
    (A\otimes B)(C \otimes D) = AC \otimes BD \in \Mat_{mr,pu}( \mathbb{C}).
  \end{equation}
\end{lemma}
\begin{proof}
  The claim follows from Lemmas~\ref{lem:tensor-product-kronecker-product} and~\ref{lem:comp-vs-tensoring}. %TODO these might become propositions...
\end{proof}

String diagrams could either be read vertically first or horizontally first.
The previous result shows that both interpretations give the same result, so we can neglect the choice of direction.
\begin{equation*}
  \inputtikz{box-order}
\end{equation*}

Moreover, as a direct application, we obtain that we can move boxes freely along their strands.

\begin{equation*}
  \inputtikz{exchange-order-matrices}
\end{equation*}

Let us now discuss some some fundamental properties of the Kronecker product.

\begin{lemma}\label{lem:props-of-tensor-products}
  Let \( A \in \Mat_{m,n}(\mathbb{C}) \) and \( B \in \Mat_{p,q}(\mathbb{C}) \).
  The Kronecker product satisfies
  \begin{thmlist}
    \item \(\overline{A \otimes B} = \overline{A} \otimes \overline{B}\),
    \item \((A \otimes B)^T = A^T \otimes B^T\),
    \item \((A \otimes B)^\dagger = A^\dagger \otimes B^\dagger\),
    \item \(I_n \otimes I_q = I_{nq}\).
  \end{thmlist}
\end{lemma}
\begin{proof}
  \begin{thmlist}
    \item The first property follows directly from the definition of complex conjugation on matrices.
    \item For the transpose, we verify the statement by an index-wise computation:
    \begin{equation*}
      (A \otimes B)^T_{jq + l, ip + k} = (A \otimes B)_{ip + k, jq + l} = a_{ij}b_{kl} = A^T_{ji}B^T_{lk} = (A^T \otimes B^T)_{jq + l, ip + k}.
    \end{equation*}
    \item The third property follows from the first two properties, since \( (A \otimes B)^\dagger = \overline{(A \otimes B)^T} = \overline{A^T \otimes B^T} = \overline{A^T} \otimes \overline{B^T} = A^\dagger \otimes B^\dagger \).
    \item The fourth property is verified by a direct computation.
  \end{thmlist}
\end{proof}

In the context of quantum computing we are mainly interested in unitary matrices, orthogonal projections, and, more abstractly, partial isometries.
\begin{lemma}\label{lem:matrix-properties}
  Let \( A \in \Mat_n(\mathbb{C}) \), \( B \in \Mat_q(\mathbb{C}) \) and
  \(C \in \Mat_{m,n}( \mathbb{C})\), \( D \in \Mat_{p,q}( \mathbb{C})\).
\begin{thmlist}
\item If \( A \) and \( B \) are self-adjoint, then \( A \otimes B \) is self-adjoint.
\item If \( A \) and \( B \) are positive semidefinite, then \( A \otimes B \) is positive semidefinite.
\item If \( A \) and \( B \) are orthogonal projections, then \( A \otimes B \) is an orthogonal projection.
\item If \( C \) and \( D \) are partial isometries, then \( C \otimes D \) is a partial isometry.
\item If \( A \) and \( B \) are unitary, then \( A \otimes B \) is unitary.
\end{thmlist}
\end{lemma}
\begin{proof}
  We verify each item by a direct computation.
\begin{thmlist}
  \item If \( A \) and \( B \) are self-adjoint, then \( (A \otimes B)^\dagger = A^\dagger \otimes B^\dagger = A \otimes B \), so \( A \otimes B \) is self-adjoint.

  \item If \( A \) and \( B \) are positive semidefinite, there exist orthonormal bases \( (v_1, \dots, v_n) \) of \( \mathbb{C}^n \) and \( (w_1, \dots, w_q) \) of \( \mathbb{C}^q \) consisting of eigenvectors of \( A \) and \( B \), respectively, with non-negative eigenvalues \( (\lambda_1, \dots, \lambda_n) \) and \( (\mu_1, \dots, \mu_q) \).
  Then
  \begin{equation*}
    (A \otimes B)(v_i \otimes w_j) = Av_i \otimes Bw_j = \lambda_i \mu_j (v_i \otimes w_j),
  \end{equation*}
  and \( (v_i \otimes w_j)_{i,j} \) forms a basis of \( \mathbb{C}^{nq} \) consisting of eigenvectors of \( A \otimes B \) with non-negative eigenvalues \( \lambda_i \mu_j \).
  Thus, \( A \otimes B \) is positive semidefinite.
  \item If \( A \) and \( B \) are orthogonal projections, then \( A^2 = A \) and \( B^2 = B \).
  We now have
  \begin{equation*}
    (A \otimes B)^2 = A^2 \otimes B^2 = A \otimes B,
  \end{equation*}
  and \( A \otimes B \) is self-adjoint by the first property.
  Therefore, \( A \otimes B \) is an orthogonal projection.
  \item \( C \) and \( D \) being partial isometries, corresponds to  \( C^\dagger C = P_C \) and \( D^\dagger D = P_D \) being orthogonal projections.
  We compute
  \begin{equation*}
    (A \otimes B)^\dagger (A \otimes B) = A^\dagger A \otimes B^\dagger B = P_A \otimes P_B,
  \end{equation*}
  which is an orthogonal projection by the third property and \(A\otimes B\) is a partial isometry.
  \item We compute
  \begin{equation*}
    (A \otimes B)^\dagger (A \otimes B)
    = A^\dagger A \otimes B^\dagger B
    = I_n \otimes I_q
    = I_{nq}
    =  I_n \otimes I_q
    = A A^\dagger \otimes B B^\dagger
    = (A \otimes B)(A \otimes B)^\dagger.
  \end{equation*}
  Thus, \( A \otimes B \) is unitary. \qedhere
\end{thmlist}
\end{proof}

\begin{corollary}\label{cor:tensor-product-of-density-matrices}
The tensor product of two density matrices \( A \in \Mat_{n}(\mathbb{C}) \) and \( B \in \Mat_{m}(\mathbb{C}) \) is a density matrix.
\end{corollary}
\begin{proof}
By Lemma~\ref{lem:matrix-properties}, \( A \otimes B \) is self-adjoint and positive semidefinite. It remains to show that \( \tr(A \otimes B) = 1 \). Using the block matrix form of the Kronecker product, we have:
\begin{equation*}
\tr(A \otimes B) = \sum_{i=1}^m \sum_{j=1}^n a_{ii}b_{jj} = \left( \sum_{i=1}^m a_{ii} \right) \left( \sum_{j=1}^n b_{jj} \right) = \tr(A) \tr(B) = 1 \cdot 1 = 1. \qedhere
\end{equation*}
\end{proof}

\subsubsection{Tensor products of linear maps}\label{sec:tensor-products-of-linear-maps}

Since working with iterated Kronecker products can be cumbersome, we want  to hide the complexity behind an layer of abstraction: tensor products of linear maps.

\begin{lemma}\label{lem:existence}
  Let \( f \from V \to W \) and \( g \from X \to Y \) be linear maps.
  There exists a unique linear map \( f \otimes g \from V \otimes X \to W \otimes Y \) such that
  \begin{equation*}
    (f \otimes g)(v \otimes x) = f(v) \otimes g(x)
  \end{equation*}
  for all \( v \in V \) and \( x \in X \).
\end{lemma}
\begin{proof}
  The map \(V\times X \to W\otimes Y\), \((v, w) \mapsto f(v) \otimes g(w)\) is bilinear implying that there is a unique linear map
  \begin{equation*}
    f\otimes g \from V\otimes W \to X\otimes Y, \qquad v\otimes w \mapsto f(v) \otimes g(w). \qedhere
  \end{equation*}
\end{proof}

A natural question is how to translate between the tensor product of linear maps and the Kronecker product of matrices.

\begin{lemma}\label{lem:tensor-product-kronecker-product}
  Consider two linear maps \(f\from V\to W\) and \(g\from X\to Y\) and suppose that \(V\) and \(X\) have the bases \(\mathcal{B}_{V} = (v_{1}, \dots , v_{n})\) and \(\mathcal{B}_{X} = (x_{1}, \dots , x_{q})\) and \(W\) as well as \(Y\) have the bases \(\mathcal{C}_{W}= (w_{1}, \dots , w_{m})\) and \(\mathcal{C}_{Y}=(y_{1}, \dots , y_{p})\).

  We endow \(V\otimes X\) and \(W\otimes Y\) with the bases
  \begin{gather*}
    \mathcal{B}_{V\otimes X} = (v_{1}\otimes x_{1}, \dots , v_{1}\otimes x_{q}, v_{2} \otimes x_{1}, \dots , v_{2}\otimes x_{q}, \dots ,v_{n}\otimes x_{1}, \dots , v_{n} \otimes x_{q})\\
    \mathcal{C}_{W\otimes Y} = (w_{1}\otimes y_{1}, \dots , w_{1}\otimes y_{p}, w_{2} \otimes y_{1}, \dots , w_{2}\otimes y_{p}, \dots ,w_{m}\otimes y_{1}, \dots , w_{m} \otimes y_{p})
  \end{gather*}
  and write \(F= {}_{\mathcal{C}_{W}}f_{\mathcal{B}_{V}}\) and \(G = {}_{\mathcal{C}_{Y}}g_{\mathcal{B}_{X}}\).
  Then, the following diagram commutes
  \begin{equation*}
    \begin{tikzcd}[ampersand replacement=\&,cramped]
      {V\otimes X} \&\& {W\otimes Y} \\
      \\
      {\mathbb{C}^{nq}} \&\& {\mathbb{C}^{mp}}
      \arrow["{f\otimes g}", from=1-1, to=1-3]
      \arrow["{\can_{\mathcal{B}_{V\otimes X}}}", from=3-1, to=1-1]
      \arrow["{\mu_{F\otimes G}}"', from=3-1, to=3-3]
      \arrow["{\can_{\mathcal{B}_{W\otimes Y}}}"', from=3-3, to=1-3]
    \end{tikzcd}
  \end{equation*}
\end{lemma}
\begin{proof}
  Given \(1\leq i \leq n\) and \(1\leq j \leq q\), we compute
  \begin{align*}
    \!\!\!\!\!(f\otimes g)&\circ \can_{\mathcal{B}_{V\otimes W}} (e_{i}\otimes e_{j})
    = (f\otimes g) v_{i}\otimes x_{j}
    = f(v_{i}) \otimes g(x_{j})\\
    & = \can_{\mathcal{C}_{W}}(Fe_{i}) \otimes \can_{\mathcal{C}_{Y}}(Ge_{j})
    = \can_{\mathcal{C}_{W\otimes Y}}( Fe_{i} \otimes Ge_{j})
    = \can_{\mathcal{C}_{W\otimes Y}}\circ \mu_{F\otimes G}(e_{i} \otimes e_{j}) \qedhere
  \end{align*}
\end{proof}

As with matrices, we want to establish compatibility between the orders of composition and tensoring.

\begin{lemma}\label{lem:comp-vs-tensoring}
  Let \(f\from V\to W\), \(g\from W\to X\) and \(h\from A \to B\), \(k \from B\to C\) be linear maps.
  We have
  \begin{equation*}
    (g\circ f)\otimes (k \circ h) = (g \otimes k)\circ (f\otimes h).
  \end{equation*}
\end{lemma}
\begin{proof}
  Given vectors \(v\in V\) and \(w\in W\), we compute
  \begin{align*}
    \big((g\circ f)&\otimes (k \circ h)\big) v\otimes w
    = \big((g\circ f) v\big) \otimes \big((k \circ h) w\big)
    = g(f(v)) \otimes k(h(w))\\
    &= (g \otimes k)\big( f(v) \otimes h(w)\big)
    = \big((g\otimes k) \circ (f\otimes h)\big) (v\otimes w). \qedhere
  \end{align*}
\end{proof}

As with matrices, we can represent tensor products of linear maps using string diagrams.

\begin{remark}\label{rmk:string-diags-tensoring-linear-maps}
  The diagram,representing the tensor product of linear  maps \(f\from V\to W\) and \(g\from X\to Y\)
  is shown below.
  \begin{equation*}
    \inputtikz{tensor-product-of-lin-maps}
  \end{equation*}
  Note that by the previous result, both interpretations of such a diagram---reading it first horizontally (\ie giving precedence to composition) or vertically (first resolving tensors)--yield the same result.
  Moreover, we can freely move boxes along their strings and do not have to pay close attention to their horizontal alignment.

  If we have a map \(f\from V_{1} \otimes \dots \otimes V_{n} \to W_{1}\otimes \dots \otimes W_{m}\), the corresponding string diagram has a spider-like shape.
  \begin{equation*}
    \inputtikz{tens-product-mult-ins-and-outs}
  \end{equation*}
\end{remark}

So far we have two restrictions left for our string diagrams.
First, we cannot bend edges, second, we cannot cross edges.
To address the first issue, we need to look at the dual of a space \(V\).


\begin{definition}\label{def:dual-basis}
  Let \(\mathcal{B} = (b_{1}, \dots , b_{n})\) be a basis of a vector space \(V\).
  The \emph{dual basis} \(\rd*{B}= (\beta_{1}, \dots , \beta_{n})\) is given by the linear maps
  \begin{equation*}
    \beta_{i} \from V \to \mathbb{C}, \qquad b_{j}\mapsto
    \begin{cases}
      1 & i=j \\
      0 & \text{otherwise.}
    \end{cases}
  \end{equation*}
\end{definition}

\begin{lemma}\label{lem:dual-basis-is-basis}
  Let \(\mathcal{B}= (b_{1}, \dots , b_{n})\) be a basis of \(V\).
  The dual basis \(\rd*{\mathcal{B}}= (\beta_{1}, \dots , \beta_{n})\) is a basis of \(\rd*{V}= \Hom_{ \mathbb{C}}(V, \mathbb{C})\).
\end{lemma}
\begin{proof}
  Consider any linear map \(f \from V\to \mathbb{C}\) and set \(\lambda_{1} = f(b_{i})\) for all \(1\leq i\leq n\).
  A direct verification on the basis \(\mathcal{B}\) of \(V\) shows that \(f=\lambda_{1} \beta_{1} + \dots + \lambda_{n}\beta_{n}\).
  Clearly, this linear combination is unique and the claim follows.
\end{proof}

\begin{proposition}\label{prop:can-iso}
  For any finite-dimensional vector space \(V\) there is an isomorphism
  \begin{align*}
    \mathrm{dk} \from \rd*{V} \otimes V &\to \End_{ \mathbb{C}}(V)\\
    \gamma \otimes v &\mapsto (w \mapsto \gamma(w)v).
  \end{align*}
\end{proposition}
\begin{proof}
  We choose basis \(\mathcal{B}=(b_{1}, \dots , b_{n})\) of \(V\) with dual basis \(\rd*{\mathcal{B}}= (\beta_{1}, \dots , \beta_{n})\).
  Moreover, we consider the standard basis \((E_{ij})_{1\leq i,j \leq n}\) of \(\Mat_{n}( \mathbb{C})\),
  where
  \begin{equation*}
    (E_{ij})_{kl}=
  \begin{cases}
    1 & k=i, j=l \\
    0 & \text{otherwise.}
  \end{cases}
\end{equation*}
  We have linear isomorphisms
  %Establish notation \(E_{ij}\).
  \begin{align*}
    k\from \rd*{V} \otimes V &\to \Mat_{n}( \mathbb{C}),\qquad  \beta_{i}\otimes b_{j}\mapsto E_{ij} \\ % or should it be E_{ji} please check.
    d \from \Mat_{n}( \mathbb{C}) &\to \End_{ \mathbb{C}}(V), \qquad A \mapsto \mu_{A}.
  \end{align*}
  Thus the composition \(\mathrm{dk} = d\circ k\) is an isomorphism as well.
\end{proof}

The space \(\End_{ \mathbb{C}}(V)\) has a distinguished element, namely \(\id_{V}\).
By chosing a basis \(\mathcal{B}\) of \(V\), this corresponds under the isomorphism \(\mathrm{dk}^{-1} \from \End_{ \mathbb{C}}(V) \to \rd*{V} \otimes V\) to \(\sum_{i=1}^{n} \beta_{i}\otimes b_{i}\).

Moreover, we can associate to any endomorphism \(f\in \End_{ \mathbb{C}}(V) \) its trace \(\tr(f) \in \mathbb{C}\).
By applying \(\mathrm{dk}^{-1}\from \End_{ \mathbb{C}}(V) \to \rd*{V} \to V\), it is not difficult to see that this corresponds to the assignment \(\gamma \otimes v \mapsto \gamma(v)\).

\begin{definition}\label{def:cap-cup}
  Let \(V\) be a finite-dimensional vector space, \(\mathcal{B} =( b_{1}, \dots , b_{n})\) a basis of \(V\) and \(\rd*{\mathcal{B}} = (\beta_{1}, \dots , \beta_{n})\) its dual basis.
  The linear maps
  \begin{subequations}
    \begin{align}
      \coev &\from \mathbb{C} \to \rd*{V}\otimes V,& \lambda &\mapsto \lambda \sum\nolimits_{i=1}^{n}\beta_{i}\otimes b_{i},\\
      \overline{\coev} &\from \mathbb{C} \to  V\otimes \rd*{V}, &\lambda &\mapsto \lambda \sum\nolimits_{i=1}^{n}b_{i} \otimes \beta_{i}
    \end{align}
  \end{subequations}
  are called right and left \emph{coevaluation}.
  Similarly, the right and left \emph{evaluations} are
  \begin{subequations}
    \begin{align}
      \ev &\from \rd*{V}\otimes V \to \mathbb{C}, &\gamma\otimes v &\mapsto \gamma(v) \\
      \overline{\ev} &\from V\otimes \rd*{V} \to \mathbb{C}, &v \otimes \gamma& \mapsto \gamma(v)
    \end{align}
  \end{subequations}
\end{definition}


We visualise evaluations and coevaluations by caps and cups
\begin{equation*}
  \inputtikz{caps-and-cups}
\end{equation*}
Note that passing from a vector space \(V\) to its dual space \(\rd*{V}\) reverses the direction of the arrows in our string diagrams
\begin{equation*}
  \inputtikz{dual-space-as-string-diagram}
\end{equation*}

\begin{lemma}\label{lem:cap-cup}
  The left and right (co)-evaluations of a finite-dimensional vector space \(V\) satisfy the snake identities
\begin{align*}
  (\id_{V} \otimes \ev_{V})(\coev_{V} \otimes \id_{V}) = \id_{V}\qquad  \text{ and}\qquad
  (\ev_{V}\otimes \id_{\rd*{V}})(\id_{\rd*{V}}\otimes \coev_{V}) = \id_{\rd*{V}}, \\
    (\overline{\ev}_{V} \otimes \id_{V})(\id_{V} \otimes \overline{\coev}_{V}) = \id_{V}\qquad  \text{ and}\qquad
  (\id_{\rd*{V}} \otimes \overline{\ev_{V}})(\overline{\coev_{V}} \otimes \id_{\rd*{V}}) = \id_{\rd*{V}}. \\
\end{align*}
\end{lemma}
In string-diagrammatic form, the previous result can be expressed as follows.
\begin{equation*}
  \inputtikz{snake-identity} % include once picture is typeset
\end{equation*}

Observation: we can now ``close'' string diagrams.
These represent a linear map \( \gamma \from \mathbb{C} \to \mathbb{C}\) and any such map corresponds to a number \(\gamma(1) \in \mathbb{C}\). %<- here we only need that that linear maps c\to c and complex numbers can be identified with each other.

\begin{proposition}\label{prop:traces}
  For any linear map \(f\ \from V\to V\), we have
  \begin{equation*}
    \tr(f) = \ev_{V} (f\otimes \id_{\rd*{V}}) \overline{\coev}_{V} = \overline{\ev_{V}} (\id_{\rd*{V}}\otimes f) \coev_{V}.
  \end{equation*}
\end{proposition}
\begin{proof}
  The claim follows from a direct computation.
  \begin{equation*}
   \ev_{V} (f\otimes \id_{\rd*{V}}) \overline{\coev}_{V} =  \sum_{i=1}^{n} \beta_{i}(f(b_{i})) = \tr(f). \qedhere
  \end{equation*}
\end{proof}


% TODO Partial traces

Given two vector spaces \(V,W\) we have a bilinear map
\begin{equation*}
  V\times W \trightarrow{(v,w)\mapsto (w,v)} W\times V \trightarrow{(w,v)\mapsto w\otimes v} W\otimes V.
\end{equation*}

\begin{definition}\label{def:tensor-flip}
  Let \(V\), \(W\) be two vector spaces.
  The \emph{tensor flip} or \emph{swap map} is the linear map
  \begin{equation}
    \sigma_{V,W}\from V\otimes W \to W\otimes V, \qquad v\otimes w \mapsto w\otimes v.
  \end{equation}
\end{definition}


We draw the tensor flip as a crossing.
\begin{equation*}
  \inputtikz{tensor-flip}
\end{equation*}


\begin{lemma}\label{lem:tensor-flip}
  Consider linear maps \(f\from V\to W\) and \(g\ \from X\to Y\).
  The tensor flip satisfies
  \begin{equation}
    \sigma_{W,Y}(f\otimes g)= (g\otimes f) \sigma_{V,X}.
  \end{equation}
\end{lemma}
\begin{proof}
  Given \(v\in V\) and \(x\in X\), we compute
  \begin{align*}
    (\sigma_{W,Y}(f\otimes g))(v\otimes x)
    & = \sigma_{W, Y} (f(v) \otimes g(x))
      = g(x) \otimes f(v) \\
    & = (g \otimes f)(x\otimes v)
      = ((g\otimes f) \sigma_{W,X})(v\otimes x).
  \end{align*}
\end{proof}

\subsection{Unitary transformations}\label{sec:unitary-transformations}
Aside from measurements, unitary transformations are a central component of quantum algorithms.

\begin{lemma}\label{lem:induced-map-on-states}
  Any unitary map \( f \from V \to V \) induces a well-defined map on states:
  \begin{equation}
    \{\text{states of } V\} \to \{\text{states of } V\}, \qquad [\ket{v}] \mapsto [f\ket{v}].
  \end{equation}
  Moreover, for any pair of states \( [\ket{v}] \) and \( [\ket{w}] \), there exists a unitary map \( f \) such that \( [f\ket{v}] = [\ket{w}] \).
\end{lemma}
\begin{proof}
  Since a state is an equivalence class of normalised vectors, we need to verify that \( f \) preserves norms and respects the equivalence relation.
  For any \( \ket{v} \in V \)
  \begin{equation*}
    \|f\ket{v}\|^2
    = \braket{f\ket{v}}{f\ket{v}}
    = \braket{\ket{v}}{f^\dagger f\ket{v}}
    = \braket{\ket{v}}{\ket{v}}
    = \|\ket{v}\|^2.
  \end{equation*}
  Moreover, if \( \ket{v} = \lambda \ket{w} \), then \( f(\ket{v}) = \lambda f(\ket{w}) \), so the induced map is well-defined.

  To show the existence of a unitary map between any two states, consider two normalised vectors \( \ket{v} \) and \( \ket{w} \).
  We fix  orthonormal bases \( (\ket{b_1}, \ket{b_2}, \dots, \ket{b_n}) \) and \((\ket{c_1}, \ket{c_2}, \dots, \ket{c_n})\) of \(V\) where \(\ket{b_{1}} = \ket v\) and \(\ket{c_{1}} = \ket w\).
  Next, we define the linear map \( f \colon V \to V \) by
  \begin{equation*}
        f \ket{b_i} = \ket{c_i} \qquad  \text{for all } i = 1, \dots, n.
  \end{equation*}
  This map \( f \) is unitary since it maps an orthonormal basis to an orthonormal basis, and it satisfies \( f\ket{v} = \ket{w} \).
\end{proof}

We now want to translate the action of unitaries on states to an operation on mixed states.

\begin{remark}\label{rmk:unitaries-on-mixed-states}
  Consider a normalised vector \(\ket{v} \in V \) and a unitary map \( f \from V \to V \).
  We write \(\bra{f v} \eqdef d(f\ket{v})\) for the image of \(f\ket{v}\) under the Riesz-map \(d \from V \to \rd*{V}\).
  The mixed state \(\rho_{f \ket{v}}\) acts on \(\ket{w} \in V\) by
  \begin{equation*}
    \rho_{f \ket{v}}\ket{w}
    = f \ket{v} \braket{ f v}{w}
    = f \ket{v} \braket{v}{f^{\dagger}\mid w}
    = (f \rho_{\ket v} f^{\dagger}) w.
  \end{equation*}
\end{remark}

\begin{lemma}\label{lem:mixed-state-action}
  Let \(m, f\in \End(V)\) be a mixed state and a unitary map, respectively.
  Then \(fmf^{\dagger}\) is again a mixed state
\end{lemma}
\begin{proof}
  Since \((fmf^{\dagger})^{\dagger} = {f^{\dagger}}^{\dagger}m^{\dagger} f^{\dagger} = f m f^{\dagger}\), we have that \(fmf^{\dagger}\) is self-adjoint.
  To show that \(fmf^{\dagger}\) is positive semi-definite, we observe that
  \begin{equation*}
    \braket{v}{fmf^{\dagger} \mid v} = \braket{f^{\dagger } v}{m \mid f v} \geq 0
  \end{equation*}
  for all \(\ket{v}\in V\).
  Finally, we have \(\tr(fmf^{\dagger}) = \tr(f^{\dagger} f m ) = \tr(m) = 1\).
\end{proof}

\subsubsection{Quantum Gates on the Bloch Sphere}\label{sec:quantum-gates-on-the-bloch-sphere}
To develop intuition for how unitaries affect quantum states, we focus on a single qubit.

\begin{definition}\label{def:rotations-on-bloch-sphere}
  Consider real numbers \(\alpha, \beta,\gamma \in \mathbb{R}\), we call the linear maps
  \begin{subequations}
    \begin{align}
      Z(\alpha) &= \ket{0}\bra{0} + e^{2\pi i \alpha}\ket{1}\bra{1} \from \mathbb{C}^{2} \to \mathbb{C}^{2}, \\
      X(\beta) &= \ket{+}\bra{+} + e^{2\pi i \beta}\ket{-}\bra{-} \from \mathbb{C}^{2} \to \mathbb{C}^{2}, \\
      Y(\gamma) &= \ket{+i}\bra{+i} + e^{2\pi i \gamma}\ket{-i}\bra{-i} \from \mathbb{C}^{2}\to \mathbb{C}^{2}.
    \end{align}
  \end{subequations}
  the \(Z\), \(X\), and \(Y\)-\emph{phase gate}, respectively.
\end{definition}

\begin{lemma}\label{lem:xyz-gates-are-unitary}
  The gates \(Z(\alpha), X(\beta), Y(\gamma) \from \mathbb{C}^{2} \to \mathbb{C}^{2}\) are unitary for all \(\alpha,\beta, \gamma \in \mathbb{R}\).
\end{lemma}
\begin{proof}
  We explicitly verify unitarity for the \( Z(\alpha) \) gate.
  The adjoint of \( Z(\alpha) \) is
  \begin{equation*}
    Z(\alpha)^\dagger = \ket{0}\bra{0} + e^{-2\pi i \alpha}\ket{1}\bra{1}
  \end{equation*}
  and we compute
  \begin{align*}
    Z(\alpha) Z(\alpha)^\dagger
    & = (\ket{0}\bra{0} + e^{2\pi i \alpha}\ket{1}\bra{1})(\ket{0}\bra{0} + e^{-2\pi i \alpha}\ket{1}\bra{1})\\
    & = \ket{0} \braket{0}{0} \bra{0}
    + e^{-2\pi i \alpha}\ket{0}\braket{0}{1}\bra{1}
    + e^{2\pi i \alpha}\ket{1}\braket{1}{0}\bra{0}
      + \ket{1}\braket{1}{1}\bra{1}\\
    & = \ket{0}\bra{0} + \ket{1}\bra{1} = I_2,
  \end{align*}
  and similarly \( Z(\alpha)^\dagger Z(\alpha) = I_2 \).
  Thus, \( Z(\alpha) \) is unitary.

  Ana analogous computation for the \(X(\beta)\) and \(Y(\gamma)\) gates concludes the proof.
\end{proof}

To visualise the action of the \( Z \), \( X \), and \( Y \) gates on the Bloch sphere,
we introduce the following rotations around the \( z \), \( x \), and \( y \)-axes
\begin{subequations}
  \begin{align}
    z(\alpha) \colon \mathbb{R}^3 &\to \mathbb{R}^3, \quad (x, y, z) \mapsto (x \cos(2\pi \alpha) + y \sin(2\pi \alpha), - x \sin(2\pi \alpha) + y \cos(2\pi \alpha), z), \\
    x(\beta) \colon \mathbb{R}^3 &\to \mathbb{R}^3, \quad (x, y, z) \mapsto TODO\\
    y(\gamma) \colon \mathbb{R}^3 &\to \mathbb{R}^3, \quad (x, y, z) \mapsto TODO
  \end{align}
\end{subequations}

\begin{lemma}\label{lem:xyz-gates-are-rotations}
  Given a state \(\ket{v} \in \mathbb{C}^{2}\), we write \(r_{\ket v} \in \mathbb{R}^{3}\) for the corresponding point in the Bloch sphere.
  For any \( \alpha, \beta, \gamma \in \mathbb{R} \), we have
  \begin{align*}
    r_{Z(\alpha) \ket{v}} = z(\alpha)r_{\ket{v}}, \qquad
    r_{X(\beta) \ket{v}} = x(\beta)r_{\ket{v}}, \qquad
    r_{Y(\gamma) \ket{v}} = y(\gamma)r_{\ket{v}}.
  \end{align*}
\end{lemma}

\begin{proof}
  We prove the claim for the \( Z(\alpha) \) gate.
  The matrix representation of \( Z(\alpha) \) with respect to the standard basis is
  \begin{equation*}
    Z(\alpha) = \begin{pmatrix}
      1 & 0 \\
      0 & e^{2\pi i \alpha}
    \end{pmatrix}.
  \end{equation*}
  Let \( \ket{v} \in \mathbb{C}^2 \) be a state, and let \( \rho_{\ket{v}} = \ket{v}\bra{v} \) be its associated density matrix.
  By Lemma~\ref{lem:characterising-the-qubit}, we have
  \begin{equation*}
    \rho_{\ket{v}} = \frac{1}{2}(I_2 + r_x \sigma_x + r_y \sigma_y + r_z \sigma_z),
  \end{equation*}
  where \( r_{\ket{v}} = (r_x, r_y, r_z) \).

  The density matrix of the transformed state \( Z(\alpha)\ket{v} \) is given by \(\rho_{Z(\alpha)\ket{v}} = Z(\alpha) \rho_{\ket{v}} Z(\alpha)^\dagger\).

  We now compute the conjugation of each Pauli matrix by \( Z(\alpha) \).
  \begin{align*}
    Z(\alpha) I_2 Z(\alpha)^\dagger
    &= I_2, \\[0.25\baselineskip]
    %
    Z(\alpha) \sigma_z Z(\alpha)^\dagger
    &=
      \begin{pmatrix}
        1 & 0 \\
        0 & e^{2\pi i \alpha}
      \end{pmatrix}
      \begin{pmatrix}
        1 & 0 \\
        0 & -1
      \end{pmatrix}
      \begin{pmatrix}
        1 & 0 \\
        0 & e^{- 2\pi i \alpha}
      \end{pmatrix}
      =
      \begin{pmatrix}
        1 & 0 \\
        0 & -1
      \end{pmatrix}
      =\sigma_{z}
    \\[0.25\baselineskip]
    %
    Z(\alpha) \sigma_x Z(\alpha)^\dagger
    & =
      \begin{pmatrix}
        1 & 0 \\
        0 & e^{2\pi i \alpha}
      \end{pmatrix}
      \begin{pmatrix}
        0 & 1 \\
        1 & 0
      \end{pmatrix}
      \begin{pmatrix}
        1 & 0 \\
        0 & e^{-2\pi i \alpha}
      \end{pmatrix}\\
    & =
      \begin{pmatrix}
        0 & e^{-2\pi i \alpha} \\
        e^{2\pi i \alpha} & 0
      \end{pmatrix}
      =  \cos(2\pi \alpha) \sigma_x + \sin(2\pi \alpha) \sigma_y, \\[0.25\baselineskip]
    Z(\alpha) \sigma_y Z(\alpha)^\dagger
    & =
      \begin{pmatrix}
        1 & 0 \\
        0 & e^{2\pi i \alpha}
      \end{pmatrix}
      \begin{pmatrix}
        0 & -i \\
        i & 0
      \end{pmatrix}
      \begin{pmatrix}
        1 & 0 \\
        0 & e^{-2\pi i \alpha}
      \end{pmatrix}\\
    & =
      \begin{pmatrix}
        0 & -ie^{-2\pi i \alpha} \\
        ie^{2\pi i \alpha} & 0
      \end{pmatrix}
      = -\sin(2\pi \alpha) \sigma_x + \cos(2\pi \alpha) \sigma_y.
  \end{align*}
  Thus, we get \(r_{Z(\alpha) \ket{v}} = z(\alpha)r_{\ket{v}}\).
\end{proof}

We have already seen that any state can be transformed into any other state via a unitary map.
However, this unitary is far from unique.
In the single-qubit case, the \( Z \) and \( X \) phase gates suffice to generate all single-qubit unitaries.

\begin{lemma}\label{lem:euler-decomposition-of-rotation}
  For any unitary map \( f \from \mathbb{C}^2 \to \mathbb{C}^2 \), there exist \( \alpha, \beta, \gamma \in [0,1) \) such that
  \begin{equation}
    e^{2\pi i \lambda} f = Z(\alpha) X(\beta) Z(\gamma) ,\qquad\quad \text{ for some } \lambda \in \mathbb{R}.
  \end{equation}
\end{lemma}
\begin{proof}
  TODO: Easy proof idea:
  Mercator projection:
  Map from \([0,1]^{2} \to S^{2}\) such that rotation by \(z(\alpha)\) horizontal movement
  rotation by \(x(\beta)\) vertical movement.
  Clearly any point in \([0,1]^{2}\) can be reached by...
\end{proof}

\subsubsection{Translating Classical Algorithms to Quantum Algorithms}\label{sec:translating-classical-algorithms-to-quantum-algorithms}
To understand how quantum algorithms generalise classical ones, we outline a non-trivial process for transforming classical computations into quantum computations.
At this stage, and without going into further details, we adopt a high-level perspective on classical algorithms, and simply model them as maps \(f \from \{0,1\}^n \to \{0,1\}^m\).

\begin{definition}\label{def:addition-modulo-2}
  The \emph{addition modulo 2} is the operation
  \begin{equation}
    \boxplus \from\{0,1\}^{2} \to\{0,1\}, \qquad (a, b) \mapsto
    \begin{cases}
      1 & a=1, b=0 \text{ or } a= 0, b=1 \\
      0 & a=0, b=0 \text{ or } a = 1, b= 1.
    \end{cases}
  \end{equation}
\end{definition}
We extend this addition to \(m\)-tuples of bits and write
\begin{equation*}
  \boxplus \from \{0,1\}^{m} \times \{0,1\}^{m} \to \{0,1\}^{m}, \qquad
  (a_{1}, \dots , a_{m}, b_{1}, \dots , b_{m}) \mapsto (a_{1} \boxplus b_{1}, \dots , a_{m} \boxplus b_{m}).
\end{equation*}

The method described below, called \emph{Bennett's trick}, is the first step for turning classical algorithms into quantum ones.

\begin{remark}\label{rmk:Bennetts-trick}
  We associate  any map \( f \from \{0,1\}^n \to \{0,1\}^m \), the map
  \begin{equation*}
    B(f) \from \{0,1\}^{n+m} \to \{0,1\}^{n+m}, \qquad (x,y) \mapsto (x, f(x) \boxplus y),
  \end{equation*}
  Note that
  \begin{equation*}
    (B(f) \circ B(f))(x,y) = (x, f(x) \boxplus y) = (x, f(x) \boxplus f(x) \boxplus y) = (x,y)
  \end{equation*}
  implying that \(B(f)\) is a bijection.
\end{remark}

Using the computational basis \(( \ket 0, \ket 1 )\) of  the qubit \( \mathbb{C}^{2}\),
we may now lift \(B(f)\) to a linear map.
To that end, given \((x,y)= (x_{1}, \dots , x_{n}, y_{1}, \dots , y_{m}) \in\{0,1\}^{n+m}\), we write
\(\ket{x,y}\eqdef \ket{x_{1}\dots x_{n}y_{1}\dots y_{m}}\).

\begin{definition}\label{def:associated-unitary}
To any map \( f \from \{0,1\}^n \to \{0,1\}^m \), we associate the linear map
\begin{equation}
    UB(f) \from (\mathbb{C}^2)^{\otimes (n+m)} \to (\mathbb{C}^2)^{\otimes (n+m)}, \qquad \ket{x,y} \mapsto \ket{x, f(x) \boxplus y}.
\end{equation}
\end{definition}

To prove that \(UB(f)\) can be part of a quantum algorithm, we need to show that it is unitary.

\begin{lemma}\label{lem:associated-unitary-is-unitary}
  For any \( f \from \{0,1\}^n \to \{0,1\}^m \), the associated map \( UB(f) \) is unitary.
\end{lemma}
\begin{proof}
  Since \(UB(f)\) is unitary if and only if it is an isometry, we simply need to verify that it preserves the inner product.
  Let \( \ket{x_1, y_1} \) and \( \ket{x_2, y_2} \) be basis vectors in \( (\mathbb{C}^2)^{\otimes (n+m)} \).
  Their scalar product is
  \begin{equation*}
    \braket{x_1, y_1}{x_2, y_2} = \delta_{x_1,x_2} \delta_{y_1,y_2}.
  \end{equation*}
  Conversely, we get
  \begin{equation*}
    = \ket{x_1, f(x_1) \boxplus y_1}, \quad UB(f)\ket{x_2, y_2} = \ket{x_2, f(x_2) \boxplus y_2}.
  \end{equation*}
  The inner product of the transformed vectors is
  \begin{equation*}
    \braket{UB(f)\ket{x_1, y_1}}{UB(f)\ket{x_2, y_21}}
    = \braket{x_1, f(x_1) \boxplus y_1}{x_2, f(x_2) \boxplus y_2}
    = \delta_{x_1,x_2} \delta_{f(x_1) \boxplus y_1, f(x_2) \boxplus y_2}.
  \end{equation*}
  Since \( x_1 = x_2 \) implies \( f(x_1) = f(x_2) \), this simplifies to \(\delta_{x_1,x_2} \delta_{y_1,y_2}\) and concludes the proof.
\end{proof}

To get a first idea why quantum algorithms believed to be more powerful than their classical analogues, it suffices to study transformations of a single bit.

\begin{remark}\label{rmk:only-few-trafos-are-quantum}
  While there are only four maps \(\{0,1\}\to\{0,1\}\), and therefore only four corresponding unitary transformations \( (\mathbb{C}^{2})^{\otimes 2} \to ( \mathbb{C}^{2})^{\otimes 2}\), we get for every real number \(\alpha \in \mathbb{R}\) a unitary map
  \begin{equation*}
    Z(\alpha) =
    \begin{pmatrix}
      1 & 0 & 0 & 0 \\
      0 & 1 & 0 & 0 \\
      0 & 0 & 1 & 0 \\
      0 & 0 & 0 & e^{2\pi i \alpha}
    \end{pmatrix}.
  \end{equation*}
  Moreover \( Z(\alpha) = Z(\beta)\) if and only if \(\alpha- \beta \in \mathbb{Z}\).
\end{remark}

\subsubsection{The (projective) unitary group}\label{sec:the-(projective)-unitary-group}
We conclude our discussion of unitary transformations with some abstract considerations.
We begin by formalising some basic observations about unitary maps, such as the associativity of their composition.

\begin{definition}\label{def:group}
  A \emph{group} is a set \( G \) together with a binary operation \( \cdot \colon G \times G \to G \), called \emph{multiplication}, satisfying the following axioms.
  \begin{thmlist}
    \item \emph{Associativity:} For all \( a, b, c \in G \), we have \( (a \cdot b) \cdot c = a \cdot (b \cdot c) \).
    \item \emph{Identity element:} There exists an element \( e \in G \) such that \( a \cdot e = a = e \cdot a \) for all \( a \in G \).
    \item \emph{Inverse element:} For every \( a \in G \), there exists a \( b \in G \) such that \( a \cdot b = e = b \cdot a \). In this case, we write \( a^{-1} \eqdef b \).
  \end{thmlist}
\end{definition}

We begin with some matrix-based examples of groups.

\begin{example}\label{ex:invertible-matrices}
  Let \(n\in \mathbb{N}\) be a natural number.
  \begin{thmlist}
    \item The set of invertible \( n \times n \) matrices over \( \mathbb{C} \), denoted \( GL_{n}(\mathbb{C}) \), forms the \emph{general linear group} under matrix multiplication.
    \item The unitary \( n \times n \) matrices, denoted \( U(n) \), forms  the \emph{unitary group} with  matrix multiplication as group operation.
    \item The \emph{special unitary group} \(SU(n)\) has as elements all unitary \( n \times n \) matrices \(A\) such that \(\det A =1\).
    Again, the group operation of \(SU(n)\) is given by matrix multiplication.
  \end{thmlist}
\end{example}
These groups have natural analogues in the setting of linear maps of finite-dimensional Hilbert spaces, such as the group of invertible linear transformations, the group of unitary transformations, and the group of special unitary transformations.

\begin{example}\label{ex:central-stuff}
  We fix a natural number \(n\in \mathbb{N}\).
  Another set of examples where the group operation is matrix multiplication is given below.
  \begin{thmlist}
    \item The group of scalar matrices \( \{\lambda I_n \mid \lambda \in  \mathbb{C}\setminus\{0\}\} \);
    \item The group \( \{\lambda I_n \mid \lambda \in \mathbb{C} \text{ and } \|\lambda\| = 1\} \) of scalar matrices of norm \(1\);
    \item The group of scalar matrices with entries that are \(n\)-th roots of unity.
    That is, \( \{\lambda I_n \mid \lambda \in \mathbb{C} \text{ and } \lambda^n = 1\} \).
  \end{thmlist}
\end{example}

In the context of quantum computing both of these sets of groups play a central role.
To that end, we need to define an equivalence relation on the set of (special) unitary matrices
\begin{equation}
  A\sim B \iff \text{there exists a } \lambda \in \mathbb{C} \text{ such that } \lambda I_{n} A= B.
\end{equation}
We write \([A] =\{ \lambda A \mid \lambda \in \mathbb{C}\setminus\{0\} \}\).
Matrix mult extends to these equivalence classes.
That is, we set
\begin{equation}\label{eq:mult-on-quotient}
  [A] \cdot [B] \eqdef [AB] \qquad \text{ for all } A, B \in \Mat_{n} ( \mathbb{C}).
\end{equation}
A direct computation shows that this operation does not  depend on the choice of representative.
\begin{definition}\label{def:pun-sun}
  Let \(n\in \mathbb{N}\).
  The \emph{ projective unitary group} is the set \(\{[A] \mid A \in U(n)\}\) endowed with the multiplication of Equation~\ref{eq:mult-on-quotient}.

  Similarly, the \emph{projective special unitary group} is the set \(\{[A] \mid A \in SU(n)\}\) endowed with the multiplication of Equation~\ref{eq:mult-on-quotient}.
\end{definition}

\begin{definition}\label{def:group-homomorphism}
  A \emph{group homomorphism} between groups \( G \) and \( H \) is a map \( f \colon G \to H \) such that \(f(a \cdot b) = f(a) \cdot f(b)\) for all \( a, b \in G \).
\end{definition}

When scaling a problem, we often want to extend the input size \( n \) to a possibly larger input size \( n + m \).

\begin{example}\label{ex:tensor-product-of-unitary-groups}
  There is a group homomorphism \( U(n) \times U(m) \to U(n + m) \) given by the block diagonal embedding
  \begin{equation*}
    U(n) \times U(m) \to U(n + m), \qquad
    (A, B) \mapsto \begin{pmatrix}
      A & 0 \\
      0 & B
    \end{pmatrix}.
  \end{equation*}
\end{example}

Note that this partially highlights the special role of entanglement since not every matrix in \( U(2^n) \) comes from a matrix of \( U(2)^{\times n} \).

\begin{example}\label{ex:equiv-of-pun-sun}
  The map \( PSU(n) \to PU(n) \) defined by
  \begin{equation*}
    PSU(n) \to PU(n), \qquad  [A] \mapsto [A],
  \end{equation*}
  is an isomorphism of groups.
\end{example}

\begin{definition}\label{def:group-action}
  A \emph{group action} of a group \( G \) on a set \( X \) is a map \( G \times X \to X \), denoted by \( (g, x) \mapsto g \cdot x \), such that for all \(x\in X\) and \(g,h \in G\) we have \( e \cdot x = x \) and \( g \cdot (h \cdot x) = (gh) \cdot x \).
\end{definition}

The example below can be considered a reformulation of Lemma~\ref{lem:induced-map-on-states}.
\begin{example}\label{ex:states}
  The group \( U(n) \) acts on the states of \( \mathbb{C}^n \) by matrix-vector multiplication
  \begin{equation*}
    U(n) \times\{\text{states of } \mathbb{C}^{n}\}, \qquad   A \cdot [\ket{v}] = [A\ket{v}].
  \end{equation*}
\end{example}

By Lemma~\ref{lem:mixed-state-action}, conjugation by unitary matrices yields an action on density matrices.
\begin{example}\label{ex:group-action-on-mixed-states}
  The group \( U(n) \) acts on the set of density matrices of \( \mathbb{C}^n \) by conjugation
  \begin{equation*}
    U(n) \times\{\text{density matrices of } \mathbb{C}^{n}\}, \qquad
    (A, \rho) \mapsto A\rho A^{\dagger}.
  \end{equation*}
\end{example}

\begin{lemma}\label{lem:grp-homomorphism}
  There is a group homomorphism
  \begin{equation*}
    \pi \from U(n) \to PSU(n), \qquad A \mapsto [A].
  \end{equation*}
\end{lemma}
\begin{proof}
 Observe that for any \(A \in U(n)\) there exists a \(\lambda\in \mathbb{C}\) such that \(\det(\lambda A)= \lambda^{n} \det(A) = 1\).
 Since, \([A] = [\lambda A]\), the map \(\pi\from U(n) \to PSU(n)\) is well-defined.
 Moreover, since \(\pi(AB) = [AB] = [A][B] = \pi(A) \pi(B)\).
\end{proof}

\begin{example}\label{ex:group-action-factor-through}
  Consider two unitary matrices \(A, B \in U(n)\) such that \([A] = [B]\).
  Then ,there exists a \(\lambda \in \mathbb{C}\) with \(\|\lambda \| = 1\) and \(\lambda A = B\).
  For any \(M \in \Mat_{n}( \mathbb{C})\), we have
  \begin{equation*}
    BMB^{\dagger} = (\lambda A) M (\lambda A)^{\dagger} = \lambda \lambda^{\dagger} AMA^{\dagger} = \|\lambda\| AMA^{\dagger}.
  \end{equation*}
  Therefore, \(PU(n) \cong PSU(n)\) acts by conjugation on all matrices, and, by restriction, on density matrices.
  \begin{equation*}
    PSU(n) \times\{\text{density matrices of } \mathbb{C}^{n}\} \to\{\text{density matrices of } \mathbb{C}^{n}\},\qquad
    ([A], \rho) \mapsto A\rho A^{\dagger}.
  \end{equation*}
\end{example}

While this goes beyond the content of this lecture the fact that transformations are governed by \(SU(n)\) (and not just \(U(n)\)) have a few theoretical advantages.
In particular \(SU(n)\) is a connected and simply-connected Lie group which means that it is---in a certain sense---completely determined by its associated Lie algebra \(\mathfrak{su}_{n}\).

\subsection{Measurements and the Born rule}\label{sec:measurements-and-the-born-rule}
In the simplest case, a quantum algorithm consists of three main steps: state preparation, application of unitary gates, and measurement of the outcomes. Having discussed states and unitaries, we now focus on measurements.
In our interpretation of quantum mechanics (specifically, the Copenhagen interpretation), measurements are represented by orthogonal projections.

\begin{lemma}\label{lem:orthogonal-projections-and-probabilities}
  For any normalised vector \(\ket{v} \in V\) and orthogonal projection \(e \from V \to V\) an orthogonal projection.
  The number \(\braket{v}{e \mid v}\) is real and satisfies \(0 \leq \braket{v}{e \mid v}\leq 1\).
\end{lemma}
\begin{proof}
  % TODO: Fix the \ket issue here!
  Since \( e \) is an orthogonal projection, it satisfies \( e^2 = e \) and \( e = e^\dagger \).
  Therefore,
  \begin{equation*}
    \braket{v}{e \mid v} = \braket{v}{e^2 \mid v} = \braket{e^\dagger v}{e v} = \braket{e v}{e v} \geq 0.
  \end{equation*}
  To show that \(1\) is an upper bound, we decompose \( \ket{v} \) into two summands \( \ket{v} = e\ket{v} + (\ket{v} - e\ket{v}) \).
  These are orthogonal, as a direct computation shows
  \begin{equation*}
    \braket{e\ket{v}}{\ket{v} - e\ket{v}} = \braket{e\ket{v}}{\ket{v}} - \braket{e\ket{v}}{e\ket{v}} = \braket{e\ket{v}}{\ket{v}} - \braket{e\ket{v}}{\ket{v}} = 0.
  \end{equation*}
  Therefore
  \begin{equation*}
      0 \leq \braket{e\ket{v}}{e\ket{v}} \leq \braket{e\ket{v}}{e\ket{v}} + \braket{\ket{v} - e\ket{v}}{\ket{v} - e\ket{v}} = \braket{\ket{v}}{\ket{v}} = 1. \qedhere
  \end{equation*}
\end{proof}

\begin{definition}\label{def:orthogonal-idempotent}
  A set \(\{e_{1}, \dots , e_{n}\}\) of projections \(e_{1}, \dots , e_{n} \from V\to V\) is \emph{(pairwise) orthogonal} if \(e_{i}e_{j} = 0 =  e_{j}e_{i}\) for all \(1\leq i< j \leq n\).
\end{definition}

\begin{lemma}\label{lem:part-of-1-means-orth}
  Consider projections  \(e_{1}, \dots , e_{n} \from V\to V\) such that \(e_{1} + \dots + e_{n} = \id_{V}\).
  Then,  \(\{e_{1}, \dots , e_{n}\}\) is pairwise orthogonal.
\end{lemma}
\begin{proof}
  Fix \(1\leq i \leq n\).
  We have
  \begin{equation*}
    e_i = e_i \id_V = e_i \left( \sum_{k=1}^n e_k \right) = e_i^2 + \sum_{k \neq i} e_i e_k = e_i + \sum_{k \neq i} e_i e_k.
  \end{equation*}
  Subtracting \( e_i \) from both sides yields \(0 = \sum_{k \neq i} e_i e_k\).
  Since \(\tr(e_{i} e_{k}) = \dim (\im e_{k} \cap \im e_{i})\geq 0\), we have \(e_i e_k = 0\) for all \( k \neq i \).
  % TODO: Make more explicit \(V = U \oplus U' \) and \(V = W \oplus W'\) for \(U =\im p_{i}\) and \(W = \im p_{k}\). Then  \(P_{i}p_{k}\) operates on\(U \cap W\)... trace is dimension
\end{proof}

Let \(\ket{v} \in V\) be a normalised vector and consider  \(\alpha\in \mathbb{R}\).
Then,
\begin{equation*}
  \braket{v}{e \mid v} = \braket{ e^{2\pi i \alpha}v}{e \mid e^{2\pi i \alpha}v}.
\end{equation*}
Thus, the number \(\braket{v}{e \mid v}\) only depends on the state \([\ket v] \in V\) and not the representative.

\begin{definition}\label{def:projective-measurement}
  Let \( V \) be a Hilbert space.
  A \emph{projective measurement} is a set \( E=\{e_1, \dots, e_n\} \) of pairwise orthogonal projections such that
  \begin{equation*}
    e_1 + \dots + e_n = \id_V.
  \end{equation*}
\end{definition}

The below defined way for computing the probability of measuring the \(i\)-th outcome is known as  the \emph{Born rule}.

\begin{definition}\label{def:born-rule}
  Let \(H\) be a Hilbert state and \( E=\{e_1, \dots, e_n\} \) a projective measurement.
  For a given state \( \ket{v} \in V \), the probability of obtaining the \( i\)-th outcome upon measurement is
  \begin{equation*}
    \braket{v}{e_i \mid v}.
  \end{equation*}
\end{definition}

Now, measuring with \(M\) an getting outcome \(i\), corresponds to the map
\begin{equation*}
  \{\text{states of }V\} \to\{ \text{states of }V\}, \qquad \ket {v} \mapsto \frac{1}{\sqrt{\braket{v}{e_{i}\mid v}}} e_{i}\ket v.
\end{equation*}
This way of updating a state after measuring is called \emph{Lüders' rule}.

\begin{example}\label{ex:onb-measures}
  Consider a Hilbert space \( V \) with an orthonormal basis \( \mathcal{B} = (\ket{b_1}, \dots, \ket{b_n}) \).
  The maps \( e_i = \ket{b_i}\bra{b_i} \colon V \to V \) are orthogonal projections, and satisfy \(\sum_{i=1}^n e_i = \id_V\).

  This measurement provides detailed information about the system, but at the expense of ``collapsing'' the state: after measurement, any state will be mapped to one of the basis vectors in \( \mathcal{B} \).

  At the other extreme, consider the trivial projective measurement \( M = (\id_V) \).
  This measurement does not alter the state of the system, but it also provides no information.

  In practice, we will often use measurements that balance between these two extremes.
\end{example}

\subsubsection{Measurements for mixed states}\label{sec:measurements-for-mixed-states}
Consider a state \(\ket{v}\) with associated density matrix \(\rho_{\ket{v}}\).
We fix an orthonormal basis \( \mathcal{B}= (b_{1}, \dots , b_{n})\) of \(V\) with \(b_{1} = v\).
For any orthogonal projection \(f \from V \to V\), we have
\begin{equation*}
  \tr(\rho_{v} f)
  = \tr(\ket{v} \bra{v\mid f})
  =\sum_{i=1}^n \braket{b_i}{v} \braket{v}{ f \mid b_i}
  = \braket{v}{v} \braket{v}{v\mid f}
  = \braket{v}{f \mid v}.
\end{equation*}

\begin{lemma}\label{lem:get-prob-values}
  Let \(e\from V\to V\) be an orthogonal projection on a Hilbert space and \(\rho \from V \to V\) a mixed stated.
  Then \(0\leq \tr(e\rho) \leq 1\) is a real number.
\end{lemma}
\begin{proof}
  Since \( e \) is an orthogonal projection, it is positive semi-definite and satisfies \( p^2 = p \).
  The mixed state \( \rho \) is self-adjoint, positive semi-definite and has trace 1.
  Thus, by the spectral theorem, there exists an orthonormal basis \(\mathcal{B}= (\ket{b_{1}}, \dots , \ket{b_{n}})\) of \(V\) such that
  \begin{equation*}
    \rho = \sum_i \lambda_i \ket{b_{i}} \bra{b_{i}}
  \end{equation*}
  with  \( 0 \leq \lambda_i  \in \mathbb{R}\) and \( \sum_i \lambda_i = 1 \).
  We compute
  \begin{equation*}
    \tr(e\rho)
    = \sum_i \lambda_i \tr(e \ket{b_i} \bra{b_i}) =
    \sum_i \lambda_i \braket{b_i}{e \mid b_i}
  \end{equation*}

  Since \( 0 \leq \braket{b_i}{p \mid b_i} \leq 1 \) for each \( 1\leq i \leq n \), it follows that \( 0 \leq \tr(e\rho) \leq 1 \).
\end{proof}

The definition below is known as the \emph{Born rule for mixed states}.
\begin{definition}\label{def:Born-rule-for-projective-measurment}
  Let \(M= (e_{1}, \dots , e_{k})\) be a projective measurement on a Hilbert space \(V\).
  The  probability of measuring the \(i\)-th outcome for the mixed state \(\rho \from V\to V\) is
  \begin{equation*}
    \tr(e_{i}\rho).
  \end{equation*}
\end{definition}

To update a mixed state after a measurement, one uses the following variant of \emph{Lüders' Rule }
\begin{equation*}
  \{\text{ mixed states of } V\} \to\{ \text{ mixed states of } V\}, \qquad
  \rho \mapsto \tfrac{e_{i}\rho e_{i}}{\tr{\rho e_{i}}}.
\end{equation*}


The big advantage about mixed states in the context of  measurements is that we can draw them as traces.
\begin{equation*}
  % TODO \inputtikz{GENERIC TRACE PIC}
\end{equation*}

\cleardoublepage
\subsection{Exercises}\label{sec:basics-exercises}

\begin{exercise}\label{exe:inner-product-is-sesquilinear}
  Show that an inner product \(\braket{\blank}{\bblank} \from V\times V \to \mathbb{C}\) is sesquilinear.

  \textit{Hint:} Use the Hermitian symmetry to move arguments from the first to the second position and vice versa.
\end{exercise}

\begin{exercise}\label{exe:orthogonal-proj-equivalent}
  Suppose \(V\) is a Hilbert space.
  Show that a linear map \(f \from V\to V\) is an orthogonal projection onto a subspace \(U \subset V\) if and only if \(f^{2} = f\), \(\im f = U\), and \(\ker f = U^{\perp}\).
\end{exercise}

\begin{exercise}\label{exe:orth-complement-direct-sum}
  Let \(U\subset V\) be a subspace of a Hilbert space.
  Show that \(U\oplus U^{\perp} = V\)
\end{exercise}

\begin{exercise}\label{exe:unitary-matrix-unitary-map}
  Suppose \(V\) is a Hilbert space with orthonormal basis \(\mathcal{B}\) and \(f\from V\to V\) is a linear map.
  Show that \(f\) is unitary if and only if its transformation matrix is unitary.
\end{exercise}

\begin{exercise}\label{exe:proof-of-lem-canonical-iso-to-dual-space}
  This exercise walks you through a guided proof of Lemma~\ref{lem:canonical-iso-to-dual-space}.
  Consider a finite-dimensional Hilbert space \( V \) and fix an orthonormal basis \( \mathcal{B} = (b_1, \dots, b_n) \) of \( V \).
  \begin{thmlist}
    \item Which axiom of the inner product ensures that \( d \) is well-defined?
    \item Show that \( d(\mathcal{B}) = ( d(b_1), \dots, d(b_n)) \) is a basis of \( \rd*{V}\).
    (Hint: Evaluate \( d(b_i) \) on \( b_j \) for all \( i, j \).)
    \item Use the anti-linearity of \( d \) to conclude that \( d \) is a bijection.
    \item Verify that the given inner product on \( \rd*{V} \) satisfies all axioms of an inner product.
  \end{thmlist}
\end{exercise}

% TODO TENSOR PRODUCTS
\begin{exercise}\label{exe:bilinear-vs-linear-hom}
  Show that there is a bijection
  \begin{align*}
    \{\text{bilinear maps } V\times W \to X \} & \to \{\text{linear maps} V\to \Hom_{ \mathbb{C}}(W,Z) \} \\
    b & \mapsto (v\mapsto b(v, \blank)) \\
    (B(v) w \longmapsfrom (v,w)) &\longmapsfrom B
  \end{align*}
\end{exercise}

\begin{exercise}\label{exe:translating-bilinear-to-sesquilinear}
  To any vector space \( V \), we assign a new vector space structure on \(V\) with the same vector addition and scalar multiplication defined by
  \begin{equation*}
    \lambda \cdot v \eqdef \overline{\lambda}v \qquad\quad \text{ for all }\lambda \in \mathbb{C}, v\in V.
  \end{equation*}
  To better distinguish between \(V\) and \(\overline{V}\), we write \(\overline{v}\in \overline{V}\) for the corresponding element \(v\in V\).

  Show that a map \( \phi \from V \times V \to \mathbb{C} \) is sesquilinear if and only if the map \( \psi \from \overline{V} \times V \to \mathbb{C} \) defined by \( \psi(\overline{v}, w) = \phi(v, w) \) is bilinear.
\end{exercise}



\cleardoublepage
\subsection{Study goals}\label{sec:study-goals}

\cleardoublepage
\section{Quantum computation via the zx calculus}\label{sec:quantum-computation-via-the-zx-calculus}
\begin{description}
  \item[Kitaev--Solvay theorem] ``All models for quantum computing are equivalent''
  \item[Qubits] Tensor products of vector spaces, homomorphisms, matrices.
  \item[Bloch sphere model]
  \item[ZX-calculus] Spiders et al
  \item[Relations between spiders]
\end{description}

\part{Quantum algorithms}
\section{Quantum algorithms}\label{sec:quantum-algorithms}

\begin{definition}\label{def:group}
  A \emph{group} is a pair \((G, \bullet)\) comprising a set \( G \)  and a binary map
  \begin{equation*}
    \bullet \from G \times G \to G, \qquad (g,h) \mapsto  g\bullet h,
  \end{equation*}
  called \emph{multiplication} or \emph{group operation}, satisfying
  \begin{description}
    \item[\emph{Associativity}] \((a \bullet b) \bullet c = a \bullet (b \bullet c)\) for all \( a, b, c \in G \),
    \item[\emph{Neutrality}] There exists an element \( e \in G \), called the \emph{neutral element} or \emph{unit} of \(G\), such that \( e \bullet a = a = a \bullet e \) for all \( a \in G \).
    \item[\emph{Invertibility}] For each \( a \in G \), there exists an \( a^{-1} \in G \) such that \( a \bullet a^{-1} = a^{-1} \bullet a = e \).
  \end{description}
  We call \(G\) \emph{abelian} if  \( a\bullet b = b \bullet a \) for all \(a, b\in G\).
\end{definition}

\begin{notation}
  Let \((G, \bullet)\) be a group.
  \begin{thmlist}
    \item In case its multiplication is apparent, from the context, we will identify \(G\) with the pair \((G, \bullet)\).
    \item Moreover, we will often use the short-hand notation
    \begin{equation*}
      ab \eqdef a \bullet b \qquad \text{ for all } a, b \in G.
    \end{equation*}
    \item In order to emphasise that a group is abelian \(G\) is abelian, we will denote its group operation by
    \begin{equation*}
      + \from G \times G \to G, \qquad (g,h) \mapsto g + h
    \end{equation*}
    and its neutral element by \(0\).
  \end{thmlist}
\end{notation}

Note that the neutral element \(e\) of a group \(G\) is unique.
In case \(f\in G\) satisfies \(af=a=fa\) for all \(a \in G\), we have
\begin{equation*}
  e=ef=f.
\end{equation*}

Similarly, inverses are unique.
Given an element \(a\in G\), and \(a^{-1}, b\in G\) such that \(a^{-1}a = e = aa^{-1}\) and \(ba = e = ab\), we have
\begin{equation*}
  a^{-1} = a^{-1}e = a^{-1}(ab) = (a^{-1}a)b = eb = b.
\end{equation*}

\begin{description}
  \item[Simon's algorithm]
  \item[Shor's algorithm]
  \item[Grover's algorithm]
  \item[HHL algorithm]
\end{description}


\part{Quantum complexity theory}
\section{Quantum computing from a complexity theory perspective}\label{sec:quantum-computing-from-a-complexity-theory-perspective}
\subsection{Some complexity theory terminology}\label{sec:some-complexity-theory-terminology}


\begin{definition}\label{def:letters-words-languages}
  A \emph{word} in the \emph{letters} \(\{0,1\}\) is a finite sequence \(w \eqdef (s_{1}, \dots , s_{n})\), where \(n\in \mathbb{N}\) and \(s_{i}\in \{0,1\}\) for all \(1 \leq i \leq n\).

  We call \(|x|\eqdef n\) the \emph{length} of \(w\) and write \(\{0,1\}^{n}\) for the set of all words of length \(n\).
  Furthermore, we set
  \begin{equation}\label{eq:set-of-words}
    \{0,1\}^{*}\eqdef \bigcup_{n\geq 0}\{0,1\}^{n}.
  \end{equation}

  A \emph{decision problem} is a subset \(L \subset \{0,1\}^{*}\).
\end{definition}

There is a bijection
\begin{equation}
  \begin{aligned}
    \{\text{decision problems}\} &\iso \{f\from\{0,1\}^{*} \to \{0,1\}\}\\
    L &\mapsto  f\from\{0,1\}^{*} \to \{0,1\}, \quad f(x)= 1 \iff x \in L \\
    L_{f}\eqdef\{x \in\{0,1\}^{*} \mid f(x) = 1\} & \longmapsfrom f
  \end{aligned}
\end{equation}
We will often view decision problems in terms of their unique associated maps \(\{0,1\}^{*}\to\{0, 1\}\)
\medskip

Besides formalising algorithms, we want to discuss their (asymptotic) efficiency.
To that end, we need to a notion of functions being ``approximately similar''.

\begin{definition}\label{def:O-notation}
  Let \(D\) be a subset of \( \mathbb{R}\) and\(f, g \from D \to \mathbb{R}\).
  We write \(f \in O(g)\) if there exists a \(M\in \mathbb{R}\) such that
  \begin{equation}
    |f(x)| \leq |M g(x) | \qquad\quad \text{ for all } x \in D.
  \end{equation}

\end{definition}


\subsection{Universal Turing machines}\label{sec:universal-turing-machines}
We now formalise the principles outlined in our construction of the map \(\mathrm{add}\) in terms of Turing machines.

\begin{definition}\label{def:Turing-machine}
  A \emph{Turing machine} is a triple \((\Gamma, Q, \delta)\) comprising
  \begin{description}
    \item[\(\Gamma\)] the \emph{alphabet} \(\{0,1,\rhd, \square \}\) of \(M\).
    It includes  the (distinguished) \emph{start symbol} \(\rhd\) and the \emph{blank symbol} \(\square\).
    \item[\(Q\)] a finite set of \emph{states}, containing a \emph{starting state} \(q_{\mathrm{start}}\) and a \emph{halting state} \(q_{\mathrm{halt}}\).
    \item[\(\delta\)] a \emph{transition function}
    \begin{equation*}
      \delta \from Q \times \Gamma \to Q \times \Gamma\times\{L,S,R\}.
    \end{equation*}
  \end{description}
\end{definition}

\begin{example}\label{ex:add-as-Turing-machine}
  TODO: give an easy introduction for a Turing machine.
\end{example}
A widely held belief is that Turing machines provide a model for ``computable functions''.

\begin{remark}[Church--Turing thesis]\label{rmk:Church-Turing-thesis}
  A function \(f \from  \mathbb{N} \to  \mathbb{N}\) can be computed by some device if it can be computed by a Turing machine.
\end{remark}


\begin{definition}\label{def:time-constructible-function}
  A map \(f \from \mathbb{N} \to \mathbb{N}\) is \emph{time-constructible} if there exists a Turing machine \(M\) such that for each input \(n\) (in binary presentation) \(M\) halts in \(O(f)\) steps with output \(f(n)\) (in  binary presentation).
\end{definition}

\begin{remark}\label{rmk:encoding-Turing-machines}
  A Turing machine \(M\) itself can be encoded as a word in the letters \(\{0,1, \square\}\).
  Explicitly, we
\end{remark}

\begin{theorem}\label{thm:existence-of-universal-TM}
  There exists a Turing machine \(U\)
\end{theorem}




\begin{definition}\label{def:deciding-a-decision-problem}
  A Turing machine \(M\) \emph{decides} a decision problem \(L \subset\{0,1\}^{*}\) if \(M\) halts for each \(x \in \{0,1\}^{*}\) and the output of \(M\) is \(1\) if \(x\in L\) and \(0\) otherwise.
\end{definition}


\begin{definition}\label{def:Dtime}
  Consider a time constructible function \(T \from \mathbb{N} \to \mathbb{N}\).
  The class \(\Dtime(T)\) consists of all decision problems \(L \subset\{0,1\}^{*}\) such that there exists a Turing machine \(M\) that halts in \(O(T)\) steps and decides \(L\).
\end{definition}


\begin{definition}\label{def:class-p}
  The \emph{complexity class} \(\Pclass\) is given by
  \begin{equation}\label{eq:p-class}
    \Pclass = \bigcup_{n\geq 1} \Dtime(t^{n}).
  \end{equation}
\end{definition}

We think of problems in the complexity class \(\Pclass\) as being precisely the  ``feasible'' problems.
The \emph{strong} Church--Turing thesis states:
\begin{remark}[strong Church--Turing thesis]\label{rmk:strong-Church-Turing-thesis}
  If a function \(f \from \mathbb{N} \to \mathbb{N}\) can be computed in polynomial time by some device, it can be computed in polynomial time by a Turing machines.
\end{remark}

Quantum computers potentially challenge the point of view.
As we will see over the span of this course, they provide ways of performing tasks efficiently (in polynomial time) for which no efficient classical solution is know.

\begin{definition}\label{def:class-np}
  A decision problem \(L\) is in \(\NPclass\) (Nondeterministic Polynomial time) if there exists a polynomial-time TM \(V\), called a \emph{verifier}, and a polynomial \(p\), such that:
  \(x \in L \iff \exists y \in \{0,1\}^{p(|x|)} \text{ s.t. } V(x, y) = 1\).
\end{definition}

\begin{remark}
  While \(\Pclass \subseteq \NPclass\), we believe \(\Pclass \neq \NPclass\). Quantum computers are interesting because they can solve certain problems in \(\NPclass\) (like integer factoring) in polynomial time, whereas we suspect no such polynomial-time classical algorithm exists.
\end{remark}

\part{Quantum error correction}
\section{Fault-tolerance}\label{sec:fault-tolerance}
\begin{description}
  \item[Quantum compilation stack]
  \item[Stabiliser codes]
  \item[Toric and surface codes]
  \item[Outlook: Toplogical quantum computing]
\end{description}
% \appendix
% \section{Linear algebra}\label{sec:linear algebra}

% \subsection{Unitary vector spaces}\label{sec:unitary vector spaces}

% The state space of a quantum system is a complex vector space equipped with an inner product.

% \begin{definition}\label{def:complex-vector-space}
%   A \emph{complex vector space} is a set \(V\) equipped with two operations,
%   \begin{equation*}
%     + \from V \times V \to V, \qquad (v,w) \mapsto v + w, \qquad
%     \cdot \from \mathbb{C} \times V \to V, \qquad (\lambda, v) \mapsto \lambda v,
%   \end{equation*}
%   called \emph{addition}, respectively \emph{scalar multiplication},
%   that satisfy
%   \begin{thmlist}
%     \item the addition is \emph{associative}: for every \(v,w,x \in V\) we have \((v+w)+x = v+(w+x)\),
%     \item the addition is \emph{commutative}: \(v+w = w+v\) for every \(v,w\in V\),
%     \item There exists an element \(0\in V\), called a \emph{zero}, such that \(v+0 = v = 0+v\) for all \(v \in V\),
%     \item the scalar multiplication is \emph{associative}: for all \(\lambda, \mu \in \mathbb{C}\) and \(v \in V\) we have  \(\lambda(\mu v) = (\lambda \mu) v\),
%     \item the scalar multiplication is \emph{unital}: we have \(1_{\mathbb{C}}v= v\)for all \(v\in V\).
%     \item the scalar multiplication is \emph{distributive} \(\lambda(v+w) = \lambda v + \lambda w\) and \((\lambda +\mu)v = \lambda v + \mu v\).
%   \end{thmlist}
% \end{definition}

% \begin{example}[The space \(\mathbb{C}^n\)]\label{ex:complex-n-space}
%   The set of all \(n\)-tuples of complex numbers,
%   \begin{equation*}
%     \mathbb{C}^n = \left\{ \begin{pmatrix} z_1 \\ \vdots \\ z_n \end{pmatrix}  \;\middle|\; z_i \in \mathbb{C} \right\},
%   \end{equation*}
%   is a complex vector space under component-wise addition and scalar multiplication.
% \end{example}

% \begin{example}[The space of complex matrices]\label{ex:mat-complex}
%   The set \(\Mat_{n,m}(\mathbb{C})\) of all \(n \times m\) matrices with complex entries forms a complex vector space. The operations are defined by:
%   \begin{thmlist}
%     \item \((A + B)_{ij} = A_{ij} + B_{ij}\)
%     \item \((cA)_{ij} = c \cdot A_{ij}\)
%   \end{thmlist}
% \end{example}

% \begin{definition}\label{def:inner-product}
%   An \emph{inner product} on a complex vector space \(V\) is a map \(\braket{ \blank }{=} \from  V \times V \to \mathbb{C}\) satisfying
%   \begin{thmlist}
%     \item \emph{sesquilinearity}: for every \(\lambda,\mu \in \mathbb{C}\) and \(v,w,x, y\in V\) we have
%     \begin{equation*}
%       \braket{\lambda v+ w}{\mu x+ y} = \overline{\lambda}\mu \braket v x + \overline{\lambda} \braket v y + \mu \braket{w}{x} + \braket{w}{y}.
%     \end{equation*}
%     \item \textit{hermitian} \(\braket v w = \overline{\braket w v}\).
%     \item \emph{positivity} \(\braket v v \geq 0\) and \(\braket v v = 0\) if and only if \(v = 0\).
%   \end{thmlist}
% \end{definition}

% \begin{example}\label{ex:standard-scalar-product}
%   The \emph{standard scalar product} on  \(\mathbb{C}^{n}\) is given by
%   \begin{equation*}
%     \braket v w \eqdef \sum_{i=1}^{n} \overline{v_{i}}w_{i}.
%   \end{equation*}

% \end{example}
% \begin{definition}\label{def:onb}
%   A basis \((e_{1}, \dots , e_{n})\) of a unitary vector space \(V\) is \emph{orthonormal basis} if
%   \begin{equation*}
%     \braket{e_{i}}{e_{j}} =
%     %     \begin{cases}
      %     %       1 & \(i=j\)\\
%     %   0 & \text{otherwise.}
%     % \end{cases}
%   \end{equation*}
% \end{definition}


% \begin{definition}\label{def:unitary-matrix}
%   A matrix \(U \in \Mat_{n}( \mathbb{C})\) is
%   \begin{thmlist}
%     \item \emph{unitary} if \(UU^{*}= \Id_{n} =  U^{*}U\),
%     \item \emph{positive}
%     \item \emph{self-adjoint}
%     \item \emph{orthogonal projection} if
%   \end{thmlist}
% \end{definition}

% \begin{remark}\label{rem:all-normal}
%   Note that all the classes of matrices defined in Definition~\ref{def:unitary-matrix} are in particular \emph{normal}.
% \end{remark}

% \begin{theorem}\label{thm:spectral-theorem}
%   A matrix \(A \in \Mat_{n}(\mathbb{C})\) is normal if and only if it is unitarily diagonalisable.
%   That is, there exists a unitary matrix \(U\) and a diagonal matrix \(D\) such that \(A = UDU^*\). Equivalently, \(A\) can be written as:
%   \begin{equation*}
%       A = \sum_{i} \lambda_i P_i,
%   \end{equation*}
%   where \(\lambda_i\) are the eigenvalues of \(A\) and \(P_i\) are orthogonal projections onto the corresponding eigenspaces.
% \end{theorem}


% \subsection{Tensor products}\label{sec:tensor-products}


% \begin{definition}\label{def:tensor-product}
%   Consider two complex vector spaces \(V, W\).
%   A \emph{tensor product} of \(V\) and \(W\) is a tuple \((V \otimes W, \blank \otimes = )\) comprising a vector space \(V \otimes W\) and a bilinear map
%   \begin{equation*}
%     \blank \otimes = \from V\times W \to V\otimes W, (v, w) \mapsto v\otimes w
%   \end{equation*}
%   such that for each bilinear map \(b \from V \times W \to X\) there is a unique linear map \(f_{b}\from V\otimes W \to X\) such that
%   \begin{equation*}
%     f_{b}(v\otimes w) = b(v,w) \qquad \text{ for all } v \in V, w\in W.
%   \end{equation*}
% \end{definition}

% \begin{theorem}\label{thm:tensor-product-existence}
%   For any two complex vector spaces \(V\) and \(W\), there exists a tensor product \((V \otimes W, \blank \otimes = )\).
%   Moreover, if \((V \boxtimes W, \blank \boxtimes = )\)  is another tensor product,
%   there is a unique isomorphism of vector spaces
%   \(f \from V\otimes W \to V \boxtimes W\) such that \(f(v\otimes w) = v \boxtimes w\).
% \end{theorem}

% \begin{lemma}\label{lem:basis-of-tp}
%   Let \(V, W\) be two complex vector spaces with basis \((v_{1}, \dots , v_{n})\) and \((w_{1}, \dots , w_{m})\), respectively.
%   Then \((v_{i}\otimes w_{j})_{1\leq i \leq n, 1\leq j \leq m}\) is a basis of \(V \otimes W\).
% \end{lemma}

% \begin{lemma}\label{lem:assoc-of-tps}
%   Let \(V,W,X\) be three complex vector spaces.
%   There is a (canonical) isomorphism
%   \(V \otimes (W\otimes X) \to (V\otimes W) \otimes X\).
% \end{lemma}



% \subsection{The Trace Operator}\label{subsec:trace}

% \begin{definition}\label{def:trace}
%   The \emph{trace} of a square matrix \(A \in \Mat_{n}(\mathbb{C})\) is the sum of its diagonal elements:
%   \begin{equation*}
%       \tr(A) \eqdef \sum_{i=1}^n A_{ii}.
%   \end{equation*}
% \end{definition}

% \begin{remark}\label{rem:trace-properties}
%   The trace is a linear functional and satisfies the \emph{cyclic property}: \(\tr(AB) = \tr(BA)\) for any \(A, B \in \Mat_{n}(\mathbb{C})\). This implies that the trace is invariant under basis changes (similarity transformations), and is equal to the sum of the eigenvalues of the matrix.
% \end{remark}


\setlength{\bibitemsep}{0\baselineskip}
\printbibliography
\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:
